---
title: "Getting started with site-level proportions"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  Estimate a proportion and its standard error at each site using
  simulated pre-kindergarten data. Learn how to read the returned
  estimates and check them with `sm_diagnose()`.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Getting started with site-level proportions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

This example estimates the proportion of students receiving free and
reduced-price meals (FRPM) at each site in 2024. We use `sm_estimate()`
to calculate the proportions and standard errors, then `sm_diagnose()`
to check the results.

The bundled `prek_sim` data describe 50 simulated pre-kindergarten sites.
They contain no real children or programs, so the results illustrate the
calculations rather than empirical Pre-K findings (see `?prek_sim`).

# Install and load

If you have not installed sitemix yet:

```r
# install.packages("pak")
pak::pak("joonho112/sitemix")
```

Then load the package and the bundled sample data:

```{r library}
library(sitemix)
data(prek_sim, package = "sitemix")
```

`prek_sim` carries four overlapping binary indicators
(`frpm`, `snap`, `wic`, `tanf`) observed across years 2021–2025.

# Estimate the FRPM proportion

Select one year and one binary indicator. The default transformation is
arcsine; the result also retains each proportion on its original scale.

```{r teaser}
frpm_2024 <- sm_estimate(
  subset(prek_sim, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)
print(
  as.data.frame(frpm_2024[
    1:5,
    c("site_id", "year", "indicator", "n", "theta_raw", "theta_hat", "se")
  ]),
  row.names = FALSE
)
```

Each row is one site in 2024. `n` is the site denominator (number
of students with non-missing FRPM status); `theta_raw` is the raw
proportion in `[0, 1]`; `theta_hat` is the same estimate on the
arcsine-stabilized scale; `se` is its delta-method standard error
on that scale. Use `theta_raw` to report percentages and keep `theta_hat`
with `se` when an analysis uses the transformed estimates.

# Read the result

The result is a `sitemix_estimates` tibble. It includes the estimates,
their standard errors, and information about how they were calculated:

```{r columns}
names(frpm_2024)
```

Start with these columns:

-   `site_id`, `year`, `indicator` — the site, year, and indicator identifying each row.
-   `n` — site denominator.
-   `theta_raw` — observed proportion in `[0, 1]` for these complete student records.
-   `theta_hat`, `se` — point estimate and SE on the
    `estimate_scale` (default `"arcsine"`).
-   `var_method` — how the standard error was calculated; the possible
    method labels are listed in [`sm_vcov()`](../reference/sm_vcov.html).
-   `flag_small_n`, `flag_below_accountability`, `flag_zero_cell`,
    `flag_suppressed` — flags for sample size, reporting thresholds,
    boundary proportions, and suppressed data. Inspect the flagged rows
    before deciding whether they belong in a particular analysis.
-   `input_mode` — `"student_level"` here; would be
    `"counts_full_suff"` or `"aggregate"` under the wrappers.

The Value section of [`sm_estimate()`](../reference/sm_estimate.html)
describes all returned columns and their conditions.

# Check estimates and uncertainty

Before passing the tibble to another analysis, run the diagnostics:

```{r diagnose}
diag <- sm_diagnose(frpm_2024, verbose = FALSE)
class(diag)
print(as.data.frame(diag), row.names = FALSE)
```

A `sitemix_diagnostics_summary` reports one row summarizing the
object, including denominator percentiles, tier counts, and checks on
the estimates and standard errors. For scalar output, check
`scalar_uncertainty_finite`, `scalar_se_positive`, and
`indicator_scale_consistent`. For output with a `V` list-column,
also inspect `v_valid` and `estimate_vcov_scale_compatible`.

Use `level = "row"` to locate individual flags. If you request covariance
matrices with `vjt = TRUE`, `level = "vcov"` returns one diagnostic row
per site-year matrix. The full walkthrough is in
[Checking estimates and handling suppressed data](a6-diagnostics-and-suppression.html).

# Check the example results

These checks confirm the expected rows and default scale for this
simulated example. Their positive-SE condition is specific to these data;
an exact census under a finite-population design can have zero sampling SE.

```{r audits}
stopifnot(nrow(frpm_2024) == 50L)  # 50 sites in 2024
stopifnot(all(frpm_2024$indicator == "frpm"))  # single indicator
stopifnot(all(frpm_2024$estimate_scale == "arcsine"))  # default scale
stopifnot(all(is.finite(frpm_2024$theta_hat)))  # no NaN/Inf
stopifnot(all(frpm_2024$se > 0))  # strictly positive SE
stopifnot("var_method" %in% names(frpm_2024))  # calculation method recorded
stopifnot(all(frpm_2024$var_method == "arcsine_vst"))  # this interior example
```

# Continue with your own data

You now have a `sitemix_estimates` tibble for one indicator in
one year. The next steps depend on what your data look like:

-   If your data are **not student rows** (you have sufficient
    counts or a publisher CSV), read
    [Choosing an input format](a2-input-formats.html) and pick the
    matching wrapper.
-   For transformations and proportions at 0 or 1, read
    [Estimating proportions for a binary indicator](a3-scenario-binomial.html).
-   If you have **multiple indicators** (e.g., FRPM + SNAP +
    WIC), read
    [Analyzing overlapping indicators and mutually exclusive categories](a4-multivariate-multinomial.html).
-   When you are ready to use the estimates in another analysis,
    read [Using estimates and covariance matrices in further analyses](a8-downstream-workflows.html).
-   For the **formal foundations** (the sampling model and
    the delta-method SE derivation), read
    [Sampling uncertainty in site-level proportions](m1-statistical-foundations.html)
    and [Binomial standard errors and transformations](m2-scalar-se-binomial.html).

# Session info

```{r session-info, echo = FALSE, results = "hide"}
# Print session info in a real render
```
