---
title: "Analyzing overlapping indicators and mutually exclusive categories"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  Choose between overlapping binary indicators and mutually exclusive
  categories, then estimate their site-level proportions and covariance.
  Examples show how to interpret raw-scale matrices alongside transformed
  standard errors and why multinomial covariance can be singular.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Analyzing overlapping indicators and mutually exclusive categories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(prek_sim, package = "sitemix")
```

Estimates for several indicators measured on the same students can have
dependent sampling errors. To retain their covariance, set `vjt = TRUE`.
First choose the estimation family from how the variables are recorded.

The examples use 50 simulated sites in `prek_sim` and an added simulated
language category. They describe no real children or programs and are not
empirical Pre-K findings. See [Choosing an input format](a2-input-formats.html)
for the distinction between student records, counts, and published aggregates.

# Choose from the observed variables

| Data | Family and argument | Information used for covariance |
|:--|:--|:--|
| Several binary indicators on the same students; a student may have more than one | `family = "multivariate"`, `indicators = c(...)` (B) | Marginal and pairwise co-occurrence counts from the same observations |
| One categorical variable; every retained student belongs to exactly one category | `family = "multinomial"`, `indicator = "..."` (C) | Category counts that sum to the shared site-year denominator |

FRPM, SNAP, WIC, and TANF are overlapping binary indicators: a student can
have any combination of these statuses. Primary language, recorded as
exactly one of English, Spanish, or Other, is a multinomial category.
The categories must be mutually exclusive and together cover all retained
observations. A set of categories with omitted observations or overlapping
membership does not meet that requirement.

For B, the same student records supply the marginal proportions and the
proportion positive for each pair of indicators. A pair's covariance can
be positive, negative, or zero; the possibility of overlapping membership
does not determine its value. With C, the category proportions sum to one,
and the covariance matrix must respect that constraint.

Both families also accept complete sufficient counts through
`sm_estimate_from_counts()`. B requires every marginal and pairwise
co-occurrence count from a common sample, in the supplied indicator order.
Count-based covariance is supported for two or three overlapping indicators;
use student records for `vjt = TRUE` with four or more. C requires every
category count, with an exact sum of `n_jt` per site-year. A category with
zero count must remain in the specified category set.

Separate published marginal counts, without the joint counts, use the D1
aggregate method. Equal denominators alone do not establish that the same
students were observed. D1 working independence does not recover joint
counts and is not a multinomial composition; see
[Estimating proportions from published aggregates](a5-published-aggregates.html).

# Estimate overlapping binary indicators

Use `indicators` to name the four binary columns and `vjt = TRUE` to
request their joint covariance:

```{r scenario-b}
benefits_2024 <- sm_estimate(
  subset(prek_sim, year == 2024),
  family     = "multivariate",
  indicators = c("frpm", "snap", "wic", "tanf"),
  vjt        = TRUE
)
head(benefits_2024, 4)
```

The result has four rows per site-year, one per indicator. Each row carries
the same site-year covariance object in `V`; these are repeated copies of
one group matrix, not four independent matrices. The first object is for
site `S001` in 2024:

```{r scenario-b-vcov}
V_S001 <- benefits_2024$V[[1L]]
round(as.matrix(V_S001), 4)
V_S001$vcov_scale
V_S001$vcov_method
```

The matrix rows and columns identify the indicators, and `vcov_method`
is `"sur"`, the package's label for covariance calculated from their
joint observations. `vcov_scale = "raw"` means that it describes the
raw proportions in `theta_raw`. This scale is a package choice; the
matrix is not automatically transformed with the row estimates.

By default, `theta_hat` and `se` use the arcsine scale. Consequently,
`sqrt(diag(V))` should be compared with `se_raw` for B, not with the
transformed `se`. Keep `vcov_scale` and the matrix's indicator order
when using the result in a joint analysis.

# Estimate mutually exclusive categories

`prek_sim` has no categorical language variable. The following code
simulates one label for every student, using the fixed seed for this
article. The specified probabilities are example settings.

```{r scenario-c-data}
language_data <- transform(
  prek_sim,
  primary_language = sample(
    c("english", "spanish", "other"),
    nrow(prek_sim),
    replace = TRUE,
    prob    = c(0.70, 0.25, 0.05)
  )
)
language_data <- language_data[
  c("student_id", "site_id", "year", "primary_language")
]
head(language_data, 5)
```

Use `indicator` to name this single categorical column and
`family = "multinomial"` to estimate all of its category proportions:

```{r scenario-c}
language_2024 <- sm_estimate(
  subset(language_data, year == 2024),
  family    = "multinomial",
  indicator = "primary_language",
  vjt       = TRUE
)
head(language_2024, 6)
```

The result has three rows per site-year, one per category. Within each
site-year, the raw proportions in `theta_raw` sum to one. The covariance
matrix for `S001` is:

```{r scenario-c-vcov}
V_lang_S001 <- as.matrix(language_2024$V[[1L]])
round(V_lang_S001, 4)
```

This matrix also uses the raw-proportion scale, while the default
`theta_hat` and `se` remain on the arcsine scale. Its method label is
`"multinomial"`. Because the category proportions have a fixed sum,
the covariance row sums are zero up to numerical rounding:

```{r simplex-check}
rowSums(V_lang_S001)
```

This sum-to-one constraint is called the **simplex** constraint. It makes
the covariance matrix singular: the category estimates cannot vary
independently. Singularity is expected here and does not itself indicate
an invalid matrix.

If `S` categories have positive observed counts, the matrix's
`positive_support` field is `S` and `matrix_rank` records `S - 1`.
This is the analytic support rank. It remains the recorded value when
an exact census under a supplied finite-population design makes the actual
covariance matrix zero. A census therefore has numerical rank zero even
if `matrix_rank` is positive.

A category with no observations at a site has an exact zero row and
column in its multinomial matrix. This zero in the estimated matrix does
not establish that membership in the category is impossible.
With the default Wilson boundary treatment and no finite-population
correction, that category still has a positive scalar `se_raw`.
Its squared value does not replace the zero matrix diagonal, which is
needed to preserve the simplex constraint. The `diag_contract` field
records this boundary exception. At an exact census, the finite-population
correction makes sampling uncertainty zero. See
[Covariance for multinomial proportions](m4-multinomial-simplex.html) for the formulas
and the distinction between scalar boundary uncertainty and the joint matrix.

# Use individual estimates or joint covariance

For analyses that treat each indicator separately, select the estimates,
standard errors, scale, and calculation method from this complete example:

```{r split}
scalar_inputs <- split(
  as.data.frame(benefits_2024[, c(
    "site_id", "year", "indicator", "theta_hat", "se",
    "estimate_scale", "var_method"
  )]),
  benefits_2024$indicator
)
names(scalar_inputs)
```

These tables keep `theta_hat` with its matching `se` and `estimate_scale`.
They omit cross-indicator covariance, so they are not sufficient for
comparisons or combinations that require joint uncertainty. Use the
complete site-year `V` matrix and its corresponding raw estimates for a
raw-scale joint analysis. Match the indicator order before calculating
contrasts. A linear contrast uses the covariance matrix without requiring
its inverse, so a valid singular multinomial matrix can still be used.
See [Using estimates and covariance matrices in further analyses](a8-downstream-workflows.html) for the
required checks and extraction examples.

# Check the example results

The following checks confirm the expected row counts, methods, raw B
covariance scale, positive semidefiniteness of the first B matrix, and
the zero-sum constraint for the first C matrix:

```{r audit}
stopifnot(nrow(benefits_2024) == 200L)  # 50 sites × 4 indicators
stopifnot(inherits(benefits_2024$V[[1]], "sm_vcov"))
stopifnot(benefits_2024$V[[1]]$vcov_method == "sur")
stopifnot(benefits_2024$V[[1]]$vcov_scale == "raw")
stopifnot(min(eigen(as.matrix(benefits_2024$V[[1]]),
                    symmetric = TRUE)$values) >= -1e-10)
stopifnot(nrow(language_2024) == 150L)  # 50 sites × 3 categories
stopifnot(language_2024$V[[1]]$vcov_method == "multinomial")
stopifnot(max(abs(rowSums(as.matrix(language_2024$V[[1]])))) < 1e-10)
```

# Examine the uncertainty in more detail

-   [Checking estimates and handling suppressed data](a6-diagnostics-and-suppression.html)
    for `level = "vcov"` diagnostics including the smallest eigenvalue,
    PSD tolerance, and estimate/covariance scale compatibility.
-   [Using estimates and covariance matrices in further analyses](a8-downstream-workflows.html) for using
    estimates and joint covariance in another analysis.
-   [Covariance for overlapping binary indicators](m3-multivariate-sur-covariance.html)
    for the formal SUR derivation.
-   [Covariance for multinomial proportions](m4-multinomial-simplex.html) for
    the formal simplex-covariance derivation.
