---
title: "Experimental models for variance smoothing"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  `sm_smooth_variance()` fits a relationship between supplied variance
  estimates and their denominators. This article describes the log-linear
  and additive models, the raw-scale offset, and the assumptions behind
  transforming fitted log-variances back to standard errors.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Experimental models for variance smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
```

# 1. What the smoother estimates

`sm_smooth_variance()` models the logarithm of the variance estimates
already present in a `sitemix_estimates` object. It fits a relationship
across eligible rows and converts the fitted values to alternative SEs.
This generalized variance function (GVF) is experimental and must be
requested explicitly. By default, it adds `se_smoothed` or
`se_raw_smoothed`, together with `var_method_smoothed`, and preserves
the original SEs.

A smaller denominator can itself explain a larger SE. For example,
the uncorrected arcsine working SE without a finite-population correction
is $s_{jt} = 1/(2\sqrt{n_{jt}})$: it is about `0.224` at $n = 5$
and `0.050` at $n = 100$. Fitting this denominator relationship does
not establish that the original uncertainty needs correction.

The response here is a log-variance estimate. In contrast, the
Fay--Herriot area-level model relates direct outcome estimates to
unobserved area means and a regression model, with specified sampling
variances [@fay_herriot_1979, p. 271]. `sm_smooth_variance()` leaves
`theta_hat` and `theta_raw` unchanged and does not estimate area-level
posterior means. The Fay--Herriot paper supplies a comparison of model
targets, rather than a derivation of this variance smoother.

The worked example in Section 7 uses `prek_sim`, a fully simulated
50-site pre-kindergarten panel. Section 8 describes a separate
simulation that generates binomial counts. Neither uses observations
of real children, sites, or programs.

# 2. Variances and predictors

Let $s_{jt}$ denote the selected input SE and $v_{jt}=s_{jt}^2$ its
variance estimate. Here $v_{jt}$ is an observed input to a regression,
not the unknown sampling variance itself. Only rows with a positive,
finite selected SE and denominator can enter the log-variance model;
the additional row-selection rules are described below.

| Quantity | Model variable | Meaning |
|:---------|:---------------|:--------|
| $\log v_{jt}$ | `log_var` | Log of the squared input `se` or `se_raw` |
| $\log n_{jt}$ | `log_n` | Log of the row denominator |
| $\ell^{\mathrm{off}}_{jt}=\log[p^*_{jt}(1-p^*_{jt})]$ | `p_offset` | Log offset for the default `scale = "se_raw"` formula |
| $\tilde s_{jt}$ | `se_smoothed` or `se_raw_smoothed` | SE obtained from the fitted log-variance |

The probability $p^*_{jt}$ used in the offset is defined in Section 4.
The `scale` argument selects the input column. In particular,
`scale = "se"` can select raw-scale SEs when the estimates were made
with `vst = "none"`; it does not automatically select the raw-offset
formula.

# 3. The default log-linear model

With `method = "loglinear"`, `scale = "se"`, `by = NULL`, and no
custom formula or weights, the function fits an ordinary least-squares
regression using `stats::lm()`:

$$
\log v_{jt} \;=\; \beta_0 + \beta_1 \log n_{jt} + \varepsilon_{jt}.
\tag{M6.1}
$$

For `scale = "se_raw"`, the default formula adds an offset with its
coefficient fixed at one:

$$
\log v^{\mathrm{raw}}_{jt} \;=\; \beta_0 + \beta_1 \log n_{jt} +
\ell^{\mathrm{off}}_{jt} +
\varepsilon_{jt}.
\tag{M6.2}
$$

The corresponding formulas are `log_var ~ log_n` and
`log_var ~ log_n + offset(p_offset)`. Columns supplied through `by`
enter as factor main effects; they do not cause separate fits. For
uncorrected arcsine inputs without FPC, $v=1/(4n)$ gives a slope of
$-1$. Corrections already present in the input SE can change that
relationship.

## Returning from log-variance to SE

Write $\hat\eta_{jt}$ for the fitted value on the log-variance scale,
including any formula offset. With the default `bias_correct = TRUE`,
the implementation calculates

$$
\tilde v_{jt}=\exp\!\left(\hat\eta_{jt}
  +\tfrac12\hat\sigma_\varepsilon^2\right),
\qquad
\tilde s_{jt}=\sqrt{\tilde v_{jt}}.
$$

The half-variance term follows from a log-normal model: if the log-scale
error is conditionally normal with mean zero and a common variance
$\sigma_\varepsilon^2$, then
$E(v_{jt}\mid\text{predictors})=
\exp(\eta_{jt}+\sigma_\varepsilon^2/2)$.
This is the log-normal mean identity documented in
[R's log-normal distribution reference](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Lognormal.html).
Jensen's inequality alone does not determine this particular correction.

The package substitutes `stats::sigma(fit)^2` for a linear model or
`summary(fit)$scale` for a GAM. Setting `bias_correct = FALSE` omits
the half-variance term; under the same log-normal model,
$\exp(\eta_{jt})$ is the conditional median of the variance input.
These calculations treat the fitted model parameters as fixed. They
do not account for uncertainty in the fitted regression or make the
result an unbiased estimate of the true sampling variance. A common
half-variance correction also needs separate justification if the
log-scale errors are non-normal or their variance changes across rows.

# 4. The Wilson-centered raw-scale offset

The binomial variance has a rate-dependent factor $p(1-p)$, which
motivates the offset in (M6.2). Substituting a boundary estimate
$\hat\pi\in\{0,1\}$ makes its logarithm undefined. For flagged boundary
rows, the implementation instead uses

$$
p^*_{jt} \;=\; \frac{\hat\pi_{jt} + z^2/(2n_{jt})}{1 + z^2/n_{jt}}
\;=\; \frac{C_{jt} + z^2/2}{n_{jt} + z^2}
$$

with $z=\Phi^{-1}(0.975)$. This is the Wilson score center
[@wilson_1927]; taking $z\approx2$ gives the related plus-four center
[@agresti_coull_1998]. Interior rows use their raw estimates, and the
result is clamped away from zero and one at machine precision.
Using this center for the offset is a package modeling choice. It
does not replace the response variance: an input SE of zero still
cannot enter the log-variance regression.

The model-frame column is not this probability itself. It is
`p_offset = log(p_star * (1 - p_star))`, the term
$\ell^{\mathrm{off}}_{jt}$ in (M6.2). A custom formula should therefore use
`offset(p_offset)`, not apply a second probability-to-log transformation.

## Inputs with a finite-population correction

Smoothing consumes the SEs already present in the input. If those SEs were
adjusted for simple random sampling without replacement (SRSWOR), the
fitted response contains that adjustment. For the plug-in arcsine
calculation with `anscombe = FALSE`, let $N>n$ be the finite population size and
$q=(N-n)/(N-1)>0$. The supplied variance is $q/(4n)$, so

$$
\log(s^2)=-\log 4-\log n+\log q,
$$

Variation in $q$ across rows can change the fitted denominator slope;
the default smoothing formula does not add an FPC offset. This equation
does not cover Anscombe or bias-corrected inputs. Those inputs retain
their own SE calculation, described in
[Binomial standard errors and transformations](m2-scalar-se-binomial.html). At an exact census, the SE is zero
and the row is excluded from the log-variance fit.

By default, the returned object keeps `se`, `se_raw`, and the sampling
information used to calculate them. If overwrite is requested and
allowed, the original SEs are saved in `se_pre_smoothing` or
`se_raw_pre_smoothing`. The unchanged FPC fields continue to describe
the original design calculation; they do not establish a new
design-based interpretation for the smoothed SE.

# 5. An additive model for a curved relationship

With `method = "gam"`, the default `scale = "se"`, `by = NULL`
model replaces the straight-line term with a smooth function:

$$
\log v_{jt} \;=\; \beta_0 + f(\log n_{jt}) + \varepsilon_{jt}.
\tag{M6.3}
$$

The default call uses `mgcv::gam()` with `s(log_n)`, a Gaussian
family with an identity link, and `method = "REML"` to estimate
smoothness. The smooth uses a penalized thin plate regression spline,
the default basis for [`mgcv::s()`](https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/s.html).
The penalty controls how much curvature is fitted. For
`scale = "se_raw"`, the formula also includes `offset(p_offset)`;
`by` columns again add factor main effects.

Wood's treatment of additive models [@wood_2017] and the
[`mgcv::gam()` documentation](https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/gam.html)
describe the fitting method. Applying it to these log-variance inputs
is the package's choice, and those sources do not establish an
improvement in sitemix SEs or interval coverage. The conversion back to variance
above uses the estimated scale parameter of the GAM; its log-normal
interpretation applies to the default Gaussian model and requires
reconsideration if that family or error structure is changed.

# 6. Which rows are fitted

After validating the input object, the function selects rows that are
not suppressed and have a finite, positive selected SE and denominator.
With the default `scope = "all"`, every such row is eligible.
`scope = "tier2"` further restricts the fit to $11\le n\le29$;
this interval is fixed and does not follow `accountability_n` or
`flag_below_accountability`. If supplied, `min_n` also requires
`n >= min_n`. The Tier 2 label in `sm_suppression_report()`
uses a separate, adjustable reporting threshold.

The fitted SEs replace values only in eligible rows of the added
column. Other rows retain their input values, including zero SEs for
exact censuses and missing SEs for suppressed rows. Thus a successful
fit does not imply that every row in the result has a positive or
finite SE. Objects containing non-identified suppression-sensitivity
estimates are rejected before fitting.

At least `min_rows` eligible rows are needed; the default is `50L`.
If fewer qualify, the function skips fitting, copies the selected
input SE column to the alternative column, and records
`status = "skipped"` in both summary attributes. It performs no
overwrite, even with `overwrite = TRUE`, and adds no new residuals,
fitted model, or pre-smoothing snapshots.

## Warnings and fit diagnostics

Four warning classes describe distinct conditions:

- `sitemix_warning_smoother_skipped`: fewer than `min_rows` rows
  qualified. Inspect the row selection before interpreting the
  alternative column as fitted values.
- `sitemix_warning_smoother_multi_year_default`: eligible rows
  span multiple years and `by = NULL`. Use `by = "year"` for a year
  fixed effect in one fit, or call the function on separate yearly
  subsets for separate fits.
- `sitemix_warning_unexpected_slope`: a loglinear fit using
  `scale = "se"` on arcsine or Anscombe-arcsine rows has a finite
  `log_n` coefficient more than 0.15 from $-1$. Interpret the slope
  alongside any corrections already included in the input SEs.
- `sitemix_warning_raw_scale_smoothing`: the first request for
  `scale = "se_raw"` after loading the package. The default formula
  includes `offset(p_offset)`; a custom formula needs its own
  appropriate rate-dependent term.

The `smoother_fit_summary` attribute records the current fit status,
formula, eligible-row count, and relationship to any covariance
matrices. After fitting, it also records rank, convergence, prediction
checks, and model summaries. The `smoothing` attribute records the
settings and eligible row indices. With `return_diagnostics = TRUE`,
a completed fit adds `residual_log_var` and the `smoother_fit`
attribute. Check the current status first: a skipped call adds no
new fitted model or residuals.

A custom formula must use `log_var` as its response and the model
variables and terms documented in `?sm_smooth_variance`. The fitted
model must have full rank, positive residual degrees of freedom, and
finite predictions for all eligible rows; a GAM must also converge.
Failure raises an error instead of substituting a fallback fit.

## Overwrite and covariance matrices

When enough rows qualify to fit, `overwrite = TRUE` is rejected with
`sitemix_error_smoothing_v_stale` if an eligible row has a `V` matrix
on the selected SE scale. Keeping that matrix while replacing its
SE could make the diagonal inconsistent. A matrix on a different
scale remains unchanged, and its relation to the selected scale is
recorded in `smoothing$v`. The insufficient-row skip described above
occurs before the overwrite rejection and leaves both SEs and `V`
unchanged.

If overwrite is allowed, `scale = "se"` replaces eligible `se`
values. With `scale = "se_raw"`, it replaces eligible `se_raw`
values and also updates `se` for rows with `estimate_scale = "none"`.
Transformed `se` values are preserved in that raw-column operation.
`var_method` changes where `se` changes; `var_method_smoothed`
describes the alternative SEs.

# 7. Checking the returned columns

This example fits the two default loglinear formulas to the 2024
FRPM estimates and checks that the original columns are preserved.
All rows in this example have finite SEs; the checks on the whole
column below therefore apply to this example, while the general
rules for missing or zero input values are given in Section 6.
The calls explicitly use `min_rows = 2`, rather than the default 50.

```{r invariants}
data(prek_sim, package = "sitemix")
est <- sm_estimate(
  subset(prek_sim, year == 2024),
  family = "binomial", indicator = "frpm"
)
est_s <- sm_smooth_variance(
  est, method = "loglinear", return_diagnostics = TRUE, min_rows = 2
)

# Check the smoothed SEs in this eligible-row example.
stopifnot(all(is.finite(est_s$se_smoothed)))
stopifnot(all(est_s$se_smoothed >= 0))

# The default transformed-scale formula has no offset.
fit_formula <- formula(attr(est_s, "smoother_fit"))
stopifnot(identical(deparse(fit_formula), "log_var ~ log_n"))

est_raw <- capture_expected_sitemix_warning(
  sm_smooth_variance(
    est, method = "loglinear", scale = "se_raw",
    return_diagnostics = TRUE, min_rows = 2
  ),
  "sitemix_warning_raw_scale_smoothing"
)
raw_formula <- formula(attr(est_raw, "smoother_fit"))
stopifnot(identical(deparse(raw_formula),
                    "log_var ~ log_n + offset(p_offset)"))

# The default call retains the original SE column.
stopifnot(all.equal(est_s$se, est$se, tolerance = 1e-12))

# Keep the original method label and record the smoothed alternative.
stopifnot(identical(est_s$var_method, est$var_method))
stopifnot("var_method_smoothed" %in% names(est_s))

# Retain the original SEs and finite-population fields.
counts <- readRDS(system.file("extdata", "prek_sim_counts.rds",
                              package = "sitemix"))
fpc_counts <- counts[counts$year == 2024,
                     c("site_id", "year", "n_jt", "c_jt_frpm")]
est_fpc <- sm_estimate_from_counts(
  fpc_counts, family = "binomial", indicator = "frpm",
  fpc = 2 * fpc_counts$n_jt, min_n = 1L
)
est_fpc_s <- sm_smooth_variance(est_fpc, method = "loglinear", min_rows = 2L)
fpc_fields <- c(
  "population_size", "sampling_fraction",
  "fpc_variance_multiplier", "fpc_se_multiplier",
  "variance_multiplier_applied", "se_multiplier_applied",
  "sampling_design", "variance_rule"
)
stopifnot(identical(est_fpc_s$se, est_fpc$se))
stopifnot(identical(est_fpc_s$se_raw, est_fpc$se_raw))
stopifnot(all(vapply(fpc_fields, function(field) {
  identical(est_fpc_s[[field]], est_fpc[[field]])
}, logical(1))))
```

The transformed-scale fit uses all 50 rows and reproduces the original
uncorrected arcsine SEs to numerical precision. That follows from their
$1/(2\sqrt n)$ relationship; it is not evidence that smoothing improves
uncertainty estimation. The raw-column fit uses the additional rate
offset. Both fits preserve the original SE and calculation-label
columns by default.

The final part sets each simulated finite population to twice the
sample size and checks the eight returned sampling fields. Preserving
those fields records the original SRSWOR calculation. It does not test
the statistical performance of smoothed SEs under finite-population
sampling.

# 8. What the fixed simulation shows

The maintainer script `inst/scripts/audit-smoothing-simulation.R`
compares unsmoothed SEs with loglinear and GAM alternatives. The
recorded full run used seed `20260712` and 60 replicates. Each
replicate generated independent binomial counts on an 84-row grid:
12 denominators from 8 to 150 crossed with seven true rates from
0.02 to 0.98. The four reporting strata crossed $n\le20$ versus
larger $n$ with near-boundary rates ($p\le0.10$ or $p\ge0.90$)
versus interior rates.

The estimates used `vst = "arcsine"`, `boundary_method = "wilson_floor"`,
and `min_n = 1`, without FPC. Each smoother used all eligible rows,
`min_rows = 50`, `bias_correct = TRUE`, and `overwrite = FALSE`.
The variance targets, calculated separately from the package helpers,
were the arcsine working approximation $1/(4n)$ and the exact binomial
proportion variance $p(1-p)/n$. The first target is not the exact
finite-sample variance of the transformed estimator.

Relative variance MSE means the average of
$[(\hat v-v)/v]^2$, using the target $v$ on the corresponding scale.
Coverage measures symmetric normal intervals for $p$ or
$\arcsin\sqrt p$, using the same point estimates with each candidate
SE. The weight measure is the average total-variation distance
between normalized $1/\hat v$ weights and normalized target $1/v$
weights; it does not evaluate a fitted hierarchical model.

The recorded raw-scale results were:

| Method | Relative variance MSE | 95% interval coverage | Weight distance |
|:-------|---------------------:|---------------------:|----------------:|
| Unsmoothed | 0.7404 | 0.9373 | 0.1984 |
| Loglinear | 1.5980 | 0.9452 | 0.2042 |
| GAM | 1.5889 | 0.9442 | 0.2047 |

Both raw-scale smoothers moved pooled coverage closer to 95%, while
more than doubling relative variance MSE and slightly increasing
weight distortion. On the transformed scale, all three methods
reproduced the deterministic variance target to numerical precision,
and all had coverage about 0.9101. Agreement with $1/(4n)$ therefore
did not imply nominal coverage.

The criteria specified in advance by the maintainer required all of the following:
at least 10% lower pooled relative MSE; no stratum with more than 10%
relative-MSE deterioration; increases in absolute coverage error of
at most 0.01 overall and 0.03 in every stratum; and at least 5% lower
weight distance. For effectively zero baseline errors, matching that
baseline counted as equality, not improvement. None of the four
scale-by-method candidates met all criteria.

These results do not support a general recommendation to smooth or
overwrite SEs. They also do not establish that smoothing is inferior
under other data-generating processes. The simulation contains no
finite-population sampling, suppression, covariance matrices,
multiple-year trends, or downstream hierarchical model. The function
remains an optional, experimental way to compare alternative SEs.

For worked comparisons with a GAM and Fréchet sensitivity results,
see [Variance smoothing and Fréchet sensitivity analysis](a7-variance-smoothing-and-frechet.html). For the scope of the
Fréchet calculation itself, see [Pairwise Fréchet bounds and projected dependence scenarios](m7-frechet-envelope-theory.html).

# References
