---
title: "Pairwise Fréchet bounds and projected dependence scenarios"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  `sm_frechet_envelope()` explores dependence when published proportions
  omit joint counts. This article derives the pairwise probability and
  covariance intervals, explains when they have a formal interpretation,
  and distinguishes them from the returned dependence stress matrices.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Pairwise Fréchet bounds and projected dependence scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
```

# 1. What the marginals identify

The D1 aggregate path accepts published marginal counts without joint
counts; the numerators and denominators yield marginal proportions.
For example, knowing the FRPM and SNAP rates
among the same students does not usually tell us how many students
belong to both groups. A joint proportion, and hence a covariance,
can vary while those two marginal proportions remain fixed.

`sm_frechet_envelope()` reports the pairwise ranges in
`raw_pairwise_intervals`. Their interpretation depends on whether
the indicators describe the same sampled units or subgroup-conditional
rates. The function requires that choice explicitly. The formal
common-unit case is called D1a; D1b is a heuristic calculation for
subgroup-conditional rates. Section 4 gives the required inputs.

The worked example below uses counts from `prek_sim`, the package's
fully simulated 50-site pre-kindergarten panel. It does not describe
real students, sites, or programs.

# 2. The pairwise joint-probability interval

Let $X$ and $Y$ be binary indicators on a common unit, with
$p=P(X=1)$, $q=P(Y=1)$, and $r=P(X=1,Y=1)$. Their joint
probabilities must form the following table:

| | $Y=1$ | $Y=0$ | Total |
|:--|--:|--:|--:|
| $X=1$ | $r$ | $p-r$ | $p$ |
| $X=0$ | $q-r$ | $1-p-q+r$ | $1-p$ |
| Total | $q$ | $1-q$ | $1$ |

All four cells must be nonnegative, giving

$$
\max(0,p+q-1) \;\le\; r \;\le\; \min(p,q).
\tag{M7.1}
$$

Conversely, every $r$ in this interval gives a nonnegative table with
the stated margins and total one. Thus both endpoints are attainable
for a pair of binary variables. These are the classical
Fréchet--Hoeffding bounds [@frechet_1951; @hoeffding_1940]. The
general copula and joint-distribution forms appear in
@nelsen2006introduction, Theorem 2.2.3 (p. 11) and Section 2.5,
equation (2.5.1) (p. 30). The table above provides a direct derivation
for the binary case used here.

For example, $p=0.25$ and $q=0.50$ allow $0\le r\le0.25$.
The endpoints have individual-level covariance $r-p\,q=-0.125$ and
$0.125$. If either marginal is zero or one, the joint interval
collapses and the covariance is zero; the joint remains unidentified
only when the interval has positive width.

The package substitutes the supplied `theta_raw` values for $p$ and
$q$. The returned interval is conditional on those margins. It is
not a confidence interval for an unknown population joint probability
and does not account for sampling error in the supplied margins.

# 3. From a joint probability to sampling covariance

Suppose $(X_i,Y_i)$ are independent, identically distributed pairs
for the same $n$ sampled units. The raw sample proportions have
covariance

$$
\operatorname{Cov}(\bar X,\bar Y)
=\frac{1}{n^2}\sum_{i=1}^n\operatorname{Cov}(X_i,Y_i)
=\frac{r-p\,q}{n}.
$$

Independence across sampled units removes the cross-unit covariance
terms. Dependence between $X_i$ and $Y_i$ within a unit is allowed.
Substituting each endpoint from (M7.1) gives the D1a raw pairwise
covariance interval. With $p=0.25$, $q=0.50$, and $n=20$, it is
$[-0.00625,0.00625]$.

The table `raw_pairwise_intervals` records `p_1`, `p_2`,
`n_common`, and the joint-probability, correlation, and covariance
endpoints. Its covariance endpoints use the raw probability scale,
even when the input row estimates and `V` use an arcsine scale.
For D1a, `covariance_construction` is
`"formal_iid_pairwise_covariance_over_n"` and `interval_scope` is
`"formal_raw_pairwise_interval"`.

For interior probabilities, the corresponding correlation is

$$
\rho=\frac{r-p\,q}{\sqrt{p(1-p)q(1-q)}}.
$$

Correlation is undefined for a degenerate marginal. The implementation
records `NA` correlation endpoints when the denominator is smaller
than machine epsilon. The direct D1a covariance formula still applies;
at an exact zero/one marginal its two endpoints are zero.

# 4. Choosing D1a or D1b

Supply `population_regime = "d1a"` or `"d1b"`; omission raises
`sitemix_error_population_regime_required`. The input must come from
the D1 aggregate path and contain complete, identified marginals.
Both `vjt = FALSE` and `vjt = TRUE` results are accepted. Suppressed
missing rows and suppression-sensitivity rows are rejected.

## D1a: common units and an IID plug-in calculation

Formal D1a requires the following information to have been recorded
by the D1 estimator:

- `sampling_relation = "same_units"`, with every site-year group
  recorded as D1a with common denominators;
- one finite, positive denominator per selected site-year and
  `n_eff = n`, requiring `anscombe = FALSE`;
- the IID plug-in variance calculation, without FPC or
  `binomial_bc`. If `V` is present, its sampling and variance
  information must agree with these conditions.

Equal denominators alone do not establish common sampled units.
The label must reflect the publisher's information. These checks
limit the formal covariance result to the `/n` calculation above;
they do not add sampling uncertainty to the marginal inputs.

For this regime, `V_independence` has diagonal $p_k(1-p_k)/n$,
reconstructed from the marginal values. This raw plug-in diagonal
can differ from an input `se_raw^2` that used a boundary fallback.
It is also distinct from a transformed-scale row `V`.

## D1b: subgroup-conditional rates

Consider FRPM rates among ELL students and among non-ELL students.
These refer to different sets of students, so treating the two rates
as marginals of one common-unit binary vector is an additional
hypothetical construction. The D1b calculation uses that construction
to explore dependence; its ranges are not formal joint-probability
or sampling-covariance bounds for those subgroup rates.

D1b accepts inputs recording `sampling_relation = "different_units"`
or `"unknown"`. A same-unit D1a input cannot be relabeled as D1b
in this call. `subgroup_conditional_action = "warn"` is the default:
it proceeds with `sitemix_warning_frechet_d1b_heuristic`. Use
`"allow"` to acknowledge the heuristic and proceed silently, or
`"error"` to stop with `sitemix_error_frechet_d1b_disallowed`.
This argument is ignored for D1a.

There is no justified common denominator for the D1b construction,
so `n_common = NA`, even if the supplied denominators happen to be
equal. Instead, the function scales each pairwise correlation endpoint
by the two input raw-scale SEs:

$$
V_{kk'}^{\mathrm{corner}}
=\rho_{kk'}^{\mathrm{corner}}s_k s_{k'}.
$$

Here $s_k$ is the row's `se_raw`. The output identifies this choice
with `covariance_construction = "heuristic_se_scaled_pairwise_correlation"`
and `interval_scope = "heuristic_pairwise_stress_range"`. Where the
correlation endpoint is undefined, the implementation uses zero for
the heuristic off-diagonal. That numerical convention does not
identify the actual covariance between subgroup estimates.

# 5. From pairwise intervals to matrices

The function assembles two matrices, one from all lower pairwise
covariance endpoints and one from all upper endpoints. It stores them
as `unprojected_negative_dependence_corner` and
`unprojected_positive_dependence_corner`. A covariance matrix must
be positive semidefinite (PSD): every linear combination of its
variables must have nonnegative variance. Pairwise feasibility alone
does not ensure this property for the assembled lower corner.

For example, consider three common-unit indicators with marginal
probabilities all equal to 0.5. Each lower joint endpoint is zero,
so the lower covariance corner has diagonal $1/(4n)$ and every
off-diagonal equal to $-1/(4n)$. Its smallest eigenvalue is
$-1/(4n)$, so it is indefinite. The three pairs cannot all attain
their lower endpoints together.

The upper corner has a different property in formal D1a. Let
$U$ be uniform on $[0,1]$ and define $X_k=1\{U\le p_k\}$.
Then $P(X_k=1,X_l=1)=\min(p_k,p_l)$ for every pair
simultaneously. This construction gives a valid joint distribution,
and the covariance matrix of means from $n$ independent copies is
PSD. Thus the common-unit all-upper corner is attainable and PSD
in exact arithmetic. Its largest pairwise entries do not make it
an upper variance bound for every linear contrast.

## PSD is necessary but not sufficient

A PSD adjustment ensures nonnegative variances for linear
combinations, but it does not establish a joint binary distribution
with the supplied marginals. For a simple example, three Bernoulli
variables with probabilities 0.5 and pairwise correlations $-0.5$
would have a PSD correlation matrix. Yet their sum would have mean
1.5 and variance zero, which is impossible for an integer-valued
sum. Each pair separately satisfies its probability bounds.

The returned `projected_negative_dependence_stress` and
`projected_positive_dependence_stress` therefore retain their role
as dependence scenarios. For `K = 2` in formal D1a, their
off-diagonals do equal the two formal covariance endpoints: the
matrices are returned unchanged from their source corners. That
special-case equality does not turn the general matrix fields into
multivariate bounds.

# 6. Projection methods and diagnostics

For `K <= 2`, neither algorithm runs, even if a shrinkage weight is
supplied. For larger matrices, an already-PSD corner is also kept
unchanged unless a fixed `shrink_alpha` was requested. When an
adjustment is needed, the two methods preserve the source diagonal:
the raw plug-in diagonal in D1a and the input raw-SE diagonal in D1b.

## Higham method

Higham's alternating-projections method finds a nearest correlation
matrix using the PSD and unit-diagonal constraints, with the correction
selected by `doDykstra = TRUE` [@higham_2002]. The default
`psd_method = "higham"` uses
[`Matrix::nearPD()`](https://stat.ethz.ch/R-manual/R-devel/library/Matrix/html/nearPD.html)
with package-specific settings. It operates on the covariance corner
with `corr = FALSE` and `keepDiag = TRUE`, preserving its original
diagonal rather than setting the diagonal to one.

The default call uses `doDykstra = TRUE` and enables the additional
eigenvalue adjustment through `do2eigen = TRUE`. Its eigenvalue and
convergence tolerances equal `psd_tol`, as does the default
`posd.tol`; the convergence norm is the infinity norm (`"I"`).
Finite stopping tolerances and the final eigenvalue adjustment mean
the returned matrix is a numerical covariance adjustment.

After `nearPD()` returns, sitemix restores the source diagonal,
makes the result symmetric, and checks its smallest eigenvalue. A fit
reported as failing to converge raises
`sitemix_error_vcov_projection_nonconvergence`. The final PSD check
allows a small numerical tolerance. Write $t$ for `psd_tol` and $d$ for
`.Machine$double.xmin`, the smallest positive normalized double:

$$
\lambda_{\min}(V)\ge
-\left(t\max_i|\lambda_i(V)|+K\,d\right).
$$

The default `psd_tol` is `1e-8`, and it must satisfy
`0 < psd_tol <= sqrt(.Machine$double.eps)`. The small absolute term
handles the machine range; the main tolerance scales with the
matrix eigenvalues. This is a numerical check, not permission to
treat an appreciably indefinite matrix as a covariance matrix.

Supported `...` controls are `base.matrix`, `doSym`, `doDykstra`,
`posd.tol`, and `conv.norm.type`; the applied values are stored in
`projection_config`. For example, `doDykstra = FALSE` turns off that
correction and changes the iteration. Other settings such as `keepDiag`, `corr`,
`eig.tol`, `conv.tol`, and `maxit` cannot be overridden through `...`.
Use `psd_tol` and `psd_max_iter` for the public tolerance and iteration
limit.

## Shrinkage toward the diagonal

With `psd_method = "shrink"`, let $C$ be the source corner and
$D$ the diagonal matrix containing its diagonal entries. The method
uses

$$
C_\alpha=\alpha C+(1-\alpha)D.
$$

Without a supplied `shrink_alpha`, a bisection search seeks the
largest retained weight satisfying the PSD tolerance. It stops when
the bracket width is at most `psd_tol` and returns the feasible end
of that bracket. Failure to narrow the bracket within `psd_max_iter`
raises `sitemix_error_vcov_projection_nonconvergence`.

A fixed `shrink_alpha` in $(0,1]$ is applied exactly for `K > 2`,
including to a corner that was already PSD. If that choice is not
PSD within tolerance, the call errors. A supplied weight requires
`psd_method = "shrink"` even for `K <= 2`, where the matrices
remain unchanged. Shrinkage controls how much of the chosen corner
to retain; it does not estimate the missing joint probabilities.

## Reading the diagnostics

`projection_diagnostics` has one row per site-year and scenario.
Start with `projection_status`, `projection_attempted`,
`converged`, and `iterations`. Status `"identity_k_le_2"` or
`"already_psd"` means no algorithm ran, so `converged = NA` is
expected. A fixed weight is recorded as `"fixed_alpha_applied"`;
a completed iterative adjustment is `"projected"`.

Compare `min_eigen_before` and `min_eigen_after` with their
`absolute_tolerance_before` and `absolute_tolerance_after` values.
Then inspect `projection_distance_absolute`, `diagonal_preserved`,
`symmetry_preserved`, and `psd_preserved`. The remaining diagnostics
include `sign_changes`, `raw_interval_violations`, and
`projected_order_reversals`, which check how the adjusted entries
relate to the original pairwise ranges. A global adjustment can
change an individual entry's interpretation; PSD alone is not a
reason to skip these checks. The deprecated `psd_diagnostics`
field provides an older summary layout.

# 7. A two-indicator check

The bundled simulated counts provide FRPM and SNAP marginals for
the same students. The following code computes a formal D1a result
and checks the `/n` covariance formula and equality of source and
returned matrices. Because this example has two indicators, it
checks the identity case; it does not exercise an iterative solver.

```{r invariants}
counts <- readRDS(system.file("extdata", "prek_sim_counts.rds",
                              package = "sitemix"))
d1 <- counts[counts$year == 2024, ]
d1_long <- rbind(
  data.frame(site_id = d1$site_id, year = d1$year,
             indicator = "frpm", c_jt = d1$c_jt_frpm, n_jt = d1$n_jt),
  data.frame(site_id = d1$site_id, year = d1$year,
             indicator = "snap", c_jt = d1$c_jt_snap, n_jt = d1$n_jt)
)
est_d1 <- capture_expected_sitemix_warning(
  sm_estimate_from_aggregates(
    d1_long, family = "multivariate",
    indicator_col = "indicator",
    sampling_relation = "same_units",
    vjt = TRUE
  ),
  "sitemix_warning_working_independence_default"
)
env <- sm_frechet_envelope(est_d1, population_regime = "d1a")

V_indep <- as.matrix(env$V_independence[[1L]])
raw_intervals <- env$raw_pairwise_intervals
V_negative <- as.matrix(env$projected_negative_dependence_stress[[1L]])
V_positive <- as.matrix(env$projected_positive_dependence_stress[[1L]])

# FR1: direct formal /n oracle
row1 <- raw_intervals[raw_intervals$site_key == env$site_keys$site_key[[1L]], ][1L, ]
stopifnot(all.equal(
  row1$pairwise_covariance_lower,
  (row1$joint_probability_lower - row1$p_1 * row1$p_2) / row1$n_common
))
stopifnot(all.equal(
  row1$pairwise_covariance_upper,
  (row1$joint_probability_upper - row1$p_1 * row1$p_2) / row1$n_common
))

# FR2: PSD after repair (relative tolerance + machine-range floor)
eig_negative <- eigen(V_negative, symmetric = TRUE, only.values = TRUE)$values
eig_positive <- eigen(V_positive, symmetric = TRUE, only.values = TRUE)$values
tol_negative <- env$psd_tol * max(abs(eig_negative)) +
  nrow(V_negative) * .Machine$double.xmin
tol_positive <- env$psd_tol * max(abs(eig_positive)) +
  nrow(V_positive) * .Machine$double.xmin
stopifnot(min(eig_negative) >= -tol_negative)
stopifnot(min(eig_positive) >= -tol_positive)

# FR3: K=2 shortcuts preserve their source corners
stopifnot(nrow(V_negative) == 2L)
stopifnot(all.equal(
  V_negative,
  as.matrix(env$unprojected_negative_dependence_corner[[1L]])
))
stopifnot(all.equal(
  V_positive,
  as.matrix(env$unprojected_positive_dependence_corner[[1L]])
))
stopifnot(all.equal(diag(V_negative), diag(V_indep), tolerance = 1e-8))
stopifnot(all.equal(diag(V_positive), diag(V_indep), tolerance = 1e-8))
```

The covariance endpoints in `raw_intervals` are on the raw
probability scale, while the input estimates use an arcsine scale.
Both returned stress matrices equal their source corners here, and
their diagonals equal `V_independence`. The eigenvalue checks use
the default relative tolerance `1e-8`; no iterative projection was
needed to obtain these matrices.

For the D1 working-independence calculation, see
[Sampling uncertainty from published aggregates](m5-aggregate-engines.html). For an applied comparison, see
[Variance smoothing and Fréchet sensitivity analysis](a7-variance-smoothing-and-frechet.html). [Understanding returned estimates and uncertainty](m8-output-contract.html)
explains how these sensitivity outputs differ from ordinary
covariance objects.

# References
