---
title: "A1 · Getting started — your first site-level estimate"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For applied researchers with site-year data who need site-level
  point estimates and sampling uncertainty for downstream analysis.
  This vignette walks through one `sm_estimate()`
  call on the bundled Alabama Pre-K sample and the column glossary
  every downstream consumer reads. You leave with a 50-site
  `sitemix_estimates` tibble, the diagnostic pass that confirms its
  uncertainty contract, and pointers to the next vignette.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A1 · Getting started — your first site-level estimate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  results    = "markup",
  comment    = "#>",
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  fig.width  = 7,
  fig.height = 4.5,
  fig.retina = 2,
  dpi        = 144,
  fig.align  = "center",
  out.width  = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

# Overview

## 1. Why you are here

Asha is the accountability lead at her district. She has one
season of student-row Pre-K enrollment data, and she has been
told to produce site-level FRPM (free and reduced-price meals)
rates with standard errors by Friday so her colleague can use them
in a downstream model. She knows what `sm_estimate()`
*should* do; she has not yet typed the first call.

If you opened this vignette, your situation is close enough.

## 2. What you will leave with

By the end of this five-minute walkthrough you will have:

-   A 50-site `sitemix_estimates` tibble for FRPM rates in 2024,
    one row per site-year-indicator.
-   A short tour of the column glossary every downstream consumer
    of `sitemix_estimates` reads.
-   The uncertainty audit (`sm_diagnose()`) showing that estimates,
    standard errors, scales, and any covariance payload are coherent.
-   Pointers to A2 (input formats) and M1 (statistical
    foundations) for when you want to go deeper.

**Prerequisites.** None — start here.

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 3. Install and load

If you have not installed sitemix yet:

```r
# install.packages("pak")
pak::pak("joonho112/sitemix")
```

Then load the package and the bundled sample data:

```{r library}
library(sitemix)
data(alprek_subset, package = "sitemix")
```

`alprek_subset` carries four overlapping binary indicators
(`frpm`, `snap`, `wic`, `tanf`) observed across years 2021–2025.

# 4. Your first estimate (the five-minute teaser)

The teaser: run `sm_estimate()` on one year of FRPM data and
print the result.

```{r teaser}
frpm_2024 <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)
print(
  as.data.frame(frpm_2024[
    1:5,
    c("site_id", "year", "indicator", "n", "theta_raw", "theta_hat", "se")
  ]),
  row.names = FALSE
)
```

Each row is one site in 2024. `n` is the site denominator (number
of students with non-missing FRPM status); `theta_raw` is the raw
proportion in `[0, 1]`; `theta_hat` is the same estimate on the
arcsine-stabilized scale; `se` is its delta-method standard error
on that scale. Down the page you will see the full column
glossary.

# 5. Read the columns

`frpm_2024` is a tibble of class `sitemix_estimates` carrying
the estimates, uncertainty, and provenance a downstream analysis
can consume. The full list:

```{r columns}
names(frpm_2024)
```

The columns you must know on first contact:

-   `site_id`, `year`, `indicator` — the keyed triple identifying
    each row.
-   `n` — site denominator.
-   `theta_raw` — raw proportion (always in `[0, 1]`).
-   `theta_hat`, `se` — point estimate and SE on the
    `estimate_scale` (default `"arcsine"`).
-   `var_method` — the row-level SE provenance; see
    [`sm_vcov()`](../reference/sm_vcov.html) for the implemented
    lexicon.
-   `flag_small_n`, `flag_below_accountability`, `flag_zero_cell`,
    `flag_suppressed` — the canonical row flags. Use these to
    audit or filter rows. A flag alone is not a smoothing trigger; any GVF
    alternative is a prespecified experimental sensitivity that must remain
    separate from canonical SEs.
-   `input_mode` — `"student_level"` here; would be
    `"counts_full_suff"` or `"aggregate"` under the wrappers.

The full glossary lives in
[`sm_estimate()`](../reference/sm_estimate.html)'s `@return`
section.

# 6. Confirm the uncertainty contract

Before passing the tibble to another analysis, run the diagnostics:

```{r diagnose}
diag <- sm_diagnose(frpm_2024, verbose = FALSE)
class(diag)
print(as.data.frame(diag), row.names = FALSE)
```

A `sitemix_diagnostics_summary` reports one row summarizing the
object, with denominator percentiles, tier counts, and intrinsic
uncertainty facts. For scalar output, check
`scalar_uncertainty_finite`, `scalar_se_positive`, and
`indicator_scale_consistent`. For output with a `V` list-column,
also inspect `v_valid` and `estimate_vcov_scale_compatible`.

For a row-level audit (every site, every flag), pass
`level = "row"`; for covariance-level (when `vjt = TRUE`), pass
`level = "vcov"`. The full walkthrough is in
[A6 · Diagnostics and suppression](a6-diagnostics-and-suppression.html).

# 7. Audit before going further

A bottom-of-file `stopifnot()` block makes the workflow assertions
explicit. If you fork this vignette into a methods appendix, keep
this chunk:

```{r audits}
stopifnot(nrow(frpm_2024) == 50L)  # 50 sites in 2024
stopifnot(all(frpm_2024$indicator == "frpm"))  # single indicator
stopifnot(all(frpm_2024$estimate_scale == "arcsine"))  # default scale
stopifnot(all(is.finite(frpm_2024$theta_hat)))  # no NaN/Inf
stopifnot(all(frpm_2024$se > 0))  # strictly positive SE
stopifnot("var_method" %in% names(frpm_2024))  # provenance column present
stopifnot(all(frpm_2024$var_method %in% c(
  "arcsine_vst", "arcsine_anscombe", "logit_delta",
  "binomial", "binomial_bc", "wilson_floor", "agresti_coull",
  "suppressed_drop", "suppression_sensitivity"
)))  # implemented base lexicon
```

# 8. What's next?

You now have a `sitemix_estimates` tibble for one indicator in
one year. The next steps depend on what your data look like:

-   If your data are **not student rows** (you have sufficient
    counts or a publisher CSV), read
    [A2 · Input formats](a2-input-formats.html) and pick the
    matching wrapper.
-   If you want the **boundary-method / VST knobs**, read
    [A3 · Scenario A — binomial](a3-scenario-binomial.html).
-   If you have **multiple indicators** (e.g., FRPM + SNAP +
    WIC), read
    [A4 · Scenarios B / C — multivariate / multinomial](a4-multivariate-multinomial.html).
-   When you are ready to use the estimates in another analysis,
    read [A8 · Downstream workflows](a8-downstream-workflows.html).
-   For the **formal foundations** (the sampling model and
    the delta-method SE derivation), read
    [M1 · Statistical foundations](m1-statistical-foundations.html)
    and [M2 · Scalar SE — binomial](m2-scalar-se-binomial.html).

# References
# Session info

```{r session-info, echo = FALSE, results = "hide"}
# Print session info in a real render
```
