---
title: "A2 · Input formats — student rows, counts, or aggregates?"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For applied researchers who have site-year data in some form but
  are not sure which sitemix entry function applies. This vignette
  walks through a three-branch decision tree (student rows /
  sufficient counts / published aggregates), gives one runnable
  example per branch, and shows the equivalence between branches.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A2 · Input formats — student rows, counts, or aggregates?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

# Overview

## 1. Why you are here

Marco is a state-agency analyst. Three different districts handed
him three different files: District A sent one row per student
(student rows); District B sent pre-aggregated count rows
(sufficient counts); District C sent the publisher's CSV (a
published aggregate). Marco needs to estimate site-level rates
from all three without spending a day deciding which sitemix
function to call.

This vignette is the decision tree.

**What you will leave with.** By the end of this article you will be able to:

1.  Decide whether your site-year data are student rows, sufficient
    counts, or a published aggregate.
2.  Call the matching entry function (`sm_estimate()`,
    `sm_estimate_from_counts()`, or `sm_estimate_from_aggregates()`).
3.  Rely on the count-versus-rows equivalence guarantee when your
    source data are already aggregated.

**Prerequisites.** [A1 · Getting started](a1-getting-started.html).

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`),
> and `alprek_subset_counts` for the sufficient-count path. It is not a real
> accountability dataset and must not be cited as empirical Pre-K results.
> Every code block runs offline with a fixed random seed.

# 2. The three-branch decision tree

Use the table below to pick your entry point.

| Your data look like                          | Use this function                                            | Scenario |
|:---------------------------------------------|:-------------------------------------------------------------|:--------:|
| One row per **student-year**                 | [`sm_estimate()`](../reference/sm_estimate.html)             | A/B/C    |
| One row per **site-year** with complete family-specific sufficient counts | [`sm_estimate_from_counts()`](../reference/sm_estimate_from_counts.html) | A/B/C |
| Publisher CSV with one row per **site-year** (or **site-year-subgroup**), with `c_jt` and `n_jt` columns or analogous | [`sm_estimate_from_aggregates()`](../reference/sm_estimate_from_aggregates.html) | D0/D1 |

Three questions to disambiguate:

1.  **"Does every row identify one student?"** Yes → branch 1
    (student rows). No → branch 2 or 3.
2.  **"Does every row carry an explicit numerator and
    denominator?"** Yes → branch 2 (counts) or 3 (aggregates).
3.  **"Is the data product a publisher CSV with a known schema
    (e.g., state accountability files)?"** Yes → branch 3.
    Otherwise → branch 2.

# 3. Branch 1 — student rows

This is the modal entry point and the one A1 demonstrated. One
row per student-year; the indicator column is logical or 0/1.

```{r student-rows}
data(alprek_subset, package = "sitemix")
est_rows <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)
head(est_rows, 3)
```

Use when your data are an enrollment file, a student-level
administrative export, or any table where each row is one
student in one year.

# 4. Branch 2 — sufficient counts

When student rows have already been aggregated to one row per
site-year with complete family-specific sufficient statistics, use
[`sm_estimate_from_counts()`](../reference/sm_estimate_from_counts.html).
The package ships a count artifact derived deterministically
from `alprek_subset`:

-   Scenario A needs `n_jt` and one `c_jt_<indicator>` column.
-   Scenario B needs `n_jt`, all ordered marginal counts, and all ordered
    pairwise co-occurrence counts. Count-input feasibility is verified for
    two or three indicators and fails closed at four or more.
-   Scenario C needs `n_jt` and at least two `c_jt_<category>` columns whose
    row sum is exactly `n_jt`.

```{r counts}
counts_path <- system.file(
  "extdata", "alprek_subset_counts.rds",
  package = "sitemix", mustWork = TRUE
)
counts <- readRDS(counts_path)
head(counts, 3)

snap_counts <- counts[
  counts$year == 2024,
  c("site_id", "year", "n_jt", "c_jt_snap")
]
est_counts <- sm_estimate_from_counts(
  snap_counts,
  family    = "binomial",
  indicator = "snap"
)
head(est_counts, 3)
```

# 5. Branch 3 — published aggregates

When the file is a publisher product (you did not aggregate it
yourself), use
[`sm_estimate_from_aggregates()`](../reference/sm_estimate_from_aggregates.html).
The wrapper supports two cases: **D0** (one numerator/denominator
per site-year) and **D1** (multiple aggregate marginals per
site-year).

D1 is a multivariate marginal model, not aggregate multinomial composition.
`family = "multinomial"` is therefore rejected on this branch; a Scenario C
composition must use student rows or complete sufficient category counts.

```{r aggregates-d0}
# Build a D0 aggregate slice from the bundled count file:
d0 <- counts[counts$year == 2024, c("site_id", "year", "n_jt", "c_jt_frpm")]
d0$indicator <- "frpm"
d0$c_jt <- d0$c_jt_frpm
d0 <- d0[c("site_id", "year", "indicator", "c_jt", "n_jt")]

est_agg <- sm_estimate_from_aggregates(
  d0,
  family    = "binomial",
  indicator = "frpm"
)
head(est_agg, 3)
```

For D1 (multiple marginals) and the publisher-side suppression
audit, read [A5 · Published aggregates D0 / D1](a5-published-aggregates.html).

# 6. The equivalence guarantee

Branches 1 and 2 must agree on the same underlying data. The
sufficient-counts identity (T2.5 in
[M2 · Scalar SE — binomial](m2-scalar-se-binomial.html))
guarantees:

```{r equivalence}
snap_2024 <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family    = "binomial",
  indicator = "snap"
)
snap_counts <- counts[counts$year == 2024, c("site_id", "year", "n_jt", "c_jt_snap")]
snap_from_counts <- sm_estimate_from_counts(
  snap_counts,
  family    = "binomial",
  indicator = "snap"
)
stopifnot(identical(snap_from_counts$site_id, snap_2024$site_id))
stopifnot(all.equal(snap_from_counts$theta_raw, snap_2024$theta_raw))
stopifnot(all.equal(snap_from_counts$theta_hat, snap_2024$theta_hat))
stopifnot(all.equal(snap_from_counts$se, snap_2024$se))
```

The two pathways agree to numerical tolerance. If you have a
choice between student rows and counts, prefer student rows for
clarity; the counts path exists for users whose source data are
already aggregated.

# 7. Audit

```{r audit}
stopifnot(nrow(est_rows) == 50L)
stopifnot(nrow(est_counts) == 50L)
stopifnot(nrow(est_agg) == 50L)
stopifnot(identical(unique(est_rows$estimate_scale), "arcsine"))
stopifnot(identical(unique(est_agg$estimate_scale), "arcsine"))
```

# 8. What's next?

-   [A3 · Scenario A — binomial](a3-scenario-binomial.html) — the
    Scenario A deep dive (boundary methods, VST choice, the
    sufficient-counts identity from §6 made formal).
-   [A4 · Scenarios B / C — multivariate / multinomial](a4-multivariate-multinomial.html)
    when your data have multiple indicators per site-year.
-   [A5 · Published aggregates D0 / D1](a5-published-aggregates.html)
    for branch-3 deep dive including subgroup pivots.
-   [M1 · Statistical foundations](m1-statistical-foundations.html)
    for the sampling-uncertainty framework underlying every input path.

# References
