---
title: "A3 · Scenario A — binomial estimates"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For applied researchers running Scenario A (one binary
  indicator per site-year, the modal sitemix call) who want to
  understand the boundary-method, reporting-transform,
  and accountability-threshold knobs. This vignette walks through
  the knobs with worked examples on `alprek_subset`.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A3 · Scenario A — binomial estimates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(alprek_subset, package = "sitemix")
```

# Overview

## 1. Why you are here

Asha (back from A1) noticed that some of her sites are tiny: only
one or two students were enrolled in 2024, and the resulting raw
proportions are 0/1 or 1/1. Her colleague asked: "Is the standard
error meaningful at a zero-cell row? And why is the default
arcsine instead of the raw proportion?" She came here for the
knobs that answer these questions.

## 2. What you will leave with

By the end you will have:

-   Comfort with the three boundary methods
    (`"wilson_floor"`, `"agresti_coull"`, `"none"`) and when to
    pick which.
-   A clear picture of the three reporting transforms
    (`"arcsine"`, `"logit"`, `"none"`) — only `"arcsine"` is
    variance-stabilizing.
-   The role of `accountability_n` and how the
    `flag_below_accountability` flag supports reporting decisions.
-   A side-by-side count vs student-row demonstration that
    confirms the T2.5 identity numerically.

**Prerequisites.** [A1 · Getting started](a1-getting-started.html).

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 3. The boundary methods

Boundary cells are rows where `theta_raw` is exactly 0 or 1 (the
numerator equals 0 or the denominator). Boundary policy matters
most on raw-scale output; under the default arcsine VST, the
row-level `se` remains the closed-form arcsine SE while `se_raw`
records the boundary-safe raw fallback. The example below uses
`vst = "none"` so the `var_method` labels are visible.

```{r boundary-wilson}
wf <- sm_estimate(
  subset(alprek_subset, year == 2021),
  family          = "binomial",
  indicator       = "tanf",
  vst             = "none",
  boundary_method = "wilson_floor"
)

ac <- sm_estimate(
  subset(alprek_subset, year == 2021),
  family          = "binomial",
  indicator       = "tanf",
  vst             = "none",
  boundary_method = "agresti_coull"
)
```

Inspect a few sites and look at the `var_method` column:

```{r boundary-compare}
side_by_side <- data.frame(
  site_id      = wf$site_id,
  n            = wf$n,
  theta_raw    = wf$theta_raw,
  wilson_se    = wf$se,
  wilson_meth  = wf$var_method,
  agresti_se   = ac$se,
  agresti_meth = ac$var_method
)
print(as.data.frame(head(side_by_side, 6)), row.names = FALSE)
```

When raw-scale `theta_raw` is `0` or `1`, the Wilson method records
`"wilson_boundary_surrogate"` and the Agresti-Coull method records
`"agresti_coull_boundary_surrogate"`. Interior rows record
`"binomial"` under raw output.

**When to pick which.** Use `"wilson_floor"` (the default,
[@wilson_1927]) for accountability-style reports: it floors
the SE at a small positive number without shifting `theta_raw`.
Use `"agresti_coull"` [@agresti_coull_1998] for the standard
z-general adjusted-Wald boundary uncertainty surrogate. Both methods
retain the observed `theta_raw = C/n`; neither replaces the point
estimate. Use `"none"` only for diagnostic comparisons (zero-cell
SEs collapse).

# 4. The reporting transforms

`vst` selects the reporting scale of `theta_hat` and `se` via three
choices:

```{r vst-grid}
arc <- sm_estimate(subset(alprek_subset, year == 2024),
                   family = "binomial", indicator = "frpm",
                   vst = "arcsine")
lgt <- sm_estimate(subset(alprek_subset, year == 2024),
                   family = "binomial", indicator = "frpm",
                   vst = "logit")
raw <- sm_estimate(subset(alprek_subset, year == 2024),
                   family = "binomial", indicator = "frpm",
                   vst = "none")
unique(arc$estimate_scale)
unique(lgt$estimate_scale)
unique(raw$estimate_scale)
```

The `estimate_scale` column records the chosen scale. The
default `"arcsine"` is the only variance-stabilizing choice; `"logit"`
is a reporting transform useful when downstream models expect log-odds;
and `"none"` retains the raw proportion scale for inspection or direct
reporting (see [M2](m2-scalar-se-binomial.html)).

# 5. The accountability threshold

`accountability_n` is the row-level denominator below which a
row is flagged `flag_below_accountability = TRUE`. The default
is 30. Smaller sites are estimated but flagged so an analyst can
apply the project's publication or modeling policy explicitly.

```{r accountability}
strict <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family           = "binomial",
  indicator        = "frpm",
  accountability_n = 50L
)
table(strict$flag_below_accountability)
```

Setting `accountability_n = 50L` flags rows with fewer than 50
observations. Use a higher threshold when your project's rules
treat sites below 50 as not publishable.

# 6. Sufficient-counts identity (T2.5 preview)

Demonstrate that the student-row path and the counts path agree
on the same underlying data:

```{r equivalence}
counts_path <- system.file("extdata", "alprek_subset_counts.rds",
                           package = "sitemix", mustWork = TRUE)
counts <- readRDS(counts_path)

snap_rows <- sm_estimate(subset(alprek_subset, year == 2024),
                         family = "binomial", indicator = "snap")
snap_cnts <- sm_estimate_from_counts(
  counts[counts$year == 2024,
         c("site_id", "year", "n_jt", "c_jt_snap")],
  family = "binomial", indicator = "snap"
)
stopifnot(all.equal(snap_rows$theta_hat, snap_cnts$theta_hat, tolerance = 1e-10))
stopifnot(all.equal(snap_rows$se, snap_cnts$se, tolerance = 1e-10))
```

The two pathways agree to `1e-10`. This is the T2.5 invariant;
the formal derivation lives in
[M2 · Scalar SE — binomial](m2-scalar-se-binomial.html).

# 7. Audit

```{r audit}
stopifnot(nrow(arc) == 50L)
stopifnot(all(arc$indicator == "frpm"))
stopifnot(all(is.finite(arc$theta_hat)))
stopifnot(all(arc$se > 0))
stopifnot(all(wf$var_method[wf$theta_raw %in% c(0, 1)] ==
                    "wilson_boundary_surrogate"))
stopifnot(all(ac$var_method[ac$theta_raw %in% c(0, 1)] ==
                    "agresti_coull_boundary_surrogate"))
```

# 8. What's next?

-   [A4 · Scenarios B / C — multivariate / multinomial](a4-multivariate-multinomial.html)
    if your project has multiple indicators per site-year.
-   [A6 · Diagnostics and suppression](a6-diagnostics-and-suppression.html)
    for the per-row audit and the accountability flag tour.
-   [M2 · Scalar SE — binomial](m2-scalar-se-binomial.html) for
    the formal delta-method derivation of every SE in this
    vignette.

# References
