---
title: "A4 · Scenarios B / C — multivariate and multinomial"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For applied researchers whose data have more than one outcome
  per site-year and need joint covariance. This vignette
  disambiguates Scenarios B (overlapping binary indicators with
  SUR-style covariance) and C (mutually exclusive categories with
  simplex covariance) and shows how to inspect the `V` list-column.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A4 · Scenarios B / C — multivariate and multinomial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(alprek_subset, package = "sitemix")
```

# Overview

## 1. Why you are here

Priya runs an accountability framework with both `FRPM` status
(binary) and a language-of-instruction category (4 mutually-
exclusive levels). Her downstream analysis compares indicators
jointly, so she needs not just SEs but **covariance** — the
`V` list-column. She also needs to decide between Scenario B
(overlapping binaries) and Scenario C (mutually exclusive
categories) before her first call.

**What you will leave with.** By the end of this article you will be able to:

1.  Distinguish overlapping-indicator (Scenario B) covariance from
    mutually-exclusive (Scenario C) covariance before your first call.
2.  Produce a `sitemix_estimates` tibble that carries a joint `V`
    list-column via `vjt = TRUE`.
3.  Read the `vcov_scale` and `vcov_method` metadata before consuming
    a covariance matrix.

**Prerequisites.** [A2 · Input formats](a2-input-formats.html).

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`),
> and a synthetic `primary_language` category for the Scenario C example. It
> is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 2. Scenario B vs Scenario C — the disambiguation

Two questions in order:

1.  **"Can a single student be in two indicators
    simultaneously?"** Yes → Scenario B (overlapping binaries).
    No → go to question 2.
2.  **"Do my indicators partition the population (each student
    belongs to exactly one)?"** Yes → Scenario C (multinomial).

Concrete examples:

-   FRPM / SNAP / WIC / TANF: a student can be in any subset.
    Scenario B.
-   Primary language ∈ {English, Spanish, Other}: a student is
    in exactly one. Scenario C.

The two scenarios share the structure "multiple components per
site-year with non-diagonal covariance," but they differ in
**which covariance**: Scenario B uses SUR-style cross-covariance
from joint proportions; Scenario C uses the multinomial simplex
covariance `(diag(π) − π π') / n`.

Both scenarios accept student rows through `sm_estimate()` and complete
sufficient counts through `sm_estimate_from_counts()`. Scenario B count input
must include every pairwise co-occurrence count; Scenario C count input must
include every category count and those counts must sum to `n_jt`. Published
aggregate marginals are a separate D1 multivariate path and are never routed
to Scenario C.

# 3. Scenario B — overlapping binary indicators

Call `sm_estimate()` with `family = "multivariate"` and an
`indicators` vector of column names. Set `vjt = TRUE` to attach
the per-row `V` list-column.

```{r scenario-b}
benefits_2024 <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family     = "multivariate",
  indicators = c("frpm", "snap", "wic", "tanf"),
  vjt        = TRUE
)
head(benefits_2024, 4)
```

Each site-year emits 4 rows (one per indicator). The covariance
across the four indicators for site `S001` is in the first
element of the `V` list-column:

```{r scenario-b-vcov}
V_S001 <- benefits_2024$V[[1L]]
round(as.matrix(V_S001), 4)
V_S001$vcov_scale
V_S001$vcov_method
```

**The `vcov_scale` caveat.** Note that
`vcov_scale = "raw"`: the covariance matrix is computed on the
raw probability scale, even though `theta_hat` is on the arcsine
scale (the default). This is because the SUR construction is
numerically stable in raw space; the matrix is not currently
transformed to the arcsine-delta scale.

The practical implication: **do not assume
`sqrt(diag(V)) == se`**. The row-level `se` is on the arcsine
scale; the matrix is on the raw scale. Always read
`V[[i]]$vcov_scale` before consuming the matrix.

# 4. Scenario C — mutually exclusive categories

`alprek_subset` does not ship with a mutually-exclusive category,
so this section constructs a deterministic synthetic example on
the same site-year grid.

```{r scenario-c-data}
language_data <- transform(
  alprek_subset,
  primary_language = sample(
    c("english", "spanish", "other"),
    nrow(alprek_subset),
    replace = TRUE,
    prob    = c(0.70, 0.25, 0.05)
  )
)
language_data <- language_data[
  c("student_id", "site_id", "year", "primary_language")
]
head(language_data, 5)
```

Now call `sm_estimate()` with `family = "multinomial"`:

```{r scenario-c}
language_2024 <- sm_estimate(
  subset(language_data, year == 2024),
  family    = "multinomial",
  indicator = "primary_language",
  vjt       = TRUE
)
head(language_2024, 6)
```

Three rows per site-year (one per category). The covariance for
site `S001`:

```{r scenario-c-vcov}
V_lang_S001 <- as.matrix(language_2024$V[[1L]])
round(V_lang_S001, 4)
```

The row sums of the simplex covariance are zero (the simplex
constraint):

```{r simplex-check}
rowSums(V_lang_S001)
```

# 5. Preparing scalar and joint uncertainty inputs

Some downstream analyses want one estimate and standard error per
indicator rather than a joint matrix. Select the canonical columns
and split them directly:

```{r split}
scalar_inputs <- split(
  as.data.frame(benefits_2024[, c(
    "site_id", "year", "indicator", "theta_hat", "se",
    "estimate_scale", "var_method"
  )]),
  benefits_2024$indicator
)
names(scalar_inputs)
```

This preserves the estimate scale and variance provenance instead of
silently translating them. Analyses that use cross-indicator
uncertainty should consume the `V` list-column and its metadata
directly; see [A8 · Downstream workflows](a8-downstream-workflows.html).

# 6. Audit

```{r audit}
stopifnot(nrow(benefits_2024) == 200L)  # 50 sites × 4 indicators
stopifnot(inherits(benefits_2024$V[[1]], "sm_vcov"))
stopifnot(benefits_2024$V[[1]]$vcov_method == "sur")
stopifnot(benefits_2024$V[[1]]$vcov_scale == "raw")
stopifnot(min(eigen(as.matrix(benefits_2024$V[[1]]),
                    symmetric = TRUE)$values) >= -1e-10)
stopifnot(nrow(language_2024) == 150L)  # 50 sites × 3 categories
stopifnot(language_2024$V[[1]]$vcov_method == "multinomial")
stopifnot(max(abs(rowSums(as.matrix(language_2024$V[[1]])))) < 1e-10)
```

# 7. What's next?

-   [A6 · Diagnostics and suppression](a6-diagnostics-and-suppression.html)
    for the `level = "vcov"` diagnostic that confirms PSD and
    reports the condition number.
-   [A8 · Downstream workflows](a8-downstream-workflows.html) — when the joint
    covariance is consumed by a downstream analysis.
-   [M3 · Multivariate SUR covariance](m3-multivariate-sur-covariance.html)
    for the formal SUR derivation.
-   [M4 · Multinomial simplex](m4-multinomial-simplex.html) for
    the formal simplex-covariance derivation.

# References
