---
title: "A5 · Published aggregates D0 / D1"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For external researchers working from publisher CSVs (state
  accountability files, district reports) rather than student
  rows. This vignette walks through the D0 (single
  numerator/denominator per site-year) and D1 (multiple
  marginals per site-year) paths, the Framing X / Framing Y
  subgroup pivots, and the three-tier publisher-side suppression
  audit.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A5 · Published aggregates D0 / D1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
```

# Overview

## 1. Why you are here

Dana is an external researcher with no microdata access. She
works exclusively from state-published accountability files:
numerator counts plus subgroup splits with cells masked at
`< 10`. She needs site-level rates and standard errors from
these files, plus raw pairwise intervals and projected stress scenarios for what the publisher
masked.

## 2. What you will leave with

By the end you will have:

-   A D0 (single-indicator) and a D1 (multi-marginal) call
    each producing a `sitemix_estimates` tibble.
-   The Framing X and Framing Y pivots for subgroup files.
-   A three-tier suppression audit confirming how many rows
    survive publisher-side masking.

**Prerequisites.** [A2 · Input formats](a2-input-formats.html).

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> The D0 and D1 rows are aggregate counts built from that sample (its
> companion `alprek_subset_counts` file), and the subgroup section adds a
> small synthetic fixture. It is not a real accountability dataset and
> must not be cited as empirical Pre-K results. Every code block runs
> offline with a fixed random seed.

# 3. Scenario D0 — one numerator and denominator per site-year

D0 is the simplest publisher case: each row is one site-year
with explicit numerator (`c_jt`) and denominator (`n_jt`).
Build a D0 slice from the bundled counts file:

```{r d0-data}
counts_path <- system.file(
  "extdata", "alprek_subset_counts.rds",
  package = "sitemix", mustWork = TRUE
)
counts <- readRDS(counts_path)

d0_frpm <- counts[counts$year == 2024,
                  c("site_id", "year", "n_jt", "c_jt_frpm")]
d0_frpm$indicator <- "frpm"
d0_frpm$c_jt <- d0_frpm$c_jt_frpm
d0_frpm <- d0_frpm[c("site_id", "year", "indicator", "c_jt", "n_jt")]
head(d0_frpm, 3)
```

Estimate:

```{r d0-est}
est_d0 <- sm_estimate_from_aggregates(
  d0_frpm,
  family    = "binomial",
  indicator = "frpm"
)
head(est_d0, 3)
unique(est_d0$estimate_scale)
```

The output is identical in shape to `sm_estimate()` Scenario A
output; the only difference is `input_mode = "aggregate"`.

# 4. Scenario D1 — multiple aggregate marginals per site-year

D1 is the case where the publisher reports several marginal
indicators per site-year (e.g., FRPM rate + SNAP rate + WIC
rate). Each marginal becomes a row; cross-marginal correlation
is unidentified from marginals alone, so the engine emits a
diagonal working-independence `V` (when `vjt = TRUE`).

Tell the engine whether those marginals describe the same observed units with
`sampling_relation`. Use `"same_units"` only with source documentation that
establishes a common sample, `"different_units"` when the marginal samples are
known to differ, and the default `"unknown"` otherwise. Equal denominators are
recorded as a denominator pattern; they do not prove common units.

```{r d1-build}
# Build a long-form D1 slice with two marginals (FRPM, SNAP):
d1 <- counts[counts$year == 2024, ]
d1_long <- rbind(
  data.frame(
    site_id   = d1$site_id, year = d1$year,
    indicator = "frpm",
    c_jt      = d1$c_jt_frpm, n_jt = d1$n_jt
  ),
  data.frame(
    site_id   = d1$site_id, year = d1$year,
    indicator = "snap",
    c_jt      = d1$c_jt_snap, n_jt = d1$n_jt
  )
)
head(d1_long, 4)
```

Estimate:

```{r d1-est}
est_d1 <- capture_expected_sitemix_warning(
  sm_estimate_from_aggregates(
    d1_long,
    family = "multivariate",
    indicator_col = "indicator",
    sampling_relation = "same_units",
    vjt = TRUE
  ),
  "sitemix_warning_working_independence_default"
)
head(est_d1, 4)
```

The working-independence `V` matrix for site `S001`:

```{r d1-vcov}
V_d1 <- as.matrix(est_d1$V[[1L]])
round(V_d1, 4)
est_d1$V[[1L]]$vcov_method
attr(est_d1, "sampling_relation")
attr(est_d1, "denominator_pattern")
attr(est_d1, "d1_regime")
```

`vcov_method = "working_independence"` — the off-diagonals are
exactly zero because cross-marginal covariance is not identified
from publisher marginals alone. For formal raw pairwise intervals and
explicitly non-bound projected stress scenarios, see
[A7 · Variance smoothing and Fréchet](a7-variance-smoothing-and-frechet.html)
and the formal derivation in
[M7 · Fréchet envelope theory](m7-frechet-envelope-theory.html).

# 5. Subgroup files — Framing X vs Framing Y

When the publisher file carries one row per `(site, year,
subgroup)` triple, you must pivot before estimation. The choice
between Framing X and Framing Y depends on your question.

Both helpers map the fixed publisher total vocabulary `ALL`,
`ALL STUDENT(S)`, `TOTAL`, and `OVERALL` to canonical `ALL`; matching ignores
case, surrounding whitespace, and punctuation between words. Framing X keeps
the publisher spelling in `source_subgroup` for audit while using canonical
`ALL` in its composite key, and Framing Y uses canonical `ALL` as the indicator.
Alias collisions fail as duplicates. Mixed school/district/state rows are not
routed automatically: split them into homogeneous reporting-level tables
before pivoting.

```{r subgroup-fixture}
# Synthetic subgroup file (5 sites × 2 subgroups × 1 year)
subgroups <- expand.grid(
  site_id  = paste0("S", sprintf("%03d", 1:5)),
  year     = 2024L,
  subgroup = c("ell", "non_ell"),
  stringsAsFactors = FALSE
)
subgroups$c_jt <- c(8, 4, 7, 5, 9, 3, 10, 6, 5, 8)
subgroups$n_jt <- c(12, 6, 11, 9, 13, 7, 15, 10, 8, 12)
head(subgroups, 4)
```

**Framing X** — each subgroup becomes its own site (composite
`site_id`):

```{r framing-x}
fx <- sm_pivot_subgroups_to_sites(
  subgroups,
  subgroup_col    = "subgroup",
  numerator_col   = "c_jt",
  denominator_col = "n_jt",
  indicator       = "frpm_take_up"
)
head(fx, 4)
```

The Framing-X default keeps `partition_target = "none"` and returns conditional
rates. The only composition options are `"denominator_composition"` and
`"case_composition"`; both require the same complete category grid and one
canonical `ALL` total in every site-year. A missing category row is an error,
not an implicit zero. Suppressed numerators rule out case composition, whereas
denominator composition is available only when every required denominator is
observed.

**Framing Y** — each subgroup becomes a marginal indicator of
the original site. Subgroups commonly describe different observational units,
so use `sampling_relation = "different_units"` unless the source establishes
otherwise:

```{r framing-y}
fy <- sm_pivot_subgroups_to_indicators(
  subgroups,
  subgroup_col    = "subgroup",
  numerator_col   = "c_jt",
  denominator_col = "n_jt"
)
head(fy, 4)

est_fy <- sm_estimate_from_aggregates(
  fy,
  family = "multivariate",
  indicators = attr(fy, "indicator_set"),
  framing = "subgroup_as_indicator",
  sampling_relation = "different_units"
)
```

Pass either pivoted tibble to
`sm_estimate_from_aggregates()` with the appropriate family.

# 6. Publisher-side suppression audit

Before estimation, count how many rows are Tier 1
(publisher-suppressed), Tier 2 (observed below
`accountability_n`), or Tier 3 (publishable). Use
[`sm_suppression_report()`](../reference/sm_suppression_report.html).

```{r suppression}
d0_with_subgroup <- d0_frpm
d0_with_subgroup$subgroup <- "all"
report <- sm_suppression_report(
  d0_with_subgroup,
  by              = c("subgroup", "year"),
  numerator_col   = "c_jt",
  denominator_col = "n_jt",
  indicator_col   = "indicator",
  subgroup_col    = "subgroup"
)
print(as.data.frame(report), row.names = FALSE)
```

If `pct_suppressed` is high, the default `suppression = "drop"`
retains unavailable audit rows with canonical estimates and SEs
missing. The legacy `suppression = "upper_bound"` option is not
imputation: with `suppression_sensitivity_acknowledge = TRUE`, it
stores a worst-case Bernoulli variance scenario only in separated
`sensitivity_*` fields. It never populates canonical estimates or
ordinary covariance. When the denominator is hidden, even those
numeric sensitivity variances remain missing.

# 7. Audit

```{r audit}
stopifnot(nrow(est_d0) == 50L)
stopifnot(identical(unique(est_d0$estimate_scale), "arcsine"))
stopifnot(nrow(est_d1) == 100L)  # 50 sites × 2 marginals
stopifnot(est_d1$V[[1]]$vcov_method == "working_independence")
stopifnot(all(diag(as.matrix(est_d1$V[[1]])) > 0))
stopifnot(max(abs(as.matrix(est_d1$V[[1]])[lower.tri(as.matrix(est_d1$V[[1]]))])) < 1e-12)
stopifnot(nrow(fx) == 10L)  # 5 sites × 2 subgroups
stopifnot(nrow(fy) == 10L)
```

# 8. What's next?

-   [A6 · Diagnostics and suppression](a6-diagnostics-and-suppression.html)
    for the full three-tier audit workflow.
-   [A7 · Variance smoothing and Fréchet](a7-variance-smoothing-and-frechet.html)
    for D1 raw pairwise Fréchet intervals and projected stress scenarios.
-   [M5 · Aggregate engines D0 / D1](m5-aggregate-engines.html)
    for the formal derivations.

# References
