---
title: "Estimating proportions from published aggregates"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  Estimate site-year proportions from published numerator and denominator
  counts. Work with one or several marginal indicators, reshape subgroup
  tables, and identify suppressed rows before estimation.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Estimating proportions from published aggregates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
```

Published tables often provide the numerator and denominator for each
site-year without individual student records. For one proportion, use the
single-indicator aggregate path (D0). For several marginal proportions, use
the multivariate aggregate path (D1).

Missing joint counts and suppressed marginal values are different problems.
Observed marginals can support individual rate estimates even when their
cross-indicator covariance is unknown. A suppressed numerator, in contrast,
cannot provide an observed proportion. The examples below keep these cases
separate; [Choosing an input format](a2-input-formats.html) compares the
aggregate path with student records and sufficient counts.

We use aggregate counts from the bundled, fully simulated `prek_sim` data
and a small generated subgroup table. These are examples of data handling,
not empirical findings about children or programs (see `?prek_sim`).

# One numerator and denominator per site-year (D0)

D0 is the simplest publisher case: each row is one site-year
with explicit numerator (`c_jt`) and denominator (`n_jt`).
Build a D0 slice from the bundled counts file:

```{r d0-data}
counts_path <- system.file(
  "extdata", "prek_sim_counts.rds",
  package = "sitemix", mustWork = TRUE
)
counts <- readRDS(counts_path)

d0_frpm <- counts[counts$year == 2024,
                  c("site_id", "year", "n_jt", "c_jt_frpm")]
d0_frpm$indicator <- "frpm"
d0_frpm$c_jt <- d0_frpm$c_jt_frpm
d0_frpm <- d0_frpm[c("site_id", "year", "indicator", "c_jt", "n_jt")]
head(d0_frpm, 3)
```

The default column names are already present, so no explicit column mapping
is needed. Estimate FRPM rates and their arcsine-scale standard errors:

```{r d0-est}
est_d0 <- sm_estimate_from_aggregates(
  d0_frpm,
  family    = "binomial",
  indicator = "frpm"
)
head(est_d0, 3)
unique(est_d0$estimate_scale)
```

Each row is one site's FRPM estimate for 2024. The main estimate and SE
columns have the same meaning as in student-level binomial output:
`theta_raw` is the observed proportion, while `theta_hat` and `se` use
`estimate_scale`. Here `input_mode = "aggregate"` records the source format.

For a publisher file with different column names, use `numerator_col` and
`denominator_col` to map them to `c_jt` and `n_jt`. An existing
`suppression_flag` is read by default; `suppression_col` can name a different
flag column. Do not turn a suppressed count into zero. See the suppression
example below for how unavailable values are retained.

# Several marginal proportions per site-year (D1)

D1 is the case where the publisher reports several marginal
indicators per site-year (e.g., FRPM rate + SNAP rate + WIC
rate). Each marginal becomes a row; cross-marginal correlation
is unidentified from marginals alone. With `vjt = TRUE`, sitemix uses a
diagonal `V` as a working-independence assumption; it does not estimate
zero dependence from the data.

Describe whether those marginals refer to the same observed units with
`sampling_relation`. Use `"same_units"` only with source documentation that
establishes a common sample, `"different_units"` when the marginal samples are
known to differ, and the default `"unknown"` otherwise. Equal denominators are
recorded as a denominator pattern; they do not prove common units.

```{r d1-build}
# Build a long-form D1 slice with two marginals (FRPM, SNAP):
d1 <- counts[counts$year == 2024, ]
d1_long <- rbind(
  data.frame(
    site_id   = d1$site_id, year = d1$year,
    indicator = "frpm",
    c_jt      = d1$c_jt_frpm, n_jt = d1$n_jt
  ),
  data.frame(
    site_id   = d1$site_id, year = d1$year,
    indicator = "snap",
    c_jt      = d1$c_jt_snap, n_jt = d1$n_jt
  )
)
head(d1_long, 4)
```

The FRPM and SNAP counts below were built from the same simulated students,
so `sampling_relation = "same_units"` is justified here. Requesting `V`
produces `sitemix_warning_working_independence_default`: the example captures
that expected warning, while any other warning stops the render.

```{r d1-est}
est_d1 <- capture_expected_sitemix_warning(
  sm_estimate_from_aggregates(
    d1_long,
    family = "multivariate",
    indicator_col = "indicator",
    sampling_relation = "same_units",
    vjt = TRUE
  ),
  "sitemix_warning_working_independence_default"
)
head(est_d1, 4)
```

Inspect the first site's working-independence matrix and recorded assumptions:

```{r d1-vcov}
V_d1 <- as.matrix(est_d1$V[[1L]])
round(V_d1, 4)
est_d1$V[[1L]]$vcov_method
attr(est_d1, "sampling_relation")
attr(est_d1, "denominator_pattern")
attr(est_d1, "d1_regime")
```

The zero off-diagonals express the working assumption. In this D1 example,
`V` is on the row's arcsine scale and its diagonal equals `se^2`. Keeping
`sampling_relation` separate from `denominator_pattern` makes the same-unit
assumption explicit; matching denominator values alone would not justify it.

To explore the unidentified dependence, see
[Variance smoothing and Fréchet sensitivity analysis](a7-variance-smoothing-and-frechet.html)
and [the Fréchet derivation](m7-frechet-envelope-theory.html). Formal pairwise
covariance ranges require complete same-unit marginals with a common finite
denominator, IID plug-in variance, `anscombe = FALSE`, and no FPC or
`binomial_bc`. Those ranges are raw-scale quantities. PSD-projected matrices
are stress scenarios and need not be attainable joint covariance bounds.

# Choose the unit of analysis for subgroup files

When the publisher file carries one row per `(site, year,
subgroup)` triple, you must pivot before estimation. The choice
between Framing X and Framing Y depends on your question.

Both helpers map the fixed publisher total vocabulary `ALL`,
`ALL STUDENT(S)`, `TOTAL`, and `OVERALL` to `ALL`; matching ignores
case, surrounding whitespace, and punctuation between words. Framing X keeps
the publisher spelling in `source_subgroup` for audit while using
`ALL` in its composite key, and Framing Y uses `ALL` as the indicator.
Alias collisions fail as duplicates. Mixed school/district/state rows are not
routed automatically: split them into homogeneous reporting-level tables
before pivoting.

```{r subgroup-fixture}
# Synthetic subgroup file (5 sites × 2 subgroups × 1 year)
subgroups <- expand.grid(
  site_id  = paste0("S", sprintf("%03d", 1:5)),
  year     = 2024L,
  subgroup = c("ell", "non_ell"),
  stringsAsFactors = FALSE
)
subgroups$c_jt <- c(8, 4, 7, 5, 9, 3, 10, 6, 5, 8)
subgroups$n_jt <- c(12, 6, 11, 9, 13, 7, 15, 10, 8, 12)
head(subgroups, 4)
```

**Framing X** — each subgroup becomes its own site (composite
`site_id`):

```{r framing-x}
fx <- sm_pivot_subgroups_to_sites(
  subgroups,
  subgroup_col    = "subgroup",
  numerator_col   = "c_jt",
  denominator_col = "n_jt",
  indicator       = "frpm_take_up"
)
head(fx, 4)
```

The default `partition_target = "none"` keeps each subgroup's own numerator
and denominator. It prepares conditional rates such as FRPM participation
within ELL students; assigning a new site ID does not turn them into shares
of the whole site's population.

For shares of a common site total, the same helper offers
`"denominator_composition"` and `"case_composition"`. Those options return
wide category counts for `sm_estimate_from_counts(family = "multinomial")`.
They require mutually exclusive categories covering the relevant total,
a complete category set, and one `ALL` row per site-year. The common
`n_jt` is the category-count sum, checked against `ALL` within the supplied
tolerance. A missing category must not be silently treated as zero.
Denominator composition needs observed denominators but permits suppressed
numerators; case composition needs observed numerators that are not suppressed.

**Framing Y** — each subgroup becomes a marginal indicator of
the original site. The ELL and non-ELL groups in this example represent
different students, so the call uses `sampling_relation = "different_units"`.
For other data, use that value only when the samples are known to differ;
keep `"unknown"` when the source does not establish their relationship:

```{r framing-y}
fy <- sm_pivot_subgroups_to_indicators(
  subgroups,
  subgroup_col    = "subgroup",
  numerator_col   = "c_jt",
  denominator_col = "n_jt"
)
head(fy, 4)

est_fy <- sm_estimate_from_aggregates(
  fy,
  family = "multivariate",
  indicators = attr(fy, "indicator_set"),
  framing = "subgroup_as_indicator",
  sampling_relation = "different_units"
)
```

In this example, both `fx` and `fy` retain the subgroup-specific fractions.
`fx` represents ten subgroup-sites; `fy` represents five original sites with
two indicators each. The pivots change that organization, not the counts.
The default indicator pivot removes an entire site-year if a selected
subgroup is missing or suppressed. Use `na_action = "keep_na"` to retain
missing coordinates, then choose the aggregate suppression handling explicitly.

# Count suppressed and small-denominator rows

Before estimation, count how many rows are Tier 1
(publisher-suppressed), Tier 2 (observed below
`accountability_n`), or Tier 3 (observed at or above that threshold).
These labels describe the supplied counts and threshold; they do not decide
whether a publisher may release a result. Use
[`sm_suppression_report()`](../reference/sm_suppression_report.html).

```{r suppression}
d0_with_subgroup <- d0_frpm
d0_with_subgroup$subgroup <- "all"
report <- sm_suppression_report(
  d0_with_subgroup,
  by              = c("subgroup", "year"),
  numerator_col   = "c_jt",
  denominator_col = "n_jt",
  indicator_col   = "indicator",
  subgroup_col    = "subgroup"
)
print(as.data.frame(report), row.names = FALSE)
```

This complete simulated slice has no suppressed rows. On a publisher file,
`pct_suppressed` reports their share on a 0–1 scale. The default
`suppression = "drop"` retains those rows with missing estimates and SEs.

If a separate variance-sensitivity analysis is warranted,
`suppression = "upper_bound"` with explicit acknowledgement stores
`sensitivity_*` fields. With an observed denominator it uses the Bernoulli
variance at p = 0.5; it does not fill in a suppressed proportion. This option
requires `vst = "arcsine"` and `anscombe = FALSE`. Hidden denominators leave
even the numeric sensitivity variances missing. Neither kind of suppressed
row can supply an ordinary `V` or Fréchet input.

# Check the example outputs

```{r audit}
stopifnot(nrow(est_d0) == 50L)
stopifnot(identical(unique(est_d0$estimate_scale), "arcsine"))
stopifnot(nrow(est_d1) == 100L)  # 50 sites × 2 marginals
stopifnot(est_d1$V[[1]]$vcov_method == "working_independence")
stopifnot(all(diag(as.matrix(est_d1$V[[1]])) > 0))
stopifnot(max(abs(as.matrix(est_d1$V[[1]])[lower.tri(as.matrix(est_d1$V[[1]]))])) < 1e-12)
stopifnot(nrow(fx) == 10L)  # 5 sites × 2 subgroups
stopifnot(nrow(fy) == 10L)
```

# Related examples

-   [Checking estimates and handling suppressed data](a6-diagnostics-and-suppression.html)
    for the full three-tier audit workflow.
-   [Variance smoothing and Fréchet sensitivity analysis](a7-variance-smoothing-and-frechet.html)
    for D1 raw pairwise Fréchet intervals and projected stress scenarios.
-   [Sampling uncertainty from published aggregates](m5-aggregate-engines.html)
    for the formal derivations.
