---
title: "Checking estimates and handling suppressed data"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  Check site estimates, standard errors, and covariance matrices with
  `sm_diagnose()`. Examine suppression and denominator thresholds in
  published input data with `sm_suppression_report()`.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Checking estimates and handling suppressed data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(prek_sim, package = "sitemix")
```

We check estimates of free and reduced-price meal participation (FRPM)
at each site in 2024. The `prek_sim` data describe 50 simulated
pre-kindergarten sites; the results illustrate calculations rather than
findings about real children or programs. For the first estimation call,
see [Getting started](a1-getting-started.html).

# Check the estimates as a group

First estimate each site's proportion and standard error:

```{r setup-est}
est <- sm_estimate(
  subset(prek_sim, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)
```

The default `sm_diagnose()` call returns one row describing the entire
result. Start with the number of estimates, their standard errors, and
the counts flagged by the denominator thresholds:

```{r diagnose-summary}
diag_s <- sm_diagnose(est, verbose = FALSE)
class(diag_s)
print(as.data.frame(diag_s)[, c(
  "n_cells", "n_flag_small_n", "n_flag_below_accountability",
  "scalar_uncertainty_finite", "scalar_se_positive",
  "indicator_scale_consistent", "diag_severity"
)], row.names = FALSE)
```

Here `n_cells` is 50: one estimate per site. `scalar_uncertainty_finite`
checks identified estimates and standard errors; intentionally missing
suppression results are counted separately. `indicator_scale_consistent`
checks that each indicator uses one estimate scale. Passing these checks
does not establish that a particular subsequent model is appropriate.

`scalar_se_positive` is `FALSE` for an exact census under simple random
sampling without replacement (SRSWOR), because its sampling standard
error is zero. Such rows receive a `zero_uncertainty_census` note. Keep
them distinguishable from unexplained zero standard errors, and do not
give zero-SE rows ordinary inverse-variance weights.

# Find rows that need attention

Use `level = "row"` to locate the estimates behind the summary counts.
It returns one row per site-year-indicator, retaining the estimates and
adding diagnostic columns:

```{r diagnose-row}
diag_r <- sm_diagnose(est, level = "row", verbose = FALSE)
class(diag_r)
head(as.data.frame(diag_r)[, c(
  "site_id", "n", "flag_small_n", "flag_below_accountability",
  "flag_zero_cell", "flag_suppressed", "diag_severity", "diag_notes"
)], 5)
```

The flags identify different reasons to inspect a row:

| Flag | Meaning |
|---|---|
| `flag_small_n` | The denominator is below the estimation call's `min_n` (default 10). |
| `flag_below_accountability` | The denominator is below `accountability_n` (default 30). |
| `flag_zero_cell` | The observed numerator is zero or equals the denominator. It is `NA` when suppression hides the numerator. |
| `flag_suppressed` | The published input row was identified as suppressed; individual-student input has no such rows. |

Read `diag_notes` alongside `diag_severity`. The severity uses the
following priority; the first applicable level is reported:

| Severity | Examples of the conditions recorded |
|---|---|
| `error` | Unexplained nonpositive SE, mixed scales within an indicator, inconsistent smoothing information, or a matrix no longer matching same-scale SEs. |
| `warning` | Suppressed-missing or variance-sensitivity rows, an estimate/covariance scale mismatch, or mixed smoothing-to-matrix scale relations. |
| `note` | Small denominators, boundary proportions, reporting thresholds, or an exact SRSWOR census. |
| `ok` | None of the preceding conditions applies. |

These values are returned data. A `diag_severity` value of `"warning"`
does not mean that the call emitted an R warning. Similarly,
`diag_warnings` and `diag_errors` contain diagnostic labels rather than
conditions raised by the call. `sm_diagnose()` validates the input first,
so an invalid object may stop the call before any table is returned.
`verbose = FALSE` suppresses the printed summary, not validation errors.

# Check covariance matrices and their scales

When estimates include `V` (`vjt = TRUE`), `level = "vcov"` returns one
row per site-year covariance matrix. For overlapping FRPM and SNAP
indicators, request the matrices during estimation:

```{r diagnose-vcov}
est_b <- sm_estimate(
  subset(prek_sim, year == 2024),
  family     = "multivariate",
  indicators = c("frpm", "snap"),
  vjt        = TRUE
)
diag_v <- sm_diagnose(est_b, level = "vcov", verbose = FALSE)
class(diag_v)
head(as.data.frame(diag_v)[, c(
  "site_id", "K", "min_eigenvalue", "psd_tol", "v_valid",
  "estimate_scale", "vcov_scale", "estimate_vcov_scale_compatible",
  "diag_severity"
)], 5)
```

Here `v_valid` is `TRUE`, but `estimate_vcov_scale_compatible` is
`FALSE`: the default estimates use the arcsine scale, while these
multivariate matrices stay on the raw probability scale. The diagnostic
records this as a warning. Use estimates and a covariance matrix on
compatible scales for a joint analysis; a valid matrix alone is not enough.

`min_eigenvalue` and `psd_tol` describe the positive-semidefinite check.
A matrix with `min_eigenvalue < -psd_tol` fails input validation before
diagnostics can summarize it. If smoothing has been used, also inspect
`smoothing_v_relation` and `v_stale` to distinguish scale differences from
a matrix that no longer matches the standard errors on its own scale.
See [Understanding the returned estimates](m8-output-contract.html) for
the scalar and matrix scale conventions.

# Examine suppression in published input

Before estimating published numerator/denominator pairs, use
`sm_suppression_report()` to count suppressed rows and compare observed
denominators with a reporting threshold. This example uses the companion
counts from the simulated data, grouped by subgroup and year:

```{r suppression}
counts_path <- system.file(
  "extdata", "prek_sim_counts.rds",
  package = "sitemix", mustWork = TRUE
)
counts <- readRDS(counts_path)
d0 <- counts[, c("site_id", "year", "n_jt", "c_jt_frpm")]
d0$indicator <- "frpm"
d0$c_jt <- d0$c_jt_frpm
d0$subgroup <- "all"
d0 <- d0[c("site_id", "year", "subgroup", "indicator", "c_jt", "n_jt")]

report <- sm_suppression_report(
  d0,
  by              = c("subgroup", "year"),
  numerator_col   = "c_jt",
  denominator_col = "n_jt",
  indicator_col   = "indicator",
  subgroup_col    = "subgroup"
)
print(as.data.frame(report)[, c(
  "subgroup", "year", "n_rows", "n_tier1", "n_tier2", "n_tier3",
  "n_denominator_missing"
)], row.names = FALSE)
```

The report's three tiers use suppression status and `accountability_n`
(30 by default):

| Report column | Rows counted |
|---|---|
| `n_tier1` | Publisher-suppressed rows. |
| `n_tier2` | Rows that are not suppressed and have an observed denominator below the threshold. |
| `n_tier3` | Rows that are not suppressed and have an observed denominator at or above the threshold. |

These are counts of input rows, not children. Meeting the threshold alone
does not establish that an estimate is suitable for publication or further
analysis. Missing denominators are also counted separately.

The report's `min_n` argument is stored as an attribute; changing it does
not change the tier counts. This differs from `sm_estimate(min_n = ...)`,
which sets `flag_small_n`. It also differs from
`sm_smooth_variance(scope = "tier2")`, which uses the fixed denominator
range 11–29 among otherwise eligible rows, irrespective of
`accountability_n`.

For published data, supply a suppression flag column or use an existing
`suppression_flag` column. A missing numerator with an observed positive
denominator also identifies suppression. A missing denominator alone
does not; a `suppression_when` predicate replaces the usual detection
rules. The report describes these inputs and does not fill in hidden values.

When Tier 1 rows are present, inspect `sensitivity_role`,
`sensitivity_numeric_variance_available`, and
`sensitivity_requires_acknowledgement` before estimation. The older
`upper_bound_*` columns remain available, but use the `sensitivity_*`
names when reading the report. In the estimator:

- `suppression = "drop"` retains the row with missing estimates and SEs.
- `suppression = "upper_bound"` requires `suppression_sensitivity_acknowledge = TRUE`,
  `vst = "arcsine"`, and `anscombe = FALSE` when suppressed rows are
  present. It leaves estimates and SEs missing and records a separate
  variance-sensitivity scenario in `sensitivity_*` fields. With a hidden
  denominator it provides no numeric sensitivity variance.

Suppressed rows cannot enter ordinary `V` matrices or Fréchet calculations;
use `vjt = FALSE` when retaining them. For worked estimation examples, see
[Published aggregates](a5-published-aggregates.html).

# Check the example results

```{r audit}
stopifnot(inherits(diag_s, "sitemix_diagnostics_summary"))
stopifnot(inherits(diag_r, "sitemix_diagnostics_row"))
stopifnot(inherits(diag_v, "sitemix_diagnostics_vcov"))
stopifnot(nrow(diag_s) == 1L)  # one row summarizing the object
stopifnot(nrow(diag_r) == 50L)  # one row per site-year-indicator
stopifnot(diag_s$scalar_uncertainty_finite == TRUE)
stopifnot(diag_s$scalar_se_positive == TRUE)
stopifnot(diag_s$indicator_scale_consistent == TRUE)
```

# Using the results

Keep reporting flags and suppression status with the estimates when
selecting rows for another analysis. [Downstream workflows](a8-downstream-workflows.html)
shows how to carry the estimates, SEs, and scales together.
[Variance smoothing and Fréchet](a7-variance-smoothing-and-frechet.html)
explores optional sensitivity analyses; a high `flag_small_n` count alone
is not a reason to replace the original standard errors.
For the sampling assumptions behind the diagnostics, see
[Statistical foundations](m1-statistical-foundations.html).

# References
