---
title: "A6 · Diagnostics and suppression"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For applied researchers who have a `sitemix_estimates` tibble
  and need to audit its uncertainty before downstream use. This vignette
  walks through all three `sm_diagnose()` levels (summary, row,
  vcov) and the three-tier `sm_suppression_report()` framework.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A6 · Diagnostics and suppression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(alprek_subset, package = "sitemix")
```

# Overview

## 1. Why you are here

Asha (from A1) and Dana (from A5) both arrive here with the
same question: "My tibble exists; how do I confirm its estimates
and uncertainty are coherent?" The answer is the same
two-function audit: `sm_diagnose()` (post-estimation row/vcov
audit) and `sm_suppression_report()` (publisher-side audit
applicable to D0 / D1 inputs).

## 2. What you will leave with

By the end you will have:

-   Comfort with all three `sm_diagnose()` levels.
-   A reading of the four canonical flag columns
    (`flag_small_n`, `flag_below_accountability`,
    `flag_zero_cell`, `flag_suppressed`).
-   A three-tier suppression report and a red-flags checklist
    you can paste into a methods appendix.

**Prerequisites.** [A1 · Getting started](a1-getting-started.html).

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> The diagnostics run on a `sitemix_estimates` tibble produced from that
> sample, and the suppression section uses its companion
> `alprek_subset_counts` count file. It is not a real accountability
> dataset and must not be cited as empirical Pre-K results. Every code
> block runs offline with a fixed random seed.

# 3. Set up an estimates tibble

```{r setup-est}
est <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)
```

# 4. `sm_diagnose(level = "summary")` — one row per object

The default `level = "summary"` returns one row summarizing the
object, with denominator percentiles, tier counts, and intrinsic
uncertainty facts.

```{r diagnose-summary}
diag_s <- sm_diagnose(est, verbose = FALSE)
class(diag_s)
print(as.data.frame(diag_s), row.names = FALSE)
```

Key columns:

-   `n_cells` — total rows.
-   `n_flag_small_n` — rows below `min_n`.
-   `n_flag_below_accountability` — rows below
    `accountability_n`.
-   `scalar_uncertainty_finite` — retained estimates and SEs are finite.
-   `scalar_se_positive` — retained scalar SEs are strictly positive.
    An exact SRSWOR census has zero uncertainty by design; it sets this fact
    to `FALSE` but is classified as the note `zero_uncertainty_census`.
-   `indicator_scale_consistent` — each indicator uses one estimate scale.
-   `v_valid` and `estimate_vcov_scale_compatible` — validity and
    scale compatibility when a `V` list-column is present.
-   `suppression_sensitivity_role` — either `none` or the explicit
    `nonidentified_variance_sensitivity` role.
-   `smoothing_v_relation` and `v_stale` — whether an experimental scalar
    smoothing result is absent, matching, incompatible, or mixed relative
    to `V`, and whether a matching-scale matrix was left stale.

The same severity priority is used at every level:

| Severity | Intrinsic meaning | Representative facts |
|---|---|---|
| `error` | Invalid or internally incoherent uncertainty | unexplained nonpositive SE, mixed scale within one indicator, invalid smoothing provenance, stale matching-scale `V` |
| `warning` | Valid object that needs an explicit analytical decision | suppressed-missing row, non-identified variance sensitivity, estimate/`V` scale mismatch, mixed smoothing-to-`V` relation |
| `note` | Descriptive fact, not invalidity | small denominator, zero cell, accountability threshold, exact SRSWOR census |
| `ok` | None of the preceding facts applies | coherent identified uncertainty without flagged context |

`sm_diagnose()` calls ordinary `sitemix_estimates` validation first. It
summarizes a validated object; it is not a substitute for validation.

# 5. `sm_diagnose(level = "row")` — one row per site-year-indicator

Use the row level when a reviewer asks "show me every flag for
every site":

```{r diagnose-row}
diag_r <- sm_diagnose(est, level = "row", verbose = FALSE)
class(diag_r)
head(diag_r, 5)
```

The row-level diagnostic includes the severity tier and every
flag column for every row.

# 6. `sm_diagnose(level = "vcov")` — covariance audit (Scenarios B/C/D1)

When the tibble was produced with `vjt = TRUE`, the `vcov`
level reports the minimum eigenvalue, its scale-aware PSD tolerance,
matrix validity, estimate/`V` scale compatibility, and smoothing/stale-`V`
facts per site-year:

```{r diagnose-vcov}
est_b <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family     = "multivariate",
  indicators = c("frpm", "snap"),
  vjt        = TRUE
)
diag_v <- sm_diagnose(est_b, level = "vcov", verbose = FALSE)
class(diag_v)
head(diag_v, 5)
```

If any row has `min_eigenvalue < -psd_tol`, the matrix is not PSD and
ordinary object validation fails before diagnostics can summarize it.
For a validated audit, require `v_valid == TRUE`; then interpret
`estimate_vcov_scale_compatible`, `smoothing_v_relation`, and `v_stale`
separately.

# 7. Suppression report (D0 / D1 only)

`sm_suppression_report()` audits the publisher's three-tier
denominator regime before estimation. Build a D0 slice with a
subgroup column:

```{r suppression}
counts_path <- system.file(
  "extdata", "alprek_subset_counts.rds",
  package = "sitemix", mustWork = TRUE
)
counts <- readRDS(counts_path)
d0 <- counts[, c("site_id", "year", "n_jt", "c_jt_frpm")]
d0$indicator <- "frpm"
d0$c_jt <- d0$c_jt_frpm
d0$subgroup <- "all"
d0 <- d0[c("site_id", "year", "subgroup", "indicator", "c_jt", "n_jt")]

report <- sm_suppression_report(
  d0,
  by              = c("subgroup", "year"),
  numerator_col   = "c_jt",
  denominator_col = "n_jt",
  indicator_col   = "indicator",
  subgroup_col    = "subgroup"
)
print(as.data.frame(report), row.names = FALSE)
```

Three-tier interpretation:

-   **Tier 1** — publisher-suppressed. `n_tier1` rows.
-   **Tier 2** — observed below `accountability_n`. Estimable
    but not publishable.
-   **Tier 3** — observed and meets threshold. The publishable
    population.

The canonical `sensitivity_role`,
`sensitivity_numeric_variance_available`, and
`sensitivity_requires_acknowledgement` fields make the post-audit decision
explicit. The `upper_bound_*` names remain legacy compatibility fields.
`suppression = "drop"` retains a canonical missing audit row. The legacy
`"upper_bound"` option is an acknowledged, non-identified variance
sensitivity stored only in `sensitivity_*` fields. It cannot be used to
construct ordinary `V` or a formal Fréchet input; a hidden denominator
yields no numeric variance claim.

# 8. Red-flags checklist (paste into methods appendix)

Use the following short checklist when reviewing a tibble
before publishing:

1.  `scalar_uncertainty_finite == TRUE` and
    `indicator_scale_consistent == TRUE`; require
    `scalar_se_positive == TRUE` except for a documented
    `zero_uncertainty_census` note.
2.  `n_flag_zero_cell == 0` (or boundary-method handling
    documented).
3.  `n_flag_below_accountability` reported and either small or
    explicitly accepted.
4.  For Scenarios B/C/D1: `v_valid == TRUE`,
    `min_eigenvalue >= -psd_tol`, and `v_stale == FALSE` on every matrix.
5.  Record any estimate/`V` scale mismatch as an intentional downstream
    scale decision; do not call it consumer readiness.
6.  For D0/D1: report Tier 1 alongside Tier 3 and label any sensitivity
    as non-identified.

# 9. Audit

```{r audit}
stopifnot(inherits(diag_s, "sitemix_diagnostics_summary"))
stopifnot(inherits(diag_r, "sitemix_diagnostics_row"))
stopifnot(inherits(diag_v, "sitemix_diagnostics_vcov"))
stopifnot(nrow(diag_s) == 1L)  # one row summarizing the object
stopifnot(nrow(diag_r) == 50L)  # one row per site-year-indicator
stopifnot(diag_s$scalar_uncertainty_finite == TRUE)
stopifnot(diag_s$scalar_se_positive == TRUE)
stopifnot(diag_s$indicator_scale_consistent == TRUE)
```

# 10. What's next?

-   [A7 · Variance smoothing and Fréchet](a7-variance-smoothing-and-frechet.html)
    to inspect an opt-in experimental GVF sensitivity alternative; a high
    `flag_small_n` count alone is not a reason to replace canonical SEs.
-   [A8 · Downstream workflows](a8-downstream-workflows.html) when the audit passes.
-   [M1 · Statistical foundations](m1-statistical-foundations.html)
    for the formal `sitemix_estimates` schema invariants the
    diagnostics check.

# References
