---
title: "A7 · Variance smoothing and Fréchet stress scenarios"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For applied researchers conducting an opt-in experimental GVF
  sensitivity analysis (`sm_smooth_variance()`) or evaluating D1
  aggregate dependence with formal pairwise intervals and projected
  stress scenarios (`sm_frechet_envelope()`). Canonical SEs remain
  primary; neither small sample size nor an unusual-looking SE is by
  itself a trigger to smooth.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A7 · Variance smoothing and Fréchet stress scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
data(alprek_subset, package = "sitemix")
```

# Overview

## 1. Why you are here

Tariq's tibble passed `sm_diagnose()` but a methods reviewer
pointed at the smallest sites and said "I don't believe these
SEs." Dana's D1 aggregate analysis (from A5) lacks publisher
correlation info; her reviewer wants a sensitivity analysis of
the unidentified cross-marginal covariance.

Both readers are in this vignette because it presents two sensitivity tools:
an experimental GVF alternative (available across scenarios) and Fréchet
pairwise intervals with projected stress scenarios (D1-specific). The first
half demonstrates the GVF alternative without recommending it; the second
half covers Fréchet analysis.

## 2. What you will leave with

By the end you will have:

-   A scale-specific experimental sensitivity column (`se_smoothed` or
    `se_raw_smoothed`) that leaves the canonical SE primary.
-   Formal raw pairwise intervals plus explicitly non-bound projected
    stress scenarios for the D1 sensitivity case.
-   Interpretation rules that keep non-identified or experimental alternatives
    separate from canonical sampling uncertainty.

**Prerequisites.** [A1 · Getting started](a1-getting-started.html).

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> The smoothing section runs on a `sitemix_estimates` tibble produced from
> that sample, and the Fréchet section uses its companion
> `alprek_subset_counts` count file. It is not a real accountability
> dataset and must not be cited as empirical Pre-K results. Every code
> block runs offline with a fixed random seed.

# 3. Variance smoothing (`sm_smooth_variance()`)

The following call demonstrates the experimental generalized
variance-function (GVF) / log-variance sensitivity alternative. It is not a
recommended response to small cells, and an unusual-looking SE is not evidence
that the fitted alternative is better:

```{r smoothing}
est <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)
est_s <- sm_smooth_variance(est, method = "loglinear")
head(est_s[, c("site_id", "n", "se", "se_smoothed")], 6)
```

`se_smoothed` is the cross-row trend-fitted SE on the same
scale as `se`. By default, `est$se` is preserved untouched;
pass `overwrite = TRUE` to replace it (with the original
preserved as `se_pre_smoothing`). The alternative's provenance
is stored in `var_method_smoothed`, while canonical `var_method`
also remains untouched under the default append-only path.
The appended column is a sensitivity candidate only; canonical `se` remains
the primary uncertainty input unless a separate, prespecified validation
supports a different choice.

This helper is not a Fay--Herriot area-level estimator: it does
not smooth point estimates or fit a between-area outcome model.
When a matching-scale `V` exists, overwrite is rejected because
retaining that matrix would leave a stale diagonal. Append-only
smoothing remains available, and an incompatible-scale `V` is
retained with its relationship recorded in `attr(x, "smoothing")`.

**How to interpret the optional sensitivity.** The smoother is a model on the
log-variance scale; it assumes the bulk of rows lies on a
common log-variance trend. If your data have a few sites that
genuinely have a different variance regime (e.g., a heavily
oversampled flagship site), the smoother will pull their SEs
toward the bulk. Inspect `se_smoothed - se` for outliers
before considering the alternative. A fixed-seed package audit over
small/large denominators and near-boundary/interior rates did not find a
candidate that met all predeclared variance-MSE, coverage, and inverse-weight
criteria. Consequently smoothing remains experimental, append-only, and
opt-in; the audit does not support replacing canonical SEs or treating the
smoothed alternative as a generally improved default.

# 4. GAM smoothing (optional)

For non-linear log-variance trends, the GAM backend uses
`mgcv::gam` under a runtime guard:

```{r gam}
if (requireNamespace("mgcv", quietly = TRUE)) {
  # Use a raw-output fixture: the default arcsine SE is an exact function of n
  # and can make a spline fit numerically degenerate on this example data.
  est_gam_input <- sm_estimate(
    subset(alprek_subset, year == 2024),
    family = "binomial", indicator = "frpm", vst = "none"
  )
  est_gam <- sm_smooth_variance(est_gam_input, method = "gam")
  head(est_gam[, c("site_id", "n", "se", "se_smoothed")], 6)
} else {
  message("mgcv is not installed; use method = \"loglinear\".")
}
```

`mgcv` is a Suggests-only dependency. If it is not installed,
fall back to `method = "loglinear"`. The raw-output fixture is deliberate:
unexpected fitting warnings fail the reproducibility audit instead of being
silently accepted as tutorial output.

# 5. Fréchet pairwise intervals and stress scenarios (D1 only)

For same-unit D1a inputs, Fréchet intervals formally constrain each raw
pairwise covariance that is unidentified from publisher marginals alone.
Build a
D1 input and call `sm_frechet_envelope()`:

```{r frechet}
counts_path <- system.file(
  "extdata", "alprek_subset_counts.rds",
  package = "sitemix", mustWork = TRUE
)
counts <- readRDS(counts_path)
d1 <- counts[counts$year == 2024, ]
d1_long <- rbind(
  data.frame(
    site_id = d1$site_id, year = d1$year,
    indicator = "frpm",
    c_jt = d1$c_jt_frpm, n_jt = d1$n_jt
  ),
  data.frame(
    site_id = d1$site_id, year = d1$year,
    indicator = "snap",
    c_jt = d1$c_jt_snap, n_jt = d1$n_jt
  )
)
est_d1 <- capture_expected_sitemix_warning(
  sm_estimate_from_aggregates(
    d1_long, family = "multivariate",
    indicator_col = "indicator",
    sampling_relation = "same_units",
    vjt = TRUE
  ),
  "sitemix_warning_working_independence_default"
)

env <- sm_frechet_envelope(est_d1, population_regime = "d1a")
class(env)
env$psd_method
```

The formal result is `raw_pairwise_intervals`, which records the marginal
probabilities, common denominator, joint-probability endpoints, correlation
endpoints, and raw covariance endpoints. The two PSD matrices are projected
stress scenarios sourced from the negative- and positive-dependence corners;
they are not multivariate lower or upper bounds. Inspect site `S001`:

```{r frechet-inspect}
head(env$raw_pairwise_intervals, 3)
V_negative_stress <- as.matrix(env$projected_negative_dependence_stress[[1L]])
V_positive_stress <- as.matrix(env$projected_positive_dependence_stress[[1L]])
round(V_negative_stress, 4)
round(V_positive_stress, 4)
summary(env)[, c("site_key", "scenario", "projection_status",
                 "sign_changes", "projected_order_reversals",
                 "raw_interval_violations",
                 "projection_distance_relative")]
```

Before projection, each off-diagonal is a raw pairwise interval endpoint.
For `K > 2`, PSD projection may change signs, reverse the elementwise order
of the two source corners, or leave a raw pairwise interval. The diagnostics
report those events directly. Treat the projected matrices only as stress
scenarios, never as bounds or as one identified sampling covariance matrix.

**D1a vs D1b.** Pass `population_regime = "d1a"` when the
marginals refer to the same sampled units, have one common finite denominator,
and use the IID plug-in rule without FPC or bias correction (formal raw
pairwise intervals). Pass `"d1b"` when the marginals are subgroup-
conditional (heuristic stress test). D1b requires explicit
acknowledgement via `subgroup_conditional_action = "allow"`.

# 6. Audit

```{r audit}
stopifnot("se_smoothed" %in% names(est_s))
stopifnot(all(est_s$se_smoothed > 0))
stopifnot(all(est_s$se == est$se))  # default doesn't overwrite
stopifnot(inherits(env, "sm_frechet_envelope"))
stopifnot(length(env$projected_negative_dependence_stress) == nrow(d1))
stopifnot(all(env$raw_pairwise_intervals$interval_scope ==
              "formal_raw_pairwise_interval"))
```

# 7. What's next?

-   [A6 · Diagnostics and suppression](a6-diagnostics-and-suppression.html)
    if you have not yet audited the tibble.
-   [A8 · Downstream workflows](a8-downstream-workflows.html) once smoothing /
    Fréchet are settled.
-   [M6 · Variance smoothing theory](m6-variance-smoothing-theory.html)
    and [M7 · Fréchet envelope theory](m7-frechet-envelope-theory.html)
    for the formal derivations and the D1a / D1b distinction.

# References
