---
title: "A8 · Downstream workflows"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For applied researchers who have a `sitemix_estimates` tibble
  and need to use its estimates and sampling uncertainty in another
  analysis. This vignette shows package-neutral scalar, split, and
  covariance workflows for reporting, meta-analysis, small-area estimation,
  and hierarchical models, built from `theta_hat`, `se`, `V`, and their
  provenance columns.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A8 · Downstream workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(alprek_subset, package = "sitemix")
```

# Overview

## 1. Why you are here

Marco has finished estimating site-level rates. One collaborator wants
a flat table with one estimate and standard error per row; another
wants the joint covariance across indicators. Both can use sitemix
outputs directly. No consumer-specific object is required.

**What you will leave with.** By the end of this article you will be able to:

1.  Export the package-neutral scalar, split, and joint-covariance
    contract from a `sitemix_estimates` tibble.
2.  Reconcile inverse-variance eligibility (`iv_eligible` and
    `exclusion_reason`) before filtering rows.
3.  Treat `sm_frechet_envelope()` output as a separate stress scenario,
    never as an ordinary `V`.

**Prerequisites.** [A1 · Getting started](a1-getting-started.html).

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> The workflows run on `sitemix_estimates` tibbles produced from that
> sample. It is not a real accountability dataset and must not be cited as
> empirical Pre-K results. Every code block runs offline with
> a fixed random seed.

# 2. The portable output contract

The core scalar fields are:

-   `site_id`, `year`, and `indicator` for row identity;
-   `theta_hat` and `se` for the estimate and standard error;
-   `estimate_scale` for their common scale; and
-   `var_method` for uncertainty provenance.

The raw-scale pair is `theta_raw` plus `se_raw`; do not mix one member of the
raw pair with one member of the transformed pair.

Portable exchange is an all-row audit table, not a pre-filtered model
matrix. Retain `flag_suppressed`, `flag_below_accountability`, and any
`estimate_status` and `sensitivity_*` columns that are present. A suppressed
missing row or a non-identified sensitivity row remains part of the audit even
though it is not ordinary scalar uncertainty. An exact SRSWOR census can have
`se = 0`; retain it as a separate state instead of attempting an infinite
inverse-variance weight.

When `V` is present, each `sm_vcov` matrix adds joint uncertainty.
Read its `vcov_scale`, `estimate_scale`, and `vcov_method` metadata
before using it.

# 3. Scalar output (Scenarios A / D0)

Estimate one indicator, diagnose it, and project every row to portable
columns. `as.data.frame()` is a transport operation, not validation, so the
diagnostic check comes first:

```{r scalar}
est_a <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)

diag_a <- sm_diagnose(est_a, verbose = FALSE)
object_metadata <- attributes(est_a)[c(
  "description", "family", "sitemix_role", "aggregate_case",
  "sampling_relation", "denominator_pattern", "d1_regime",
  "d1_regime_by_group", "suppression", "smoothing"
)]
audit_columns <- setdiff(names(est_a), c("V", "K"))
scalar_input <- as.data.frame(est_a[, audit_columns, drop = FALSE])

status <- if ("estimate_status" %in% names(scalar_input)) {
  scalar_input$estimate_status
} else {
  rep("identified", nrow(scalar_input))
}
exact_census <- rep(FALSE, nrow(scalar_input))
if ("sampling_fraction" %in% names(scalar_input)) {
  exact_census <-
    is.finite(scalar_input$sampling_fraction) &
    scalar_input$sampling_fraction == 1 & scalar_input$se == 0
}
scalar_input$iv_eligible <-
  status == "identified" & is.finite(scalar_input$theta_hat) &
  is.finite(scalar_input$se) & scalar_input$se > 0
scalar_input$exclusion_reason <- ifelse(
  scalar_input$iv_eligible, "eligible",
  ifelse(
    status != "identified", status,
    ifelse(
      exact_census, "exact_census_not_inverse_weighted",
      ifelse(
        !is.finite(scalar_input$theta_hat) | !is.finite(scalar_input$se),
        "nonfinite_scalar_uncertainty",
        "nonpositive_se_unexplained"
      )
    )
  )
)
head(scalar_input[, c(
  "site_id", "year", "indicator", "theta_hat", "se", "estimate_scale",
  "var_method", "flag_suppressed", "flag_below_accountability",
  "iv_eligible", "exclusion_reason"
)], 5)
```

Reconcile `iv_eligible` and `exclusion_reason` counts before filtering. Carry
`object_metadata` as an explicit sidecar when those attributes matter, because
a plain data frame does not preserve the `sitemix_estimates` contract. Rename
`theta_hat` and `se` only at the boundary where another package specifies
different argument names. Keep `estimate_scale`, `var_method`, and audit
columns alongside them so the translation remains reviewable. Never relabel
`sensitivity_var` as an ordinary SE or inverse-variance weight.

# 4. One scalar table per indicator (Scenarios B / C / D1)

For a multi-indicator result, remove `V` deliberately by projecting all rows
before filtering or splitting. Directly selecting a partial indicator set from
a `V`-bearing object is rejected because it would leave a partial covariance
group:

```{r split}
est_b <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family     = "multivariate",
  indicators = c("frpm", "snap"),
  vjt        = TRUE
)

audit_columns_b <- setdiff(names(est_b), c("V", "K"))
scalar_b <- as.data.frame(est_b[, audit_columns_b, drop = FALSE])
scalar_by_indicator <- split(scalar_b, scalar_b$indicator)
names(scalar_by_indicator)
```

This workflow deliberately omits cross-indicator covariance. It is
appropriate only when the downstream analyses are genuinely separate.

# 5. Joint covariance output

For a joint analysis, validate at summary, row, and covariance levels before
conversion. A consumed group must contain only identified coordinates, a
valid non-stale `V`, valid smoothing provenance, and one complete indicator
set. A suppression or sensitivity row cannot be removed while its matrix is
retained.

Make one covariance-index row per `(site_id, year)` tuple. Align rows with
`indicator_order`; do not assume the incoming data-frame order is the matrix
order. This Scenario-B example deliberately chooses a named contrast on the
raw scale because its reported arcsine estimates are not compatible with the
raw covariance:

```{r joint}
diag_b <- sm_diagnose(est_b, verbose = FALSE)
diag_b_row <- sm_diagnose(est_b, level = "row", verbose = FALSE)
diag_b_vcov <- sm_diagnose(est_b, level = "vcov", verbose = FALSE)

status_b <- if ("estimate_status" %in% names(diag_b_row)) {
  diag_b_row$estimate_status
} else {
  rep("identified", nrow(diag_b_row))
}
stopifnot(
  diag_b$v_present,
  diag_b$v_valid,
  all(status_b == "identified"),
  !any(diag_b_row$flag_suppressed),
  !identical(diag_b$smoothing_provenance_valid, FALSE),
  identical(diag_b$v_stale, FALSE),
  all(diag_b_vcov$v_valid),
  all(diag_b_vcov$psd_ok),
  all(diag_b_vcov$repeated_v_equal),
  !any(diag_b_vcov$row_sum_zero_ok %in% FALSE),
  !any(diag_b_vcov$diag_severity == "error")
)

reported_ready <- all(diag_b_vcov$estimate_vcov_scale_compatible)
stopifnot(!reported_ready)
analysis_scale <- "raw"       # explicit analytical choice
contrast <- c(frpm = 1, snap = -1)

estimate_table <- as.data.frame(est_b)
group_keys <- unique(estimate_table[c("site_id", "year")])
covariance_table <- group_keys
covariance_table$indicator_order <- I(vector("list", nrow(group_keys)))
covariance_table$V <- I(vector("list", nrow(group_keys)))
covariance_table$vcov_scale <- rep(NA_character_, nrow(group_keys))
covariance_table$vcov_method <- rep(NA_character_, nrow(group_keys))
covariance_table$diag_contract <- rep(NA_character_, nrow(group_keys))
covariance_table$point_scale <- rep(NA_character_, nrow(group_keys))
covariance_table$contrast_estimate <- rep(NA_real_, nrow(group_keys))
covariance_table$contrast_variance <- rep(NA_real_, nrow(group_keys))

for (g in seq_len(nrow(group_keys))) {
  take <- estimate_table$site_id == group_keys$site_id[g] &
    estimate_table$year == group_keys$year[g]
  block <- estimate_table[take, , drop = FALSE]
  V <- block$V[[1L]]
  stopifnot(all(vapply(block$V, identical, logical(1), V)))
  stopifnot(identical(V$site_id, group_keys$site_id[g]))
  stopifnot(identical(V$year, group_keys$year[g]))
  stopifnot(!anyDuplicated(block$indicator))
  stopifnot(setequal(block$indicator, V$indicator_order))
  coordinate <- match(V$indicator_order, block$indicator)
  stopifnot(!anyNA(coordinate), length(coordinate) == nrow(block))
  aligned <- block[coordinate, , drop = FALSE]
  matrix <- as.matrix(V)       # revalidates the sm_vcov object
  stopifnot(identical(aligned$indicator, V$indicator_order))
  stopifnot(identical(rownames(matrix), V$indicator_order))
  stopifnot(identical(colnames(matrix), V$indicator_order))
  stopifnot(all(block$K == length(V$indicator_order)))

  if (identical(analysis_scale, "raw")) {
    stopifnot(identical(V$vcov_scale, "raw"))
    stopifnot(V$diag_contract %in% c(
      "row_se_raw_squared",
      "row_se_raw_squared_except_boundary_surrogates"
    ))
    point <- aligned$theta_raw
    point_scale <- "raw_probability"
  } else {
    stopifnot(reported_ready)
    point <- aligned$theta_hat
    point_scale <- unique(aligned$estimate_scale)
  }
  weight <- unname(contrast[V$indicator_order])
  stopifnot(!anyNA(weight), all(is.finite(point)))
  contrast_variance <- drop(crossprod(weight, matrix %*% weight))
  stopifnot(is.finite(contrast_variance), contrast_variance >= -1e-14)

  covariance_table$indicator_order[[g]] <- V$indicator_order
  covariance_table$V[[g]] <- V
  covariance_table$vcov_scale[g] <- V$vcov_scale
  covariance_table$vcov_method[g] <- V$vcov_method
  covariance_table$diag_contract[g] <- V$diag_contract
  covariance_table$point_scale[g] <- point_scale
  covariance_table$contrast_estimate[g] <- sum(weight * point)
  covariance_table$contrast_variance[g] <- max(0, contrast_variance)
}

head(covariance_table[c(
  "site_id", "year", "point_scale", "vcov_scale", "vcov_method",
  "diag_contract", "contrast_estimate", "contrast_variance"
)], 5)
```

In this example `theta_hat` uses the arcsine scale while `V` uses the
raw scale, so `diag_b$estimate_vcov_scale_compatible` is `FALSE`.
The example does not ignore that warning: it explicitly uses `theta_raw` with
the raw `V` and records `point_scale = "raw_probability"`. If
`analysis_scale = "reported"`, compatibility must be `TRUE`; there is no
automatic transformation. A named contrast uses
$a^\top V a$ and does not require `solve(V)`, so valid singular multinomial
matrices and exact-census zero matrices remain consumable.

# 6. Fréchet projections are separate stress scenarios

`sm_frechet_envelope()` has a different role from the ordinary `V` above.
Its `raw_pairwise_intervals` are formal pairwise intervals only for approved
D1a input. The canonical projected fields
`projected_negative_dependence_stress` and
`projected_positive_dependence_stress` are separately labeled PSD stress
scenarios with `projected_scenario_role = "stress_scenario_not_bound"`.

Validate an envelope by calling `summary()` before extraction, carry its
population regime, scope, scenario, projection method/status, distance,
sign-change, raw-interval-violation, and order-reversal diagnostics, and rerun
the downstream calculation once per scenario. Do not wrap a projected matrix
as `sm_vcov`, insert it into ordinary `V`, average the two scenarios, or report
them as multivariate lower and upper bounds. Even when the `K = 2` projection
is an identity, the formal claim comes from `raw_pairwise_intervals`; D1b
remains heuristic.

# 7. Diagnostics before downstream use

Use `sm_diagnose()` as an intrinsic contract check:

| Diagnostic fact | Interpretation |
|:----------------|:---------------|
| `scalar_uncertainty_finite` | identified canonical estimates and SEs are finite |
| `scalar_se_positive` | identified non-census scalar SEs are strictly positive |
| `n_zero_uncertainty_census` | exact SRSWOR census rows are valid but are not inverse-weighted |
| `indicator_scale_consistent` | each indicator uses one estimate scale |
| `v_present` / `v_valid` | covariance payload is present and structurally valid |
| `estimate_vcov_scale_compatible` | row estimates and covariance use compatible scales |
| `n_suppressed_missing` / `n_suppression_sensitivity` | coordinates that cannot enter an ordinary joint group |
| `smoothing_provenance_valid` / `v_stale` | whether canonical `V` remains coherent after smoothing activity |

These are properties of the uncertainty output itself, not claims
about compatibility with any particular downstream package.

The canonical pair is `se` plus `var_method`. If an analysis deliberately
chooses the experimental append-only smoother, it must instead pair
`se_smoothed` (or `se_raw_smoothed` on the raw scale) with
`var_method_smoothed`, preserve the canonical columns for audit, and record
the smoothing diagnostic. No portable workflow substitutes a smoothed SE
automatically.

# 8. Audit

```{r audit}
stopifnot(inherits(est_a, "sitemix_estimates"))
stopifnot(inherits(est_b, "sitemix_estimates"))
stopifnot(diag_a$scalar_uncertainty_finite)
stopifnot(diag_a$scalar_se_positive)
stopifnot(diag_a$indicator_scale_consistent)
stopifnot(diag_b$v_present)
stopifnot(diag_b$v_valid)
stopifnot(!diag_b$estimate_vcov_scale_compatible)
stopifnot(identical(diag_b$v_stale, FALSE))
stopifnot(nrow(scalar_input) == nrow(est_a))
stopifnot(sum(table(scalar_input$exclusion_reason)) == nrow(est_a))
stopifnot(length(covariance_table$V) == 50L)
stopifnot(all(lengths(covariance_table$indicator_order) == 2L))
stopifnot(all(covariance_table$point_scale == "raw_probability"))
stopifnot(all(covariance_table$contrast_variance >= 0))
```

# 9. What's next?

-   [A9 · Case study — Alabama Pre-K](a9-case-study-alabama-prek.html)
    for the estimate → audit → canonical uncertainty export pipeline,
    with any experimental sensitivity kept separate and explicitly labeled.
-   [M8 · Output contract](m8-output-contract.html)
    for the formal scalar and covariance invariants.

# References
