---
title: "A complete analysis with simulated pre-kindergarten data"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  Follow a complete analysis of FRPM proportions in simulated
  pre-kindergarten data, from selecting a year to checking and exporting
  estimates. The example also shows how to describe the calculation and
  keep an optional smoothing comparison separate.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A complete analysis with simulated pre-kindergarten data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(prek_sim, package = "sitemix")
```

This example estimates the proportion of students receiving free and
reduced-price meals (FRPM) at each of 50 simulated sites in 2024. We prepare
a table of rates and sampling uncertainty and flag small denominators for
review. A reporting threshold is a chosen rule in the example, not a finding
about whether a real program may publish its data.

The bundled `prek_sim` panel contains no real children or programs. Every
result below is a calculation on simulated records, so it should not be
reported as an empirical Pre-K result. See [Getting started](a1-getting-started.html)
for the first estimation call on its own.

# Select the year

```{r prep}
target_year <- 2024L
dat <- subset(prek_sim, year == target_year)
nrow(dat)
length(unique(dat$site_id))
```

Restricting to 2024 leaves 50 sites in this sample.

# Estimate FRPM proportions

```{r estimate}
est <- sm_estimate(
  dat,
  family    = "binomial",
  indicator = "frpm"
)
head(est[, c("site_id", "n", "theta_raw", "theta_hat", "se",
             "flag_small_n", "flag_below_accountability")], 5)
```

The default `vst = "arcsine"` returns an arcsine-stabilized estimate
and its standard error; `estimate_scale` records that choice. `theta_raw`
retains the proportion for reporting. Both small-sample and accountability
flags are returned without removing the flagged rows.

# Review the estimates

```{r diagnose}
diag <- sm_diagnose(est, verbose = FALSE)
print(as.data.frame(diag), row.names = FALSE)
```

The summary reports `n_flag_small_n`, `n_flag_below_accountability`,
and checks such as `scalar_uncertainty_finite`,
`scalar_se_positive`, and `indicator_scale_consistent`.

# An optional variance-smoothing comparison

This analysis uses the original `se`. Small sample size alone is not a
reason to replace it with a smoothed value. The optional GVF helper is
experimental, and improved precision or model performance is not guaranteed.
If a separate, prespecified sensitivity analysis calls for smoothing, create
another object and retain the additional SE column:

```{r smoothing-sensitivity, eval = FALSE}
est_sensitivity <- sm_smooth_variance(
  est,
  method = "loglinear",
  overwrite = FALSE
)
# Keep est_sensitivity$se_smoothed alongside the original est$se.
```

The optional chunk is marked `eval = FALSE`; the reporting example below
uses the original estimates. See [the variance-smoothing methods](m6-variance-smoothing-theory.html)
for the model assumptions and the scope of the package's simulation checks.

# Export estimates and retain the review flags

```{r uncertainty-export}
analysis_input <- data.frame(
  unit_id        = est$site_id,
  year           = est$year,
  indicator      = est$indicator,
  estimate       = est$theta_hat,
  std_error      = est$se,
  estimate_scale = est$estimate_scale,
  var_method     = est$var_method,
  n              = est$n,
  proportion     = est$theta_raw,
  se_raw         = est$se_raw,
  flag_small_n   = est$flag_small_n,
  flag_below_accountability = est$flag_below_accountability,
  flag_zero_cell = est$flag_zero_cell,
  flag_suppressed = est$flag_suppressed
)
# If a prespecified sensitivity was run, add it under a separate name:
# analysis_input$std_error_gvf_sensitivity <- est_sensitivity$se_smoothed
head(analysis_input, 5)
```

The table retains all sites. For a percentage table, use `100 * proportion`
and, when needed, `100 * se_raw`. For an arcsine-scale analysis, use
`estimate` with `std_error`, keeping `estimate_scale` and `var_method`.
The flags and denominator remain available to explain later row selection.
[Using estimates in further analyses](a8-downstream-workflows.html) shows
additional handling for suppressed rows, exact censuses, and covariance
matrices; this example contains complete student records.

# Describe what was calculated

A methods description should identify the data, estimate scale, uncertainty
calculation, and any row-selection rules. For this simulated example:

> We calculated FRPM proportions for 50 simulated pre-kindergarten sites in
> 2024 using sitemix [@lee_2026_sitemix]. We retained raw proportions for
> percentage reporting and arcsine-transformed estimates with their
> delta-method standard errors for further analysis. We used the package's
> default boundary setting, which supplies a Wilson-based variance surrogate
> if a proportion is 0 or 1 [@wilson_1927]. Sites with n < 30 were flagged
> but retained in the exported table. The table contains the original
> unsmoothed standard errors, their scale and calculation method, and the
> denominator and review flags. These simulated results are not estimates
> for real children or programs.

# Check the exported table

```{r audit}
stopifnot(nrow(est) == 50L)
stopifnot(all(est$indicator == "frpm"))
stopifnot(diag$scalar_uncertainty_finite == TRUE)
stopifnot(diag$scalar_se_positive == TRUE)
stopifnot(diag$indicator_scale_consistent == TRUE)
stopifnot(all(est$estimate_scale == "arcsine"))
stopifnot(all(is.finite(est$theta_hat)))
stopifnot(nrow(analysis_input) == nrow(est))
stopifnot(identical(analysis_input$std_error, est$se))
stopifnot(!"se_smoothed" %in% names(est))
```

# Methods and related examples

-   [Sampling uncertainty in site-level proportions](m1-statistical-foundations.html)
    for the sampling-uncertainty framing behind the case study.
-   [Binomial standard errors and transformations](m2-scalar-se-binomial.html)
    for the arcsine and boundary-method derivations.
-   [Experimental models for variance smoothing](m6-variance-smoothing-theory.html)
    for the experimental variance model and its simulation findings.
-   [Understanding returned estimates and uncertainty](m8-output-contract.html)
    for the returned columns and covariance-scale requirements.

# Cite the software

```r
citation("sitemix")
```

The bibliography entry used in this article is `lee_2026_sitemix`.
Use `citation("sitemix")` to obtain citation details for your installed version.

# References
