---
title: "A9 · Case study — Alabama Pre-K"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  An end-to-end applied case study using the bundled
  `alprek_subset`. The case study mirrors a realistic project arc:
  question → data prep → site-level estimation → diagnostics →
  optional experimental GVF sensitivity → canonical uncertainty export → methods-
  appendix summary. Use this vignette as the citable URL when you
  write up your own sitemix workflow.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A9 · Case study — Alabama Pre-K}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
data(alprek_subset, package = "sitemix")
```

# Overview

## 1. The reporting question

Asha is the accountability lead for a 50-site Alabama Pre-K
program. The question for the legislative report: **what are
the site-level FRPM (free and reduced-price meals) take-up rates
in 2024, and which sites meet the stated publication criteria?**

The audience is the legislature; the deliverable is a table and
a paragraph in a methods appendix.

## 2. What you will be able to do

By the end of this case study you will be able to complete this full
`sitemix` pipeline:

1.  **Data prep** — slice `alprek_subset` to the year of
    interest.
2.  **Site-level estimation** with `sm_estimate()`.
3.  **Diagnostics** with `sm_diagnose()`.
4.  **Optional sensitivity** with the experimental GVF helper, without
    replacing canonical SEs.
5.  **Direct uncertainty export** with canonical columns.
6.  **Methods-appendix paragraph** for the legislative report.

**Prerequisites.** [A1 · Getting started](a1-getting-started.html) for the
one-call basics; this capstone assembles the full arc.

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 3. Data prep

```{r prep}
target_year <- 2024L
dat <- subset(alprek_subset, year == target_year)
nrow(dat)
length(unique(dat$site_id))
```

Restricting to 2024 leaves 50 sites in this sample.

# 4. Site-level estimation

```{r estimate}
est <- sm_estimate(
  dat,
  family    = "binomial",
  indicator = "frpm"
)
head(est[, c("site_id", "n", "theta_raw", "theta_hat", "se",
             "flag_small_n", "flag_below_accountability")], 5)
```

The default `vst = "arcsine"` returns an arcsine-stabilized estimate
and its standard error; `estimate_scale` records that choice.

# 5. Diagnostics

```{r diagnose}
diag <- sm_diagnose(est, verbose = FALSE)
print(as.data.frame(diag), row.names = FALSE)
```

The summary reports `n_flag_small_n`, `n_flag_below_accountability`,
and intrinsic facts such as `scalar_uncertainty_finite`,
`scalar_se_positive`, and `indicator_scale_consistent`.

# 6. Optional experimental GVF sensitivity

Small sample size alone is not a trigger to smooth. A prespecified simulation
study found no evidence for promoting either GVF backend to the default,
so the reporting workflow proceeds with canonical `se`. If a separate analysis
prespecifies a GVF sensitivity comparison, create an append-only candidate
under a separate object and column name:

```{r smoothing-sensitivity, eval = FALSE}
est_sensitivity <- sm_smooth_variance(
  est,
  method = "loglinear",
  overwrite = FALSE
)
# Keep est_sensitivity$se_smoothed separate from canonical est$se.
```

This optional code is deliberately not part of the default executed path and
makes no general improvement claim.

# 7. Direct uncertainty export

```{r uncertainty-export}
analysis_input <- data.frame(
  unit_id        = est$site_id,
  year           = est$year,
  indicator      = est$indicator,
  estimate       = est$theta_hat,
  std_error      = est$se,
  estimate_scale = est$estimate_scale,
  var_method     = est$var_method
)
# If a prespecified sensitivity was run, add it under a separate name:
# analysis_input$std_error_gvf_sensitivity <- est_sensitivity$se_smoothed
head(analysis_input, 5)
```

The export is an ordinary data frame. Any downstream package can map
`estimate` and `std_error` to its own interface while retaining scale
and variance-method provenance.

# 8. Methods-appendix paragraph (citable template)

> *We computed site-level FRPM rates for the 50 sites in the
> 2024 cohort using the sitemix R package
> [@lee_2026_sitemix].
> Site-level estimates used the arcsine variance-stabilizing
> transform with the default Wilson boundary surrogate
> [@wilson_1927]. Rows below the accountability threshold of
> `n = 30` were flagged but retained in the uncertainty table.
> We exported canonical unsmoothed standard errors and reported the
> estimate scale and variance method with every estimate. Experimental
> GVF alternatives were not substituted for canonical uncertainty.*

# 9. Audit

```{r audit}
stopifnot(nrow(est) == 50L)
stopifnot(all(est$indicator == "frpm"))
stopifnot(diag$scalar_uncertainty_finite == TRUE)
stopifnot(diag$scalar_se_positive == TRUE)
stopifnot(diag$indicator_scale_consistent == TRUE)
stopifnot(all(est$estimate_scale == "arcsine"))
stopifnot(all(is.finite(est$theta_hat)))
stopifnot(nrow(analysis_input) == nrow(est))
stopifnot(identical(analysis_input$std_error, est$se))
stopifnot(!"se_smoothed" %in% names(est))
```

# 10. What's next?

-   [M1 · Statistical foundations](m1-statistical-foundations.html)
    for the sampling-uncertainty framing behind the case study.
-   [M2 · Scalar SE — binomial](m2-scalar-se-binomial.html)
    for the arcsine and boundary-method derivations.
-   [M6 · Variance smoothing theory](m6-variance-smoothing-theory.html)
    for the experimental sensitivity model and its NO-GO simulation decision.
-   [M8 · Output contract](m8-output-contract.html)
    for the package-neutral column and covariance contract.

# 11. Citation

```r
citation("sitemix")
```

A canonical BibTeX entry lives in `vignettes/references.bib`
under the key `lee_2026_sitemix`.

# References
