---
title: "Sampling uncertainty in site-level proportions"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  Define the site-level proportions, standard errors, and covariance
  matrices returned by sitemix. Distinguish the binomial working model
  from finite-population sampling, explain the arcsine approximation,
  and identify assumptions left to a subsequent analysis.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Sampling uncertainty in site-level proportions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

A site-level proportion and its standard error describe different things:
the estimated rate and the uncertainty attributed to sampling. This article
defines those quantities, introduces the binomial working model used by the
basic calculation, and explains why the selected estimate scale matters.
The runnable example uses the fully simulated `prek_sim` data; it does not
provide empirical findings about real children or programs.

# Estimates and their sampling uncertainty

For a panel of sites $j = 1, \ldots, J$ observed across years
$t = 1, \ldots, T$, write the result for each site-year-indicator as:

$$
\mathcal{D}_{jt,k} \;=\; \bigl(\hat\theta_{jt,k}, \;s_{jt,k}, \;n_{jt}, \;V_{jt}\bigr).
$$

$\hat\theta_{jt,k}$ is the direct point estimate on the selected scale for
indicator $k$ at site $j$ in year $t$; $s_{jt,k}$ is its scalar standard
error on the row's `estimate_scale`; $n_{jt}$ is the denominator
(the same across $k$ for student-row scenarios); $V_{jt}$ is the
optional joint covariance across indicators.

The three supported input forms are student records, sufficient counts,
and published aggregates. They lead to the binomial (A), multivariate (B),
multinomial (C), and single- or multiple-marginal aggregate (D0/D1) calculations.
For D1, denominators may differ across indicators; write $n_{jt,k}$ when
that distinction is needed.

The scale of $V_{jt}$ must be read separately from the scale of
$\hat\theta_{jt,k}$. In particular, B and C store raw-proportion covariance
even when the reported estimates use a transformation. The [output guide](m8-output-contract.html)
describes which scalar variance matches the matrix diagonal.

# Notation and returned fields

Let $\pi$ denote a raw probability and $\theta = g(\pi)$ the same quantity
on the selected scale, with $g(p)=p$ for raw output. For identified estimates:


| Symbol | Meaning | Code column | Range |
|:-------|:--------|:------------|:------|
| $j$ | Site index | `site_id` | $1, \ldots, J$ |
| $t$ | Year index | `year` | integer |
| $k$ | Indicator / category index | `indicator` | $1, \ldots, K$ |
| $\pi_{jt,k}$ | Underlying raw probability | not observed | $[0,1]$ |
| $\theta_{jt,k}=g(\pi_{jt,k})$ | Target on the selected scale | not observed | depends on $g$ |
| $\hat\theta_{jt,k}$ | Direct estimate | `theta_hat` | scale-dep. |
| $s_{jt,k}$ | Scalar SE on `estimate_scale` | `se` | $\ge 0$; zero only at an exact SRSWOR census or a boundary cell on the diagnostic `vst = "none"`, `boundary_method = "none"` path |
| $V_{jt}$ | Within-site covariance | `V` (list-col) | PSD; $K\times K$ |
| $n_{jt}$ | Site-year denominator | `n` | $> 0$ integer |
| $C_{jt,k}$ | Numerator (success count) | input `c_jt` or `c_jt_<indicator>` | $0 \le C \le n$ |
| $\hat\pi_{jt,k}$ | Raw-scale proportion | `theta_raw` | $[0, 1]$ |

Suppressed rows can have missing estimate and SE fields; the table is not
a claim that every returned row is identified. Similarly, a recorded bound
on a hidden denominator is not an observed sample size. Each later article
defines additional symbols where they are used.

# A binomial working model

For one binary indicator, suppose the retained observations are independent
Bernoulli trials with a common probability $\pi_{jt}$ within a site-year.
Conditional on the observed denominator, this gives

$$
C_{jt} \mid n_{jt}, \pi_{jt} \;\sim\; \mathrm{Binomial}(n_{jt}, \pi_{jt}),
\tag{M1.1}
$$

where $\pi_{jt} \in [0,1]$. The maximum-likelihood estimate is
$\hat\pi_{jt}=C_{jt}/n_{jt}$. Under this model,
$\operatorname{Var}(\hat\pi_{jt})=\pi_{jt}(1-\pi_{jt})/n_{jt}$.
For an interior probability, applying the first-order delta method to
$g(p)=\arcsin\sqrt p$ gives the baseline working standard error:

$$
\boxed{
\begin{aligned}
\hat\theta_{jt} &\;=\; \arcsin\sqrt{\hat\pi_{jt}}, \\
s_{jt} &\;=\; \frac{1}{2\sqrt{n_{jt}}}.
\end{aligned}}
\tag{M1.2}
$$

The variance result is a first-order approximation, not the exact
finite-sample variance of a transformed binomial count. In sitemix,
(M1.2) is the baseline arcsine calculation without FPC, Anscombe adjustment,
or `binomial_bc`; boundary uncertainty requires the additional rules in
[Binomial standard errors and transformations](m2-scalar-se-binomial.html). The derivative used by the delta method is
not finite at 0 or 1. The [delta-method derivation](https://stat238.berkeley.edu/spring-2026/lectureten238spring2026/)
shows the interior asymptotic result. The transformation work of Anscombe
[@anscombe_1948] and the package's chosen adjustment are discussed separately
in M2.

With complete observed numerators and denominators, D0 and each D1 marginal
use the same scalar calculations under matching options. This numerical
identity does not establish the sampling assumptions of a publisher's data,
and it does not identify D1 cross-indicator covariance.

## Sampling without replacement from a finite population

When the target is a fixed finite population and observations are selected
by simple random sampling without replacement, the success count has a
hypergeometric distribution rather than an exact binomial distribution.
For population size $N_{jt}$, population success fraction $P_{jt}$, and
sample size $n_{jt}$, its variance implies

$$
\operatorname{Var}(\hat\pi_{jt})=
\frac{P_{jt}(1-P_{jt})}{n_{jt}}
\frac{N_{jt}-n_{jt}}{N_{jt}-1}, \qquad N_{jt}>1.
$$

This follows from the [hypergeometric moments](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Hypergeometric.html).
The population fraction in this identity is unknown outside a census.
sitemix uses the supplied `fpc` and observed proportions with the selected
variance rule; plug-in and design-corrected estimates of this variance have
different applied multipliers. A census has zero sampling uncertainty under
that design. M2 explains the estimators and their conditions. Neither a
binomial working model nor an FPC accounts automatically for selection bias,
measurement error, or an unspecified complex sampling design.

# What a subsequent model adds

sitemix supplies direct site estimates and sampling uncertainty. It does not
fit a model for the distribution of site-level rates or shrink their point
estimates. A separate analysis may report rates, compare groups, or fit a
hierarchical model using the estimates and their uncertainty.

For example, the model of Fay and Herriot treats the sampling variances as known
while estimating regression and between-area variation [@fay_herriot_1979,
pp. 271–272]. Treating an estimated sitemix SE as fixed in such a model is
an additional working assumption, not a statement that the sampling variance
has been observed without error.

A Bayesian analysis could use a prior $G$ to calculate

$$
\tilde\theta_{jt} \;=\; \mathrm{E}[\theta_{jt} \mid \hat\theta_{jt}, s_{jt}; G],
\tag{M1.3}
$$

after specifying a likelihood on the same scale. The prior, likelihood,
fitting method, and suitability of that model are decisions outside sitemix.
The optional [variance-smoothing helper](m6-variance-smoothing-theory.html)
models SEs; it does not perform this outcome-model calculation.

# Check an identified example

The following binomial example returns 50 identified estimates. Check that
the expected columns are present, estimates and SEs are finite, the selected
scale is recorded, and row diagnostics contain no error-severity result:

```{r invariants}
data(prek_sim, package = "sitemix")
est <- sm_estimate(
  subset(prek_sim, year == 2024),
  family    = "binomial",
  indicator = "frpm"
)

# Finite estimates in this complete example
stopifnot(all(is.finite(est$theta_hat)))

# Required columns and reported scale
stopifnot(all(c("theta_hat", "se", "estimate_scale") %in% names(est)))
stopifnot(identical(unique(est$estimate_scale), "arcsine"))

# Finite, nonnegative standard errors
stopifnot(all(is.finite(est$se)))
stopifnot(all(est$se >= 0))

# Inspect row diagnostics
diag_r <- sm_diagnose(est, level = "row", verbose = FALSE)
stopifnot(all(c("flag_small_n", "flag_zero_cell") %in% names(diag_r)))
stopifnot(!any(diag_r$diag_severity == "error", na.rm = TRUE))
```

The checks pass silently. They describe this example, not every possible
input: suppressed values remain missing, exact censuses may have zero SE,
and a valid raw covariance can differ in scale from the reported estimates.
For matrix examples, see M3, M4, and M5.

# Assumptions to carry into an analysis

- `se` and `V` are calculated under the selected sampling and variance
  rules. Assess those assumptions against how observations were collected.
- The package does not estimate covariance between different sites or years.
  A model involving such dependence needs additional information or assumptions.
- Keep the chosen estimate, SE, and matrix scales together. A later
  transformation needs a corresponding treatment of uncertainty.
- Missing or separately assumed suppression-sensitivity values are not
  identified estimates. Keep their status when preparing a model input.

# Further derivations

-   [Binomial standard errors and transformations](m2-scalar-se-binomial.html) for
    the delta-method calculation, corrections, and student/count equivalence.
-   [Covariance for overlapping binary indicators](m3-multivariate-sur-covariance.html)
    for the Scenario B extension of $V_{jt}$.
-   [Understanding returned estimates and uncertainty](m8-output-contract.html)
    for returned fields, scales, and diagnostic conditions.
-   [Getting started with site-level proportions](a1-getting-started.html) for the
    first worked estimation example.

# References
