---
title: "M2 · Scalar SE — binomial pipeline"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For methodologists checking the binomial closed-form SE
  pipeline. This vignette derives the arcsine, logit, and
  identity transforms; the Wilson floor and Agresti-Coull
  boundary corrections; and the **T2.5 sufficient-counts
  identity** that ties `sm_estimate()` on student rows to
  `sm_estimate_from_counts()` on aggregated counts.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M2 · Scalar SE — binomial pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

# Overview

This article is written for **methodologists** who need the formulas
behind every binomial standard error `sitemix` reports and *why* each
reporting transform, correction, and boundary policy takes this form. We cover, in
order:

1. the row-level binomial closed form and what `sitemix` encodes;
2. the local notation this derivation adds;
3. the three reporting transforms and their delta-method SEs;
4. the boundary policies for $\hat\pi_{jt} \in \{0, 1\}$;
5. the `var_method` provenance lexicon;
6. the implementation invariants, including the T2.5 sufficient-counts
   identity;
7. the limitations that carry forward.

**Established vs. novel.** *Established:* the arcsine variance-stabilizing
transform [@anscombe_1948], the Wilson score interval and its boundary floor
[@wilson_1927], and the Agresti–Coull adjustment [@agresti_coull_1998].
*This package:* the `var_method` provenance lexicon and the T2.5
sufficient-counts identity linking student-row and count inputs.

| Result | Attribution |
|:-------|:------------|
| arcsine VST | Anscombe (1948) |
| Wilson score / boundary floor | Wilson (1927) |
| Agresti–Coull adjustment | Agresti & Coull (1998) |
| `var_method` lexicon & T2.5 identity | this package (sitemix) |

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 1. What an analyst sees, what `sitemix` encodes

Inside Scenario A, the row-level closed form is

$$
C_{jt} \mid n_{jt}, \pi_{jt} \;\sim\; \mathrm{Binomial}(n_{jt}, \pi_{jt}),
\qquad \hat\pi_{jt} = C_{jt}/n_{jt}.
$$

`sm_estimate()` returns $\hat\pi_{jt}$ in `theta_raw` and the
reported-scale estimate in `theta_hat`. This vignette derives
the SE column $s_{jt}$ for each reporting transform, correction,
finite-population branch, and boundary policy.

# 2. Notation map (extends M1)

| Symbol | Meaning | Code | Range |
|:-------|:--------|:-----|:------|
| $C_{jt}$ | Numerator | `c_jt` | $0 \le C \le n$ |
| $\pi_{jt}$ | True proportion | (latent) | $[0, 1]$ |
| $\hat\pi_{jt}$ | Raw estimate | `theta_raw` | $[0, 1]$ |
| $g(p)$ | Reporting transform | (one of three) | — |
| $g'(p)$ | Delta-method Jacobian | — | $> 0$ on $(0, 1)$ |

# 3. The three reporting transforms

**Proposition 1 (first-order arcsine VST).** *For the arcsine transform*
$g(p) = \arcsin\sqrt{p}$, *the delta-method SE is*

$$
s_{jt} \;=\; \frac{1}{2\sqrt{n_{jt}}}.
\tag{M2.1}
$$

**Proof sketch.** Compute
$g'(p) = 1/(2\sqrt{p(1-p)})$ and apply
$s = g'(\hat\pi) \cdot \sqrt{\hat\pi(1-\hat\pi)/n}$. The product
simplifies; the $\hat\pi(1-\hat\pi)$ cancels. $\square$

**Attribution.** Classical variance-stabilizing transform; see
@anscombe_1948 for the Anscombe correction.

**Proposition 2 (logit reporting transform).** *For*
$g(p) = \log(p/(1-p))$,
*the delta-method SE is*

$$
s_{jt} \;=\; \sqrt{\frac{1}{n_{jt} \hat\pi_{jt} (1 - \hat\pi_{jt})}}.
\tag{M2.2}
$$

**Attribution.** Standard delta-method application.

The logit is not variance-stabilizing: its working SE still depends on
$\hat\pi_{jt}$.

**Identity reporting scale.** $g(p) = p$ returns
$s_{jt} = \sqrt{\hat\pi_{jt}(1-\hat\pi_{jt})/n_{jt}}$.
The identity is also not variance-stabilizing.

## Scope of the displayed baselines

Equations (M2.1) and (M2.2), and the identity-scale expression above,
are first-order, no-FPC, no-bias-correction working formulas for legal
interior cells. The cancellation leading to (M2.1) is exact algebra for
the package's implemented delta-method variance; it is not an exact
finite-sample variance identity for the transformed binomial random
variable.

When a fixed population size $N$ is supplied, the plug-in and boundary
branches use

$$
q_{jt}=\begin{cases}
0, & N_{jt}=n_{jt},\\
(N_{jt}-n_{jt})/(N_{jt}-1), & N_{jt}>n_{jt},
\end{cases}
$$

including the exact one-unit census $N=n=1$, and multiply the applicable
baseline SE by $\sqrt{q_{jt}}$ [@cochran_1977]. For an interior row with
`bias_correction = "binomial_bc"`, the uncorrected denominators $n$ become
$n-1$:

$$
s_{\mathrm{raw,bc}}=\sqrt{\frac{\hat\pi(1-\hat\pi)}{n-1}},\qquad
s_{\mathrm{asin,bc}}=\frac{1}{2\sqrt{n-1}},\qquad
s_{\mathrm{logit,bc}}=\frac{1}{\sqrt{(n-1)\hat\pi(1-\hat\pi)}}.
$$

With a supplied $N$, those corrected interior SEs use the applied
design multiplier $\sqrt{(N-n)/N}$ rather than $\sqrt q$. Boundary rules
take precedence over `binomial_bc` and retain their named surrogate plus
the conventional $\sqrt q$ multiplier. The package records canonical
and actually applied multipliers separately, and finite-population
correction never changes `n_eff`.

# 4. Boundary policies

When $\hat\pi_{jt} \in \{0, 1\}$, $g'$ diverges or the variance
collapses. `sitemix` exposes three policies:

-   **`"wilson_floor"`** (default) — return $s_{jt}$ at the
    boundary using the Wilson score standard error
    $$
    s^{W}(p,n) =
    \frac{\sqrt{p(1-p)/n + z^2/(4n^2)}}{1 + z^2/n},
    $$
    evaluated at $p \in \{0,1\}$ for boundary rows
    [@wilson_1927]. This preserves `theta_raw`; under the default
    arcsine VST, the row-level `se` remains the arcsine
    $1/(2\sqrt{n})$ while `se_raw` records the boundary-safe raw
    fallback.
-   **`"agresti_coull"`** — use the z-general adjusted-Wald
    construction $\tilde n=n+z^2$,
    $\tilde\pi=(C+z^2/2)/\tilde n$, and
    $\sqrt{\tilde\pi(1-\tilde\pi)/\tilde n}$ as a named boundary
    uncertainty surrogate [@agresti_coull_1998]. It does not replace
    observed `theta_raw = C/n` or its raw-scale `theta_hat`.
-   **`"none"`** — diagnostic only; SEs collapse to zero at
    boundaries.

# 5. The `var_method` lexicon

The row-level provenance is scale-specific:

| `var_method`         | Triggered when                                |
|:---------------------|:----------------------------------------------|
| `arcsine_vst`        | `vst = "arcsine"`, without n-1 correction      |
| `arcsine_delta_binomial_bc` | `vst = "arcsine"`, interior `binomial_bc` row |
| `arcsine_anscombe`   | `vst = "arcsine"`, `anscombe = TRUE`           |
| `logit_delta`        | `vst = "logit"`, interior row                  |
| `logit_delta_binomial_bc` | `vst = "logit"`, interior `binomial_bc` row |
| `binomial`           | `vst = "none"`, interior row                   |
| `binomial_bc`        | `bias_correction = "binomial_bc"`              |
| `wilson_boundary_surrogate` | boundary row on raw output, `boundary_method = "wilson_floor"` |
| `agresti_coull_boundary_surrogate` | boundary row on raw output, `boundary_method = "agresti_coull"` |
| `suppressed_drop` | aggregate suppression retained as unavailable |
| `suppression_sensitivity` | acknowledged non-identified variance sensitivity; canonical estimate/SE columns remain missing |

For transformed output, the row-level `var_method` remains the
transform label (for example, `arcsine_vst`); when `vjt = TRUE`,
the associated `sm_vcov` records the boundary rule in
`scalar_correction_rule`.

The lexicon is locked at sm_vcov's `@section var_method
convention`; this vignette is the derivation home.

# 6. Implementation invariants

| ID    | Layer              | Claim                                                              |
|:------|:-------------------|:-------------------------------------------------------------------|
| T1    | engine-binomial    | `theta_hat = asin(sqrt(C/n))` for `vst = "arcsine"`.               |
| T2    | engine-binomial    | `se = 1/(2*sqrt(n))` for `vst = "arcsine"`, interior row.           |
| T2.5  | cross-pipeline     | `sm_estimate(rows)$theta_hat == sm_estimate_from_counts(counts)$theta_hat` to `1e-10`. |
| T3    | engine-binomial    | `theta_hat = log(C/(n-C))` for `vst = "logit"`, interior row.       |
| T4    | wilson floor       | At boundary on raw output, `se` equals the Wilson score SE under `wilson_floor`. |
| T5    | finite population  | An uncorrected interior arcsine SE is multiplied by `sqrt((N-n)/(N-1))`. |
| T6    | corrected FPC      | An interior `binomial_bc` arcsine SE uses denominator `n-1` and multiplier `sqrt((N-n)/N)`. |

Verify on `alprek_subset`:

```{r invariants}
data(alprek_subset, package = "sitemix")
est <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family = "binomial", indicator = "frpm", vst = "arcsine"
)
interior <- which(est$theta_raw > 0 & est$theta_raw < 1)

# T1: closed-form theta_hat
manual_theta <- asin(sqrt(est$theta_raw[interior]))
stopifnot(all.equal(est$theta_hat[interior], manual_theta, tolerance = 1e-10))

# T2: closed-form SE at interior rows
manual_se <- 1 / (2 * sqrt(est$n[interior]))
stopifnot(all.equal(est$se[interior], manual_se, tolerance = 1e-10))

# T2.5: sufficient-counts identity
counts <- readRDS(system.file("extdata", "alprek_subset_counts.rds",
                              package = "sitemix"))
est_counts <- sm_estimate_from_counts(
  counts[counts$year == 2024,
         c("site_id", "year", "n_jt", "c_jt_frpm")],
  family = "binomial", indicator = "frpm"
)
stopifnot(all.equal(est$theta_hat, est_counts$theta_hat, tolerance = 1e-10))
stopifnot(all.equal(est$se, est_counts$se, tolerance = 1e-10))

# T3/T4: use a tiny fixture so boundary and logit behavior are explicit
toy <- data.frame(
  site_id = c(rep("A", 10), rep("B", 10)),
  year = 2024,
  y = c(rep(1, 10), rep(c(1, 0), each = 5))
)
toy_raw <- sm_estimate(
  toy, family = "binomial", indicator = "y",
  vst = "none", boundary_method = "wilson_floor"
)
z <- qnorm(0.975)
manual_wilson <- z / (2 * 10 * (1 + z^2 / 10))
stopifnot(all.equal(toy_raw$se[toy_raw$site_id == "A"],
                    manual_wilson, tolerance = 1e-10))
stopifnot(toy_raw$var_method[toy_raw$site_id == "A"] ==
            "wilson_boundary_surrogate")

toy_logit <- sm_estimate(
  subset(toy, site_id == "B"),
  family = "binomial", indicator = "y", vst = "logit"
)
stopifnot(all.equal(toy_logit$theta_hat, log(0.5 / 0.5), tolerance = 1e-10))

logit_boundary <- tryCatch(
  sm_estimate(toy, family = "binomial", indicator = "y", vst = "logit"),
  error = identity
)
stopifnot(inherits(logit_boundary, "sitemix_error_estimate_var_method"))

# T5/T6: plug-in and design-corrected finite-population branches
fpc_counts <- data.frame(
  site_id = "F", year = 2024L, n_jt = 8L, c_jt_rate = 3L
)
fpc_plugin <- sm_estimate_from_counts(
  fpc_counts, family = "binomial", indicator = "rate",
  vst = "arcsine", fpc = 20, min_n = 1L
)
stopifnot(all.equal(
  fpc_plugin$se,
  (1 / (2 * sqrt(8))) * sqrt(12 / 19),
  tolerance = 1e-12
))
stopifnot(all.equal(fpc_plugin$variance_multiplier_applied, 12 / 19,
                    tolerance = 1e-12))

fpc_corrected <- sm_estimate_from_counts(
  fpc_counts, family = "binomial", indicator = "rate",
  vst = "arcsine", bias_correction = "binomial_bc",
  fpc = 20, min_n = 1L
)
stopifnot(all.equal(
  fpc_corrected$se,
  (1 / (2 * sqrt(7))) * sqrt(12 / 20),
  tolerance = 1e-12
))
stopifnot(all.equal(fpc_corrected$variance_multiplier_applied, 12 / 20,
                    tolerance = 1e-12))
stopifnot(identical(fpc_corrected$n_eff, 8))
```

All seven invariants hold to their documented tolerances.

# 7. Limitations

-   The Anscombe correction uses $(C+3/8)/(n+3/4)$—equivalently,
    it adds $3/8$ to both success and failure counts. The unshifted
    `theta_raw = C/n` is preserved for audit traceability.
-   Logit delta-method requires interior $\hat\pi$; boundary
    rows under `vst = "logit"` raise
    `sitemix_error_estimate_var_method`. Use `vst = "arcsine"` or
    `vst = "none"` when boundary cells are present.

# 8. Where to go next

-   [M3 · Multivariate SUR covariance](m3-multivariate-sur-covariance.html)
    extends Scenario A to overlapping binary indicators.
-   [M5 · Aggregate engines D0 / D1](m5-aggregate-engines.html)
    extends Scenario A to published aggregate inputs (D0 reuses
    this binomial engine).
-   [A3 · Scenario A — binomial](a3-scenario-binomial.html) for
    the applied face of this derivation.

# References
