---
title: "Covariance for overlapping binary indicators"
author: "JoonHo Lee (jlee296@ua.edu)"
date: "2026-07-12"
abstract: >
  Derives the covariance of binary proportions measured on the
  same students and explains how sitemix estimates it from marginal
  and joint counts. Distinguishes the raw-scale matrix from scalar
  transformed estimates, with finite-population corrections and
  matrix checks illustrated on simulated data.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Covariance for overlapping binary indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

# Joint uncertainty for binary proportions

Scenario B estimates several binary proportions for each site-year, such
as participation in FRPM, SNAP, WIC, and TANF. A student can contribute to
more than one indicator. Joint counts describe how the indicators vary
together and determine the off-diagonal entries of their estimated
covariance matrix.

With `family = "multivariate"` and `vjt = TRUE`, `sm_estimate()` returns
this matrix in the `V` list-column. Each indicator row for the same
site-year carries the same $K \times K$ `sm_vcov` object. The derivation
below first assumes independent observations across students; sampling
without replacement from a fixed population is considered afterward.

The example uses `prek_sim`, a simulated 50-site pre-kindergarten panel
included in the package (see `?prek_sim`). Its results describe the
simulation, not real children, sites, or programs. Code runs offline
with a fixed random seed.

# 1. The observations and assumptions

Within one site-year, let $y_i=(X_{i1},\ldots,X_{i K})^\top$ record the $K$
binary indicators for student $i$. Assume the $n=n_{jt}$ student vectors
are independent and identically distributed. Indicators within a vector
may be dependent. Neither measuring them on the same students nor
observing students positive on both indicators guarantees nonzero
covariance.

Marginal counts and joint counts must refer to the same retained students.
A joint count $C_{jt,kk'}$ counts students positive on both $k$ and $k'$.
Knowing only the two marginal counts does not determine that joint count;
equal denominators alone do not establish that the observations were
paired. The independent-marginal aggregate calculation is discussed in
[Sampling uncertainty from published aggregates](m5-aggregate-engines.html).

# 2. Covariance of the sample proportions

The site-year subscripts are suppressed below except where they identify
the sample size or counts.

| Symbol | Meaning |
|:-------|:--------|
| $K$ | Number of binary indicators, `length(indicators)` |
| $\pi_k$, $\pi_{kk'}$ | Population probabilities of indicator $k$ and of both $k$ and $k'$ |
| $\hat\pi_k$ | Sample proportion, returned as `theta_raw` |
| $\hat\pi_{kk'}=C_{jt,kk'}/n$ | Sample joint proportion; `p_xy` in the example |
| $Q$ | Sum of centered student-vector cross-products |
| $\widehat V=\Sigma_{jt}$ | Estimated covariance matrix of the sample proportions, `V[[i]]$matrix` |

For a single student, $X_{i k}X_{i k'}$ is one exactly when both indicators
are positive, so its expectation is $\pi_{kk'}$. Independence across
students eliminates cross-student covariance terms in the covariance of
the sample means. Thus,

$$
\boxed{
\operatorname{Cov}(\hat\pi_k,\hat\pi_{k'})
\;=\; \frac{\pi_{kk'} - \pi_k\pi_{k'}}{n_{jt}},
}
\tag{M3.1}
$$

The covariance is positive, zero, or negative according to whether
$\pi_{kk'}$ is above, equal to, or below $\pi_k\pi_{k'}$. For example,
$\pi_k=\pi_{k'}=1/2$ and $\pi_{kk'}=1/4$ give zero covariance even though
some students are positive on both indicators. Taking $k=k'$ recovers
the familiar variance $\pi_k(1-\pi_k)/n$ of a binomial proportion.

# 3. Estimating the matrix and its relation to SUR

The seemingly unrelated regressions (SUR) method of Zellner uses generalized
least squares to estimate a system of regression equations, allowing
dependence between their disturbances [@zellner_1962]. Here each outcome
can be viewed as an equation with an intercept alone, and the estimated
intercepts are the sample proportions. Equation (M3.1) follows directly
from the covariance of those sample means. This connection explains the
package's `vcov_method = "sur"` label; the following formulas specify the
particular covariance estimates used by `sitemix`.

Replacing the unknown probabilities in (M3.1) with empirical proportions
gives

$$
\widehat V_{\mathrm{plugin},kk'}
=\frac{\hat\pi_{kk'}-\hat\pi_k\hat\pi_{k'}}{n},
\qquad
\widehat V_{\mathrm{plugin}}=\frac{Q}{n^2},
\qquad
Q=\sum_{i=1}^{n}(y_i-\bar y)(y_i-\bar y)^\top.
$$

This is the default matrix before a finite-population correction or
boundary adjustment. It uses $n$ to estimate the covariance of an
individual student vector, then divides by $n$ again to estimate the
covariance of the sample proportions. It is a plug-in estimate, with
expectation $(n-1)/n$ times the population covariance in (M3.1).

For $n>1$, `bias_correction = "binomial_bc"` instead gives
$\widehat V_{\mathrm{bc}}=Q/[n(n-1)]$. In this intercept-only setting,
$Q/(n-1)$ is the unbiased residual covariance estimate, and another
division by $n$ estimates the covariance of the sample means. The
degrees-of-freedom argument here concerns an estimated mean in each
equation; a general SUR regression can have additional, differing
regressors. The name SUR alone does not specify the package's denominator.

These formulas apply to the empirical covariance before the Wilson
boundary surrogate. That option can replace a constant indicator's zero
diagonal, as described below. The returned Scenario B matrix is on the
raw proportion scale, with `vcov_scale = "raw"`, and its diagonal matches
`se_raw^2`. The row's `theta_hat` and `se` use the selected scalar
transformation, which is arcsine by default. Matrix entries therefore
describe uncertainty in `theta_raw`, rather than in those transformed
estimates [@lee_2026_sitemix].

# 4. Sampling without replacement

Supplying `fpc = N` assumes a simple random sample without replacement
(SRSWOR) of the same $n$ students for all indicators from a fixed population
of size $N\ge n$. The population size must refer to that site-year and
sampling frame. Define

$$
q=\begin{cases}
0, & N=n,\\
(N-n)/(N-1), & N>n.
\end{cases}
$$

For a binary count, the hypergeometric variance uses this factor relative
to the binomial population variance; see the Details section of
[R's hypergeometric documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Hypergeometric.html).
The corresponding factor applies to cross-covariances when the indicator
vectors are sampled jointly. Substituting sample proportions gives the
plug-in matrix. Using the sample covariance $Q/(n-1)$ instead gives the
design-unbiased estimate for $n>1$. Before boundary replacement, the
implemented matrices are

$$
\widehat V_{\mathrm{plugin}}=qQ/n^2,\qquad
\widehat V_{\mathrm{design}}=
\begin{cases}
0_{K\times K}, & N=n,\\
(N-n)Q/[Nn(n-1)], & N>n\text{ and }n>1.
\end{cases}
\tag{M3.2}
$$

The second rule is selected by `bias_correction = "binomial_bc"`.
Relative to each rule's no-FPC matrix, the applied variance multipliers
are $q$ for the plug-in rule and $1-n/N$ for `binomial_bc`. Thus multiplying
the bias-corrected no-FPC matrix by $q$ would give a different result.
[Sampling uncertainty in site-level proportions](m1-statistical-foundations.html) develops the distinction for a single
binary indicator.

Each multiplier applies to the whole empirical matrix, including the
off-diagonal entries. A nonnegative multiplier preserves symmetry and
positive semidefiniteness. A positive multiplier also preserves
correlations wherever both variances are positive. At the census $N=n$,
all matrix entries and standard errors are zero, and correlations are
undefined. The explicit census branch includes $N=n=1$; otherwise the
`binomial_bc` matrix requires $n>1$.

For a constant observed indicator, empirical cross-covariances and the
empirical diagonal are zero. The default `boundary_method = "wilson_floor"`
can replace that diagonal with a positive Wilson surrogate, multiplied by
$q$ when FPC is supplied. Even if `binomial_bc` is requested, a boundary
surrogate uses $q$, while interior entries use the selected matrix rule.
At a census, $q=0$ makes this surrogate zero as well. Off-diagonal entries
remain zero for a constant indicator. These substituted diagonals are
regularized uncertainty estimates, so the design-unbiased interpretation
of the empirical formula does not extend to them. Agresti--Coull boundary
uncertainty is not supported with `vjt = TRUE`.

# 5. Joint information and feasible matrices

Scenario B needs at least two binary indicators. With student rows,
`sitemix` obtains the marginal and pairwise counts from the retained
jointly observed vectors. With `from_counts = TRUE`, supply `n_jt`, one
`c_jt_<indicator>` column per indicator, and every pairwise
`c_jt_<first>_<second>` column in the order specified by `indicators`.
Those columns are required even for `vjt = FALSE`.

For two indicators, the pairwise count must satisfy

$$
\max(0,C_k+C_{k'}-n)\le C_{kk'}\le\min(C_k,C_{k'}).
$$

For three indicators, these pairwise bounds alone are insufficient. The
package also checks that some integer count positive on all three
indicators could produce the supplied marginals and pairs. Passing a
positive-semidefinite check alone does not establish the existence of
such a joint table.

| Input | Covariance support |
|:------|:-------------------|
| Jointly observed student rows | Two or more indicators |
| Marginal and pairwise counts, $K=2$ or $K=3$ | Supported after the count-feasibility checks |
| Marginal and pairwise counts, $K\ge4$ | Accepted for scalar output with `vjt = FALSE`; covariance output requires student rows |

For $K\ge4$, the package does not verify full joint feasibility from
pairwise counts alone and rejects that input when `vjt = TRUE` with
`sitemix_error_input_indicator_count`. The counts may be feasible; the
error describes the limit of the available verification.

For observed vectors, $Q$ is a sum of outer products and is positive
semidefinite (PSD), meaning $a^\top Q a\ge0$ for every vector $a$. Feasible
joint counts yield the same construction. Degenerate data can give a
singular matrix and remain valid. Nonnegative whole-matrix scaling and
the nonnegative Wilson diagonal addition preserve PSD. Numerical
roundoff is checked with a tolerance proportional to matrix scale, plus
a machine-range term. Scenario B records `psd_repair = "none"` because it
does not project or otherwise repair the matrix.

# 6. Reading and checking the returned covariance

Use `as.matrix(V[[i]])` to obtain the numeric matrix and
`V[[i]]$indicator_order` to align its rows and columns with the estimates.
For Scenario B, `vcov_method = "sur"`, `vcov_scale = "raw"`, and
`diag_contract = "row_se_raw_squared"`. The numerical rank is recorded
in `matrix_rank`; singularity alone does not make the matrix invalid.

When FPC is supplied, `population_size`, `sampling_fraction`, and
`sampling_design` describe the sampling assumption.
`fpc_variance_multiplier` records the conventional $q$;
`variance_multiplier_applied` records the factor used for the coordinate's
rule. The latter can differ between an interior indicator and a Wilson
boundary indicator. These FPC fields do not change `n_eff`.

The `sitemix_estimates` validator checks the matrix and its recorded
settings against all rows in the site-year group, including the common
denominator, indicator order, scalar corrections, and boundary rules.
For the full list of `sm_vcov` fields, including methods used by other
scenarios, see `?sm_vcov`.

The following example reconstructs the estimated cross-covariance for the
first site's 2024 data. Its `p_xy`, `p_x`, and `p_y` are empirical
proportions, so the comparison checks the plug-in estimate of M3.1.
The subsequent checks examine PSD and the recorded output settings.

```{r invariants}
data(prek_sim, package = "sitemix")
est <- sm_estimate(
  subset(prek_sim, year == 2024),
  family = "multivariate",
  indicators = c("frpm", "snap"),
  vjt = TRUE
)
V1 <- as.matrix(est$V[[1L]])

# Reconstruct the empirical plug-in cross-covariance
slice <- subset(prek_sim, year == 2024 & site_id == est$site_id[1L])
n <- nrow(slice)
p_x <- mean(slice$frpm)
p_y <- mean(slice$snap)
p_xy <- mean(slice$frpm * slice$snap)
manual_sigma <- (p_xy - p_x * p_y) / n
stopifnot(all.equal(V1[1, 2], manual_sigma, tolerance = 1e-10))

# Check PSD using the matrix-scale tolerance
eig_V1 <- eigen(V1, symmetric = TRUE, only.values = TRUE)$values
tol_V1 <- 64 * nrow(V1) * .Machine$double.eps * max(abs(eig_V1)) +
  64 * nrow(V1) * .Machine$double.xmin
stopifnot(min(eig_V1) >= -tol_V1)

# Check the method, scales, and diagonal relation recorded in the output
stopifnot(est$V[[1L]]$vcov_method == "sur")
stopifnot(est$V[[1L]]$vcov_scale == "raw")
stopifnot(est$V[[1L]]$diag_contract == "row_se_raw_squared")
stopifnot(est$estimate_scale[[1L]] == "arcsine")
stopifnot(est$V[[1L]]$psd_repair == "none")
```

The first check compares two ways to calculate the same sample estimate,
with tolerance `1e-10`. It does not compare the estimate with unknown
population probabilities. The PSD check permits only the small negative
eigenvalues allowed by the matrix-scale tolerance. The final checks confirm
that the matrix describes raw proportions even though the scalar output
uses the default arcsine transformation.

```{r raw-diagonal}
first_site <- subset(est, site_id == est$site_id[1L])
first_site <- first_site[match(est$V[[1L]]$indicator_order,
                               first_site$indicator), ]
stopifnot(all.equal(unname(diag(V1)), first_site$se_raw^2,
                    tolerance = 1e-10))
data.frame(
  indicator = first_site$indicator,
  matrix_variance_raw = unname(diag(V1)),
  scalar_variance_raw = first_site$se_raw^2,
  scalar_variance_arcsine = first_site$se^2
)
```

The two raw-variance columns match, including any boundary surrogate used
by Scenario B. The arcsine variances describe uncertainty on a different
scale and should not be substituted for the raw diagonal.

## Check both finite-population rules

For a numerical comparison, suppose the same first-site sample was drawn
without replacement from a population twice its sample size. This
population size is an example assumption, not information in `prek_sim`.
Both observed proportions in this slice are interior, so no boundary
diagonal is replaced.

```{r finite-population-matrix}
N <- 2 * n
finite_plugin <- sm_estimate(
  slice, family = "multivariate", indicators = c("frpm", "snap"),
  fpc = N, vjt = TRUE
)
finite_bc <- sm_estimate(
  slice, family = "multivariate", indicators = c("frpm", "snap"),
  bias_correction = "binomial_bc", fpc = N, vjt = TRUE
)
q <- (N - n) / (N - 1)
expected_bc <- V1 * n / (n - 1) * (1 - n / N)
stopifnot(all.equal(as.matrix(finite_plugin$V[[1L]]), V1 * q,
                    tolerance = 1e-10))
stopifnot(all.equal(as.matrix(finite_bc$V[[1L]]), expected_bc,
                    tolerance = 1e-10))
data.frame(
  rule = c("plugin", "binomial_bc"),
  fpc_variance_multiplier = c(
    finite_plugin$V[[1L]]$fpc_variance_multiplier[1L],
    finite_bc$V[[1L]]$fpc_variance_multiplier[1L]
  ),
  variance_multiplier_applied = c(
    finite_plugin$V[[1L]]$variance_multiplier_applied[1L],
    finite_bc$V[[1L]]$variance_multiplier_applied[1L]
  )
)

census <- sm_estimate(
  slice, family = "multivariate", indicators = c("frpm", "snap"),
  fpc = n, vjt = TRUE
)
stopifnot(all(as.matrix(census$V[[1L]]) == 0),
          all(census$se_raw == 0), all(census$se == 0))
```

The matrix comparisons include every off-diagonal entry. The displayed
FPC fields distinguish the conventional factor from the factor actually
used with `binomial_bc`. The final call treats the observed students as
the entire population and returns zero sampling uncertainty for that
population mean.

# 7. Where to go next

-   [Covariance for multinomial proportions](m4-multinomial-simplex.html) —
    covariance for mutually exclusive, exhaustive categories.
-   [Sampling uncertainty from published aggregates](m5-aggregate-engines.html)
    for the D1 working-independence variant of $V$.
-   [Analyzing overlapping indicators and mutually exclusive categories](a4-multivariate-multinomial.html)
    for examples of reading binary and categorical estimates together
    with their covariance matrices.

# References
