---
title: "M3 · Multivariate SUR covariance"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For methodologists working on Scenario B (overlapping binary
  indicators with SUR-style covariance). Derives the
  cross-indicator covariance matrix, locks the `sm_vcov` class
  spec, and runs the PSD invariant on `alprek_subset`.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M3 · Multivariate SUR covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

# Overview

This article is written for **methodologists** working on Scenario B —
overlapping binary indicators whose within-site cross-covariance is
non-zero — who need the exact covariance matrix `sitemix` assembles and
*why* it takes this form. We cover, in order:

1. the joint within-site structure an analyst sees and what `sitemix`
   encodes;
2. the local notation this derivation adds;
3. the SUR cross-covariance derivation and its finite-population
   corrections;
4. the `sm_vcov` class spec;
5. PSD validation of the assembled matrix;
6. the implementation invariants.

**Established vs. novel.** *Established:* the
seemingly-unrelated-regression cross-covariance structure [@zellner_1962]
and design-based finite-population corrections under SRSWOR. *This package:*
the `sm_vcov` object contract — scale/method provenance, PSD validation, and
the diagonal exceptions — that packages these results for site-year panels.

| Result | Attribution |
|:-------|:------------|
| SUR cross-covariance structure | Zellner (1962) |
| SRSWOR finite-population correction | standard design-based sampling |
| `sm_vcov` object contract & PSD validation | this package (sitemix) |

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 1. What an analyst sees, what `sitemix` encodes

For Scenario B with $K$ overlapping binary indicators (e.g.,
FRPM, SNAP, WIC, TANF), the within-site joint distribution is
*not* a simple product of binomials — students can be in
multiple indicators simultaneously, so the per-row covariance
across indicators is non-zero. `sm_estimate(family =
"multivariate", vjt = TRUE)` returns a `V` list-column whose
$j$-th element is a $K \times K$ `sm_vcov` object.

# 2. Notation map (extends M2)

| Symbol | Meaning | Code | Range |
|:-------|:--------|:-----|:------|
| $K$ | Indicator-component count | `length(indicators)` | $\ge 2$ |
| $C_{jt,kk'}$ | Joint count (both $k$ and $k'$ positive) | (intermediate) | $\ge 0$ |
| $p_{xy}$ | Joint proportion estimate | (intermediate) | $[0, \min(p_x, p_y)]$ |
| $\sigma_{kk'}$ | Cross-covariance element | `V[[i]]$matrix[k,k']` | symmetric |
| $\Sigma_{jt}$ | Assembled $K \times K$ matrix | `V[[i]]$matrix` | PSD up to `1e-12` |

# 3. The SUR cross-covariance derivation

**Theorem 1 (population SUR cross-covariance).** *For two overlapping
binary indicators $X_k$ and $X_{k'}$ measured on the same $n$
IID students per site-year,*

$$
\boxed{
\operatorname{Cov}(\hat\pi_k,\hat\pi_{k'})
\;=\; \frac{\pi_{kk'} - \pi_k\pi_{k'}}{n_{jt}},
}
\tag{M3.1}
$$

*where $\pi_k=\Pr(X_k=1)$, $\pi_{k'}=\Pr(X_{k'}=1)$, and
$\pi_{kk'}=\Pr(X_k=1,X_{k'}=1)$ are population probabilities.*

**Proof sketch.** Treat each student as a Bernoulli trial in
$2^K$-cell multinomial space. The sample covariance of the two
binary marginals reduces to $\mathrm{Cov}(X_k, X_{k'})/n$,
which simplifies as above. $\square$

**Attribution.** Standard SUR result; see @zellner_1962.

The package does not observe the population probabilities in (M3.1).
It substitutes the empirical proportions. If
$Q=\sum_i(y_i-\bar y)(y_i-\bar y)^\top$, the default no-FPC matrix is
the plug-in estimate $\widehat V_{\mathrm{plugin}}=Q/n^2$; the
`binomial_bc` alternative is $Q/[n(n-1)]$. Its diagonal therefore
recovers the corresponding estimated Scenario A raw-scale variance,
not the unknown population variance itself. This raw-scale matrix
convention is the one used for Scenario B in `sitemix`
[@lee_2026_sitemix].

For a fixed population of size $N$ sampled without replacement, define

$$
q=\begin{cases}
0, & N=n,\\
(N-n)/(N-1), & N>n.
\end{cases}
$$

The first branch includes the exact one-unit census $N=n=1$. The approved SRSWOR
whole-matrix rules are

$$
\widehat V_{\mathrm{plugin}}=qQ/n^2,\qquad
\widehat V_{\mathrm{design}}=
\begin{cases}
0_{K\times K}, & N=n,\\
(N-n)Q/[Nn(n-1)], & N>n\text{ and }n>1.
\end{cases}
\tag{M3.2}
$$

These are standard without-replacement sampling corrections
[@cochran_1977].

The second rule is selected by **bias_correction = "binomial_bc"**.
Both scale the entire matrix rather than replacing only its diagonal, so
symmetry, PSD, and correlation structure are preserved. **N = n** is a valid
census and produces a zero matrix. A Wilson boundary surrogate remains a
named scalar/diagonal regularizer and carries the conventional $q$ multiplier.

The four implemented interior-cell cases are therefore:

| Variance rule | Infinite-population / no supplied $N$ | Fixed-population SRSWOR |
|:--------------|:---------------------------------------|:--------------------------|
| plug-in | $Q/n^2$ | $qQ/n^2$ |
| `binomial_bc` | $Q/[n(n-1)]$ | $(N-n)Q/[Nn(n-1)]$ |

The same multiplier applies to every matrix entry. A diagonal-only
correction would change correlations and is not the implemented estimand.
For a constant indicator, empirical cross-covariances are exactly zero.
`boundary_method = "wilson_floor"` may replace that indicator's zero diagonal
with a positive, explicitly named scalar/diagonal surrogate, but it never
creates off-diagonal covariance. Agresti--Coull boundary regularization is
illegal with `vjt = TRUE`.

# 4. The `sm_vcov` class spec (canonical home)

This vignette is the method location for the `sm_vcov` class
spec. The primary fields used by Scenario B are:

-   **`matrix`** — the $K \times K$ symmetric PSD numeric
    matrix.
-   **`vcov_scale`** — one of `"raw"`, `"arcsine_delta"`,
    `"logit_delta"`, or reserved `"reference_raw"`. Under
    Scenario B's SUR construction, `vcov_scale = "raw"` regardless
    of the row's
    `estimate_scale` (the SUR formula is numerically stable in
    raw space).
-   **`vcov_method`** — set to `"sur"` for Scenario B output.
-   **`indicator_order`** and **`matrix_rank`** — record the
    component order and numerical rank.
-   **Finite-population provenance** — population size, coordinate-aligned
    sampling fraction, conventional FPC, actually applied multiplier,
    sampling design, and variance rule. These fields do not alter n_eff.
-   **Diagonal contract** — "row_se_raw_squared" for Scenario B, making
    explicit that the raw matrix diagonal matches se_raw squared, not
    transformed se squared.

For a package-produced Scenario B `sitemix_estimates` object, validation
cross-checks those matrix fields against every row, including the site-year,
estimate scale, common denominator, scalar correction rule, and boundary
rule. Repeating the same tampered `V` over all group rows therefore cannot
bypass the contract. A stand-alone user-constructed `sm_vcov` remains a more
general carrier until it is attached to package output.

The full lexicons are documented at `?sm_vcov`: `vcov_method`
is `NA`, `"sur"`, `"multinomial"`, or
`"working_independence"`; `vcov_scale` is `"raw"`,
`"arcsine_delta"`, `"logit_delta"`, or `"reference_raw"`.

# 5. PSD validation

The assembled $\Sigma_{jt}$ is PSD by construction for
non-degenerate joint counts. Numerical roundoff can produce a
smallest eigenvalue slightly below zero; the constructor validates
PSD-ness up to the documented tolerance. The Scenario B path records
`psd_repair = "none"` because no repair is applied.

# 6. Implementation invariants

| ID    | Layer              | Claim                                                             |
|:------|:-------------------|:------------------------------------------------------------------|
| SUR1  | engine-multivariate | (M3.1) holds exactly for two indicators sharing the same $n$.      |
| SUR2  | sm_vcov            | Every assembled `V[[i]]$matrix` satisfies the documented scale-aware PSD tolerance. |
| SUR3  | sm_vcov            | `vcov_method = "sur"` and `vcov_scale = "raw"` for Scenario B.    |
| SUR4  | correction         | Plug-in and `binomial_bc` scale the whole matrix, including off-diagonals. |
| SUR5  | diagonal contract  | `diag(V) = se_raw^2`; transformed row `se^2` is intentionally different. |
| SUR6  | boundary           | Constant-indicator off-diagonals are zero; Wilson is a named diagonal surrogate. |

Verify on `alprek_subset`:

```{r invariants}
data(alprek_subset, package = "sitemix")
est <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family = "multivariate",
  indicators = c("frpm", "snap"),
  vjt = TRUE
)
V1 <- as.matrix(est$V[[1L]])

# SUR1: closed form cross-covariance
slice <- subset(alprek_subset, year == 2024 & site_id == est$site_id[1L])
n <- nrow(slice)
p_x <- mean(slice$frpm)
p_y <- mean(slice$snap)
p_xy <- mean(slice$frpm * slice$snap)
manual_sigma <- (p_xy - p_x * p_y) / n
stopifnot(all.equal(V1[1, 2], manual_sigma, tolerance = 1e-10))

# SUR2: PSD
eig_V1 <- eigen(V1, symmetric = TRUE, only.values = TRUE)$values
tol_V1 <- 64 * nrow(V1) * .Machine$double.eps * max(abs(eig_V1)) +
  64 * nrow(V1) * .Machine$double.xmin
stopifnot(min(eig_V1) >= -tol_V1)

# SUR3: lexicon
stopifnot(est$V[[1L]]$vcov_method == "sur")
stopifnot(est$V[[1L]]$vcov_scale == "raw")
stopifnot(est$V[[1L]]$diag_contract == "row_se_raw_squared")
stopifnot(est$estimate_scale[[1L]] == "arcsine")
stopifnot(est$V[[1L]]$psd_repair == "none")
```

The three checked invariants (SUR1–SUR3) hold (algebraic tolerance
`1e-10`; PSD uses the scale-aware relative plus machine-range rule
above). SUR4–SUR6 in the table above are stated but not checked in
this chunk.

# 7. Where to go next

-   [M4 · Multinomial simplex](m4-multinomial-simplex.html) —
    the sister derivation for Scenario C (mutually exclusive
    categories).
-   [M5 · Aggregate engines D0 / D1](m5-aggregate-engines.html)
    for the D1 working-independence variant of $V$.
-   [A4 · Scenarios B / C — multivariate / multinomial](a4-multivariate-multinomial.html)
    for the applied face of (M3.1) and the `vcov_scale = "raw"`
    convention.

# References
