---
title: "M4 · Multinomial simplex covariance"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For methodologists working on Scenario C (mutually exclusive
  categories summing to the site-year denominator). Derives the
  simplex covariance, documents the rank deficiency, and runs
  the simplex row-sum invariant on a synthetic example.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M4 · Multinomial simplex covariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
library(sitemix)
```

# Overview

This article is written for **methodologists** working on Scenario C —
mutually exclusive categories that sum to the site-year denominator — who
need the exact simplex covariance `sitemix` assembles and *why* it takes
this form. We cover, in order:

1. the within-site multinomial model an analyst sees and what `sitemix`
   encodes;
2. the local notation this derivation adds;
3. the simplex covariance and its finite-population corrections;
4. the rank deficiency and the analytic support rank;
5. the implementation invariants.

**Established vs. novel.** *Established:* the standard multinomial covariance,
simplex kernel, and support-rank result ($S-1$ on positive support)
[@agresti2013categorical]. *This package:* recording that analytic support rank
as `matrix_rank`, retaining it as structural metadata when a census makes the
realized covariance zero, and the diagonal-contract exception for zero-support
categories.

| Result | Attribution |
|:-------|:------------|
| multinomial covariance structure | standard [@agresti2013categorical] |
| full-simplex support rank ($S-1$) | standard [@agresti2013categorical] |
| analytic `matrix_rank` metadata & boundary diagonal exception | this package (sitemix) |

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 1. What an analyst sees, what `sitemix` encodes

For Scenario C with $K$ mutually exclusive categories (e.g.,
primary language $\in$ {English, Spanish, Other}), the
within-site joint distribution is multinomial:

$$
(C_{jt,1}, \ldots, C_{jt,K}) \mid n_{jt}, \pi_{jt}
\;\sim\; \mathrm{Multinomial}(n_{jt}, \pi_{jt}),
$$

with the simplex constraint $\sum_k \pi_{jt,k} = 1$.

# 2. Notation map (extends M2)

| Symbol | Meaning | Code | Range |
|:-------|:--------|:-----|:------|
| $\pi_{jt,k}$ | Category-$k$ probability | `theta_raw` (per row) | $\sum_k = 1$ |
| $C_{jt,k}$ | Category-$k$ count | `c_jt_k` | $\sum_k = n_{jt}$ |
| $\Sigma_{jt}$ | Simplex covariance | `V[[i]]$matrix` | analytic rank $S-1$ for positive support $S$ |

# 3. The simplex covariance

**Proposition 1 (Multinomial covariance).** *For the multinomial
sampling model on $K$ mutually exclusive categories,*

$$
\boxed{
\Sigma_{jt} \;=\; \frac{1}{n_{jt}}\bigl(\mathrm{diag}(\pi_{jt}) - \pi_{jt}\pi_{jt}^{\top}\bigr).
}
\tag{M4.1}
$$

**Proof sketch.** Direct from the moment-generating function of
the multinomial; the variance of $C_{jt,k}$ is
$n \pi_k (1 - \pi_k)$ and the covariance of $C_{jt,k}$ and
$C_{jt,k'}$ (for $k \ne k'$) is $-n \pi_k \pi_{k'}$. Divide by
$n^2$ to obtain the covariance matrix of the estimated category
proportions. $\square$

**Attribution.** Classical multinomial-sampling theory; see
@agresti2013categorical.

Under fixed-population SRSWOR with population size $N$, write
$M=\mathrm{diag}(\hat\pi)-\hat\pi\hat\pi^\top$ and define

$$
q=\begin{cases}
0, & N=n,\\
(N-n)/(N-1), & N>n.
\end{cases}
$$

The first branch includes $N=n=1$. The plug-in and
design-corrected whole-matrix rules are

$$
\widehat V_{\mathrm{plugin}}=qM/n,\qquad
\widehat V_{\mathrm{design}}=
\begin{cases}
0_{K\times K}, & N=n,\\
(N-n)M/[N(n-1)], & N>n\text{ and }n>1.
\end{cases}
\tag{M4.2}
$$

The finite-population factors follow standard SRSWOR theory
[@cochran_1977].

The design-corrected rule is selected by
**bias_correction = "binomial_bc"**. Multiplying the whole matrix preserves
PSD and the simplex identity $V\mathbf 1=0$; a census (**N = n**) has zero
sampling covariance. The covariance metadata records both the conventional
FPC and the actually applied plug-in/design multiplier without changing
n_eff. Without a supplied finite-population design, the corresponding rules
are $M/n$ and $M/(n-1)$.

# 4. Rank deficiency

Let $S=\#\{k:C_{jt,k}>0\}$ be the observed positive support. The simplex
constraint $\sum_k \pi_{jt,k} = 1$ implies
$\mathbf{1}^{\top}\Sigma_{jt} = \mathbf{0}$ (the all-ones
vector is in the kernel), and every zero-count category contributes a zero
row and column. Therefore the analytic simplex support rank is $S-1$:
full support has rank $K-1$, one zero category has rank $K-2$, and degenerate
support $S=1$ has rank zero. `matrix_rank` records this analytic rank rather
than a tolerance-dependent eigendecomposition. In a census, the realized
sampling covariance is exactly zero but `matrix_rank` still records $S-1$;
`sampling_design` and `sampling_fraction` make that distinction explicit.
The constructor validates PSD-ness, the simplex kernel, and the support-rank
metadata; no repair is applied in the current Scenario C path.

The `sm_vcov` constructor records `vcov_method = "multinomial"`
and `vcov_scale = "raw"`. As in Scenario B, the matrix is on
the raw scale even when `theta_hat` is on the arcsine scale; this
matches sitemix's Scenario C contract
[@lee_2026_sitemix].

For interior categories, the diagonal contract is
"row_se_raw_squared". At a zero-support category, the Wilson scalar
surrogate remains positive while the matrix diagonal remains zero to preserve
the simplex. That intentional exception is recorded as
"row_se_raw_squared_except_boundary_surrogates". With whole-matrix
`binomial_bc`, `V$variance_rule` and the applied multiplier record one uniform
global design-corrected matrix rule. The zero row and column are unchanged by
that scaling, while their row-level scalar provenance remains plug-in or a
Wilson surrogate. Validation permits this intentional scalar/matrix
provenance mismatch only when the matrix row and column are exactly zero.

# 5. Implementation invariants

| ID  | Layer            | Claim                                                          |
|:----|:-----------------|:----------------------------------------------------------------|
| MN1 | engine-multinomial | Row sums of `theta_raw` across categories equal 1 (tolerance `1e-12`). |
| MN2 | sm_vcov          | Row sums of `V[[i]]$matrix` equal 0 (tolerance `1e-12`).         |
| MN3 | sm_vcov          | `vcov_method = "multinomial"` and `vcov_scale = "raw"`.          |
| MN4 | sm_vcov          | `matrix_rank = positive_support - 1` (analytic support rank).     |
| MN5 | sm_vcov          | Whole-matrix `variance_rule` and applied multiplier are uniform.  |

Verify on a synthetic primary-language slice:

```{r invariants}
data(alprek_subset, package = "sitemix")
language_data <- transform(
  alprek_subset,
  primary_language = sample(c("english", "spanish", "other"),
                            nrow(alprek_subset), replace = TRUE,
                            prob = c(0.70, 0.25, 0.05))
)
est <- sm_estimate(
  subset(language_data, year == 2024),
  family    = "multinomial",
  indicator = "primary_language",
  vjt       = TRUE
)
V1 <- as.matrix(est$V[[1L]])

# MN1: rows of theta_raw across categories sum to 1 per site-year
per_site <- aggregate(theta_raw ~ site_id, data = est, FUN = sum)
stopifnot(all(abs(per_site$theta_raw - 1) < 1e-12))

# MN2: row sums of Sigma are zero
stopifnot(max(abs(rowSums(V1))) < 1e-12)

# MN3: lexicon
stopifnot(est$V[[1L]]$vcov_method == "multinomial")
stopifnot(est$V[[1L]]$vcov_scale == "raw")
stopifnot(est$V[[1L]]$psd_repair == "none")

# MN4: analytic support rank
stopifnot(est$V[[1L]]$matrix_rank == est$V[[1L]]$positive_support - 1L)

# MN5: one global matrix variance rule and multiplier
stopifnot(length(unique(est$V[[1L]]$variance_rule)) == 1L)
stopifnot(length(unique(est$V[[1L]]$variance_multiplier_applied)) == 1L)
```

All five invariants hold (`1e-12` exact tolerance where numerical).

# 6. Where to go next

-   [M3 · Multivariate SUR covariance](m3-multivariate-sur-covariance.html)
    — the sibling derivation for Scenario B (overlapping rather
    than mutually exclusive).
-   [M5 · Aggregate engines D0 / D1](m5-aggregate-engines.html)
    for the aggregate-input analog (cross-marginal covariance
    unidentified). D1 is a multivariate marginal sensitivity path, not
    aggregate multinomial composition; Scenario C requires student rows or
    complete sufficient category counts.
-   [A4 · Scenarios B / C — multivariate / multinomial](a4-multivariate-multinomial.html)
    for the applied face including the simplex row-sum invariant.

# References
