---
title: "M5 · Aggregate engines D0 / D1"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For methodologists working from publisher-aggregate inputs.
  Formalizes the D0 (single-indicator binomial pass-through)
  and D1 (multi-marginal working-independence) engines, proves
  the non-identification of cross-marginal covariance from
  marginals alone, and notes that publisher suppression is an
  operational concept handled outside the formal model.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M5 · Aggregate engines D0 / D1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
```

# Overview

This article is written for **methodologists** working from
publisher-aggregate inputs — per-site-year counts already aggregated by the
publisher — who need the exact engines `sitemix` dispatches for the D0 and D1
cases and *why* they take this form. We cover, in order:

1. the aggregate inputs an analyst sees and what `sitemix` encodes;
2. the local notation this derivation adds;
3. the D0 single-indicator binomial pass-through engine;
4. the D1 multi-marginal working-independence engine and its regime mapping;
5. suppression as an operational (non-statistical) layer;
6. the implementation invariants.

**Established vs. novel.** *Established:* the Fréchet–Hoeffding
non-identification result for aggregate marginals
[@frechet_1951; @hoeffding_1940]. *This package:* the D0 pass-through and D1
working-independence conventions, the formal-D1a vs heuristic-D1b regime
mapping, and suppression as an operational (non-statistical) layer.

| Result | Attribution |
|:-------|:------------|
| Fréchet–Hoeffding non-identification | Fréchet (1951), Hoeffding (1940) |
| D0/D1 engines + regime mapping + suppression layer | this package (sitemix) |

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 1. What an analyst sees, what `sitemix` encodes

Aggregate inputs differ from student-row inputs in what they
hand the estimator: not a per-student table, but per-site-year
counts already aggregated by the publisher. `sitemix` supports
two cases:

-   **D0** — one numerator and denominator per site-year for a
    single indicator.
-   **D1** — multiple aggregate marginals per site-year; the
    cross-marginal joints are not in the file.

The two engines share the aggregate-input parsing pipeline but
differ in covariance assembly.

# 2. Notation map

| Symbol | Meaning | Code | Range |
|:-------|:--------|:-----|:------|
| $C_{jt}^{\mathrm{agg}}$ | D0 numerator | `c_jt` | $\ge 0$ |
| $n_{jt}^{\mathrm{agg}}$ | D0 denominator | `n_jt` | $> 0$ |
| $C_{jt,k}$ (D1) | Marginal-$k$ numerator | `c_jt_k` | $\ge 0$ |
| $n_{jt,k}$ (D1) | Marginal-$k$ denominator | `n_jt_k` | $> 0$ |
| $\Sigma_{jt}^{\mathrm{D1}}$ | D1 covariance | `V[[i]]$matrix` | diagonal |

# 3. D0 engine — binomial pass-through

**Proposition 1 (D0 reuses Scenario A).** *The D0 engine
applied to $(C_{jt}^{\mathrm{agg}}, n_{jt}^{\mathrm{agg}})$
and the Scenario A counts engine produce identical canonical numerical
fields when counts are valid, non-suppressed and integer-identical and all
estimator options match. Formally, for every such row,*

$$
\hat\pi_{jt}^{D0} \;=\; C_{jt}^{\mathrm{agg}} / n_{jt}^{\mathrm{agg}},
\qquad
s_{jt}^{D0} = s_{jt}^{A},
\tag{M5.1}
$$

*to numerical tolerance.*

The invariant covers `theta_raw`, `theta_hat`, `se_raw`, `se`, and `n`.
Input-path provenance, aggregate-case metadata, suppression state, object
attributes, and path-specific validation errors can differ and are not part
of the identity.

**Attribution.** Standard pass-through; the binomial sampling
model (M1.1) is invariant to whether $C$ arrives as a sum of
student rows or a publisher-aggregated count.

# 4. D1 engine — working independence

For multiple aggregate marginals per site-year, the cross-
marginal covariance $\sigma_{kk'}$ requires the *joint* count
$C_{jt,kk'}$, which the publisher did not provide. The D1
engine therefore emits a **working-independence diagonal**:

$$
\boxed{
\Sigma_{jt}^{D1} \;=\; \mathrm{diag}\bigl(s_{jt,1}^2, \ldots, s_{jt,K}^2\bigr).
}
\tag{M5.2}
$$

The off-diagonals are set to zero as an explicit working assumption; they are
not identified zeros. `vcov_method = "working_independence"` records this
choice.

With a keyed site-year population size **fpc = N**, each D1 diagonal uses
its own marginal denominator $n_k$. Plug-in rows use

$$
q_k=\begin{cases}
0, & N=n_k,\\
(N-n_k)/(N-1), & N>n_k,
\end{cases}
$$

including $N=n_k=1$ [@cochran_1977]; binomial_bc rows use the design multiplier
$(N-n_k)/N$. The single covariance object stores one population size plus
coordinate-aligned sampling fractions and multipliers. Thus varying
denominators remain visible, while n_jt and n_eff stay missing at matrix
level when there is no honest common scalar. The diagonal contract
"row_se_squared" verifies that every working-independence diagonal equals
its row's scale-specific squared SE.

Sampling-unit provenance is a separate fact from denominator shape. The D1
API accepts `sampling_relation = "same_units"`, `"different_units"`, or
`"unknown"` (the default), mapping to `d1_regime = "D1a"`, `"D1b"`, or
`"unknown"`. Equal denominators alone never establish that the marginals were
observed on the same units. `denominator_pattern` records `"common"` or
`"varying"` independently, and `d1_regime_by_group` preserves both facts for
every site-year. Every D1 group must contain the same ordered indicator set.

**Proposition 2 (Non-identification).** *Without
$C_{jt,kk'}$, the cross-marginal covariance is unidentified
from the marginals. Any value in the pairwise Fréchet-Hoeffding interval
is feasible pairwise.*

**Proof sketch.** Construct two joint distributions on
$\{0,1\}^K$ that share marginals but differ in cross-covariance.
The Fréchet-Hoeffding bounds give the pairwise range
[@frechet_1951; @hoeffding_1940]. For $K > 2$, pairwise feasible
choices need not define a globally compatible joint distribution or
PSD covariance matrix; that is why M7 preserves the formal raw pairwise
intervals but labels PSD-projected corners only as stress scenarios. $\square$

**Attribution.** Standard copula-theory result; see M7 and
[@nelsen2006introduction].

# 5. Suppression as an operational layer

Tier 1 (publisher-suppressed), Tier 2 (observed below
accountability), and Tier 3 (observed and publishable) are
**operational, not statistical** concepts. They live in the
`sm_suppression_report()` audit, not in the
statistical model derived here. `suppression = "drop"` retains an
unavailable audit row with canonical point and SE fields missing.
The legacy `"upper_bound"` label is available only as an explicitly
acknowledged, separated worst-case variance sensitivity; it does not
create an identified point, ordinary covariance, or formal Fréchet
input. A hidden denominator cannot support a numeric variance claim.
The formal model in §3–§4 uses identified rows only.

# 6. Implementation invariants

| ID  | Layer              | Claim                                                          |
|:----|:-------------------|:---------------------------------------------------------------|
| AG1 | engine-aggregate-d0 | (M5.1) — qualified canonical numerical fields agree to tolerance; path provenance may differ. |
| AG2 | engine-aggregate-d1 | (M5.2) — D1 `V` off-diagonals are exactly zero.                  |
| AG3 | engine-aggregate-d1 | `vcov_method = "working_independence"` and diagonal equals row `se^2`. |

Verify on `alprek_subset`:

```{r invariants}
counts <- readRDS(system.file("extdata", "alprek_subset_counts.rds",
                              package = "sitemix"))

# AG1: D0 = Scenario A
d0 <- counts[counts$year == 2024, c("site_id", "year", "n_jt", "c_jt_frpm")]
d0$indicator <- "frpm"
d0$c_jt <- d0$c_jt_frpm
d0 <- d0[c("site_id", "year", "indicator", "c_jt", "n_jt")]

est_a <- sm_estimate_from_counts(
  counts[counts$year == 2024, c("site_id", "year", "n_jt", "c_jt_frpm")],
  family = "binomial", indicator = "frpm"
)
est_d0 <- sm_estimate_from_aggregates(
  d0, family = "binomial", indicator = "frpm"
)
canonical_numeric <- c("theta_raw", "theta_hat", "se_raw", "se", "n")
stopifnot(all(vapply(canonical_numeric, function(field) {
  isTRUE(all.equal(est_a[[field]], est_d0[[field]], tolerance = 1e-10))
}, logical(1))))
stopifnot(!identical(est_a$input_mode, est_d0$input_mode))

# AG2 + AG3: D1 working independence
d1_long <- rbind(
  data.frame(site_id = d0$site_id, year = d0$year,
             indicator = "frpm",
             c_jt = counts[counts$year == 2024, ]$c_jt_frpm,
             n_jt = counts[counts$year == 2024, ]$n_jt),
  data.frame(site_id = d0$site_id, year = d0$year,
             indicator = "snap",
             c_jt = counts[counts$year == 2024, ]$c_jt_snap,
             n_jt = counts[counts$year == 2024, ]$n_jt)
)
est_d1 <- capture_expected_sitemix_warning(
  sm_estimate_from_aggregates(
    d1_long, family = "multivariate",
    indicator_col = "indicator",
    sampling_relation = "same_units",
    vjt = TRUE
  ),
  "sitemix_warning_working_independence_default"
)
V1 <- as.matrix(est_d1$V[[1L]])
grp <- est_d1[est_d1$site_id == est_d1$site_id[1L] &
                est_d1$year == est_d1$year[1L], ]
off_diag <- V1[upper.tri(V1)]
stopifnot(max(abs(off_diag)) < 1e-12)
stopifnot(est_d1$V[[1L]]$vcov_method == "working_independence")
stopifnot(attr(est_d1, "sampling_relation") == "same_units")
stopifnot(attr(est_d1, "denominator_pattern") == "common")
stopifnot(attr(est_d1, "d1_regime") == "D1a")
stopifnot(all.equal(unname(diag(V1)), unname(grp$se^2), tolerance = 1e-12))
```

All three invariants hold.

# 7. Where to go next

-   [M7 · Fréchet envelope theory](m7-frechet-envelope-theory.html)
    for formal raw pairwise intervals and the separate K > 2 projected
    stress-scenario semantics.
-   [A5 · Published aggregates D0 / D1](a5-published-aggregates.html)
    for the applied walkthrough.
-   [M2 · Scalar SE — binomial](m2-scalar-se-binomial.html) for
    the row-level SE derivation reused by D0.
-   [M6 · Variance smoothing theory](m6-variance-smoothing-theory.html)
    for the experimental append-only log-variance smoother that can be
    applied to the same row-level SEs.

# References
