---
title: "M6 · Variance smoothing theory"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For methodologists evaluating the experimental generalized
  variance-function/log-variance smoother in `sm_smooth_variance()`.
  Derives the log-linear smoother, documents the Wilson-centered
  offset, and locks its provenance and validation conditions.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M6 · Variance smoothing theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
```

# Overview

This article is written for **methodologists** evaluating the experimental
generalized-variance-function/log-variance smoother in `sm_smooth_variance()`
who need its exact specification and *why* it stays opt-in and append-only. We
cover, in order:

1. what an analyst sees and what `sitemix` encodes;
2. the local notation this derivation adds;
3. the default log-linear log-variance smoother;
4. the Wilson-centered raw-scale offset;
5. the GAM extension;
6. the conditions and classed-warning failure policy;
7. the implementation invariants;
8. the fixed-seed simulation gate.

**Established vs. novel.** *Established:* generalized-additive-model smoothing
[@wood_2017] and the Fay–Herriot area-level model as a contrast
[@fay_herriot_1979]. *This package:* the opt-in, append-only
generalized-variance-function smoother (explicitly NOT a Fay–Herriot
estimator), the Wilson-centered log-variance offset, and the classed-warning
failure policy.

| Result | Attribution |
|:-------|:------------|
| GAM smoothing | Wood (2017) |
| Fay–Herriot area-level model (contrast only) | Fay & Herriot (1979) |
| append-only GVF smoother + offset + warning policy | this package (sitemix) |

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 1. What an analyst sees, what `sitemix` encodes

In the small-cell regime, the arcsine transformed-scale SE
$s_{jt} = 1/(2\sqrt{n_{jt}})$ is larger by construction: a site with $n = 5$
has an SE of `0.224`, while a site with $n = 100$ has `0.050`.
This denominator-driven difference is canonical sampling uncertainty, not
evidence that either SE is defective. A generalized variance function (GVF)
can fit a cross-row trend on the log-variance scale as an optional experimental
sensitivity, but the fixed-seed simulation audit did not support default promotion or a
general improvement claim.

`sm_smooth_variance()` is the experimental sitemix implementation.
It is not a Fay--Herriot area-level estimator: it does not model
the underlying point estimates or produce small-area posterior means.
By default it appends `se_smoothed` for `scale = "se"` or
`se_raw_smoothed` for `scale = "se_raw"`, plus
`var_method_smoothed`, while preserving canonical uncertainty columns.

# 2. Notation map

| Symbol | Meaning | Code | Range |
|:-------|:--------|:-----|:------|
| $v_{jt}$ | Row variance ($s_{jt}^2$) | `var` (internal) | $> 0$ |
| $\log v_{jt}$ | Log-variance | `log_var` | $\mathbb{R}$ |
| $\log n_{jt}$ | Log-denominator | `log_n` | $\mathbb{R}$ |
| $p^*_{jt}$ | Boundary-safe probability: raw estimate in the interior, Wilson center at a boundary | `.sm_smoothing_offset_p()` | $(0, 1)$ |
| $\ell^{\mathrm{off}}_{jt}=\log[p^*_{jt}(1-p^*_{jt})]$ | Rate-dependent log offset supplied to the model frame | `p_offset` | $(-\infty,\log(1/4)]$ |
| $\tilde s_{jt}$ | Smoothed SE | `se_smoothed` | finite, $\ge 0$ |

# 3. The log-linear smoother (default)

**Proposition 1 (Loglinear smoother).** *For the default
`scale = "se"` path, fit the OLS regression*

$$
\boxed{
\log v_{jt} \;=\; \beta_0 + \beta_1 \log n_{jt} + \varepsilon_{jt}.
}
\tag{M6.1}
$$

*For `scale = "se_raw"`, add the rate-dependent offset*

$$
\log v^{raw}_{jt} \;=\; \beta_0 + \beta_1 \log n_{jt} +
\mathrm{offset}\{\log[p^{\mathrm{off}}_{jt}(1-p^{\mathrm{off}}_{jt})]\} +
\varepsilon_{jt}.
\tag{M6.2}
$$

*Predicted log-variance back-transforms with an optional Jensen
correction:*
$\tilde v_{jt} = \exp(\widehat{\log v_{jt}} + \tfrac{1}{2}\hat\sigma^2_{\varepsilon})$,
*and* $\tilde s_{jt} = \sqrt{\tilde v_{jt}}$.

This regression is a GVF/log-variance smoother, not a small-area
outcome model.

The slope $\beta_1$ is expected to be near $-1$ (the arcsine
working variance is $1/(4n)$ in the no-FPC, uncorrected baseline, so
$\log v \approx -\log(4n)$ in that scope).
The function emits
`sitemix_warning_unexpected_slope` when the fitted slope
deviates by more than 0.15. With FPC or a relevant bias correction,
departure from $-1$ can be a predictable design effect; interpret the warning
against the recorded provenance rather than as evidence of estimator failure.

# 4. The Wilson-centered raw-scale offset

For boundary rows ($\hat\pi \in \{0, 1\}$), the raw-scale
variance $\hat\pi(1-\hat\pi)/n$ is zero and the log is
undefined. The Wilson-centered offset used in the raw-scale path is

$$
p^*_{jt} \;=\; \frac{\hat\pi_{jt} + z^2/(2n_{jt})}{1 + z^2/n_{jt}}
\;=\; \frac{C_{jt} + z^2/2}{n_{jt} + z^2}
$$

substitutes a finite proportion at the boundary, matching the
Wilson score center [@wilson_1927]. The plus-four Agresti-Coull
form is a useful approximation when $z \approx 2$
[@agresti_coull_1998]. This keeps the raw-scale log-variance
regression numerically well-defined.

The model-frame column is not this probability itself. It is
`p_offset = log(p_star * (1 - p_star))`, the quantity shown directly
inside the offset in (M6.2). A custom formula should therefore use
`offset(p_offset)`, not apply a second probability-to-log transformation.

## Finite-population inputs and provenance

Smoothing consumes the SEs already present in the input. If those SEs were
design-adjusted under SRSWOR, the fitted response therefore already contains
that adjustment. For an uncorrected arcsine row,

$$
\log(s^2)=-\log 4-\log n+\log q,
$$

so the simple $-1$ slope reference applies only when FPC is absent (and no
relevant bias correction changes the baseline). In append mode, smoothing
leaves canonical `se`, `se_raw`, and all eight FPC/design fields unchanged.
In overwrite mode, the selected original SE is retained in
`se_pre_smoothing` or `se_raw_pre_smoothing`; the unchanged FPC fields still
describe the pre-smoothing design estimator, not a newly design-derived
smoothed SE. An FPC offset or an FPC rejection gate would be a future design
change and is not part of the v0.2.0 contract.

# 5. The GAM extension

For non-linear log-variance trends, `method = "gam"` swaps the
OLS fit for a `mgcv::gam` fit with a smooth term on $\log n$:

$$
\log v_{jt} \;=\; f(\log n_{jt}) + \varepsilon_{jt},
\tag{M6.3}
$$

with $f$ a thin-plate spline. Raw-scale GAM smoothing uses the
same `p_offset` convention as raw-scale loglinear smoothing. The
`mgcv` dependency is runtime-guarded [@wood_2017].

# 6. Conditions and failure policy

`sm_smooth_variance()` emits three classed warnings:

-   **`sitemix_warning_smoother_multi_year_default`** — when
    eligible rows span multiple years and `by = NULL`. The
    smoother pools years by default; pass `by = "year"` to add
    a year fixed effect.
-   **`sitemix_warning_unexpected_slope`** — when the
    log-linear slope deviates from $-1$ by more than 0.15.
-   **`sitemix_warning_raw_scale_smoothing`** — when smoothing
    on raw-scale SE. The default raw-scale formula includes
    `offset(p_offset)`; the warning reminds users that custom
    raw-scale formulas should include an analogous rate-dependent
    term.

Matching-scale matrix staleness is not a warning. With
`overwrite = TRUE`, a matching-scale `V` raises
`sitemix_error_smoothing_v_stale` before fitting or stamping
provenance. An incompatible-scale `V` remains unchanged and the exact
relationship is recorded in the smoothing attribute and fit summary.
Custom formulas require exact left-hand side `log_var`, documented
helper variables and terms, a full-rank converged fit with positive
residual degrees of freedom, and finite predictions of the exact
eligible-row length. There is no silent non-finite fallback.

# 7. Implementation invariants

| ID  | Layer            | Claim                                                       |
|:----|:-----------------|:------------------------------------------------------------|
| SM1 | smooth            | Smoothed SE is finite and non-negative.                     |
| SM2 | smooth            | Default loglinear formulas differ by scale as documented.    |
| SM3 | smooth            | Default `overwrite = FALSE` preserves the input `se` exactly.|
| SM4 | provenance        | Append-only smoothing preserves canonical `var_method`.      |
| SM5 | matrix contract   | Matching-scale `V` blocks overwrite; incompatible `V` is explicit. |
| SM6 | FPC provenance    | Append mode preserves canonical SEs and all eight FPC/design fields. |

Verify on `alprek_subset`:

```{r invariants}
data(alprek_subset, package = "sitemix")
est <- sm_estimate(
  subset(alprek_subset, year == 2024),
  family = "binomial", indicator = "frpm"
)
est_s <- sm_smooth_variance(
  est, method = "loglinear", return_diagnostics = TRUE, min_rows = 2
)

# SM1: finite nonnegative smoothed SE
stopifnot(all(is.finite(est_s$se_smoothed)))
stopifnot(all(est_s$se_smoothed >= 0))

# SM2: default transformed-scale formula has no offset
fit_formula <- formula(attr(est_s, "smoother_fit"))
stopifnot(identical(deparse(fit_formula), "log_var ~ log_n"))

est_raw <- capture_expected_sitemix_warning(
  sm_smooth_variance(
    est, method = "loglinear", scale = "se_raw",
    return_diagnostics = TRUE, min_rows = 2
  ),
  "sitemix_warning_raw_scale_smoothing"
)
raw_formula <- formula(attr(est_raw, "smoother_fit"))
stopifnot(identical(deparse(raw_formula),
                    "log_var ~ log_n + offset(p_offset)"))

# SM3: default preserves se
stopifnot(all.equal(est_s$se, est$se, tolerance = 1e-12))

# SM4: alternative provenance does not rewrite canonical provenance
stopifnot(identical(est_s$var_method, est$var_method))
stopifnot("var_method_smoothed" %in% names(est_s))

# SM6: append-only smoothing preserves design-adjusted inputs and provenance
counts <- readRDS(system.file("extdata", "alprek_subset_counts.rds",
                              package = "sitemix"))
fpc_counts <- counts[counts$year == 2024,
                     c("site_id", "year", "n_jt", "c_jt_frpm")]
est_fpc <- sm_estimate_from_counts(
  fpc_counts, family = "binomial", indicator = "frpm",
  fpc = 2 * fpc_counts$n_jt, min_n = 1L
)
est_fpc_s <- sm_smooth_variance(est_fpc, method = "loglinear", min_rows = 2L)
fpc_fields <- c(
  "population_size", "sampling_fraction",
  "fpc_variance_multiplier", "fpc_se_multiplier",
  "variance_multiplier_applied", "se_multiplier_applied",
  "sampling_design", "variance_rule"
)
stopifnot(identical(est_fpc_s$se, est_fpc$se))
stopifnot(identical(est_fpc_s$se_raw, est_fpc$se_raw))
stopifnot(all(vapply(fpc_fields, function(field) {
  identical(est_fpc_s[[field]], est_fpc[[field]])
}, logical(1))))
```

SM1--SM4 and SM6 test the implemented behavior directly. SM5 is covered by
the package's matching/incompatible-scale matrix contract tests.

# 8. Fixed-seed simulation gate

The package's maintainer audit is
`inst/scripts/audit-smoothing-simulation.R`. It fixes seed `20260712` and
crosses 12 denominators (`n = 8` through `150`) with seven true rates
(`0.02` through `0.98`). The four prespecified strata are small/large
denominator by near-boundary/interior rate. The transformed target is the
deterministic contract variance $1/(4n)$; the separately coded raw target is
$p(1-p)/n$. Neither target calls an internal `sitemix` variance helper.

The promotion gate requires at least 10% pooled relative-MSE improvement, no
stratum with more than 10% MSE degradation, no material coverage degradation,
and at least 5% improvement in total-variation distance between estimated and
oracle inverse-variance weights. In the 60-replicate audit, every
loglinear/GAM and raw/transformed candidate received **NO-GO**. In
particular, raw-scale MSE was more than twice the unsmoothed comparator, while
the transformed comparator was already deterministic and offered no material
improvement to recover. Coverage occasionally moved closer to 95%, but that
single metric did not override the failed joint gate.

This is a bounded contract audit, not evidence of universal inferiority. It
does not span every data-generating process or hierarchical model. The policy
decision is therefore narrow: retain the existing function name and
`scope = "all"` only for compatibility inside an explicitly experimental,
opt-in, append-only helper; do not promote smoothing, canonical overwrite, or
a smoothed inverse-variance analysis as a default. A future promotion requires
a new, prespecified simulation that passes all criteria.

# 9. Where to go next

-   [M7 · Fréchet envelope theory](m7-frechet-envelope-theory.html)
    — the sibling sensitivity tool for the D1 case.
-   [A7 · Variance smoothing and Fréchet](a7-variance-smoothing-and-frechet.html)
    for the applied face including the GAM demo.

# References
