---
title: "M7 · Fréchet pairwise intervals and projected stress theory"
author: "JoonHo Lee"
date: "2026-07-12"
abstract: >
  For methodologists evaluating D1 cross-marginal sensitivity.
  Derives the Fréchet-Hoeffding pairwise intervals on the joint
  cross-marginal probability, distinguishes the formal D1a
  (common-population) regime from the heuristic D1b (subgroup-
  conditional) regime, and separates those intervals from PSD-projected
  corner stress scenarios.
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: true
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{M7 · Fréchet pairwise intervals and projected stress theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r setup, include = FALSE, error = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, results = "markup", comment = "#>", collapse = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  fig.width = 7, fig.height = 4.5, fig.retina = 2, dpi = 144,
  fig.align = "center", out.width = "100%"
)
options(warn = 2)
set.seed(
  1L, kind = "Mersenne-Twister", normal.kind = "Inversion",
  sample.kind = "Rejection"
)
capture_expected_sitemix_warning <- function(expr, expected_class) {
  captured <- list()
  value <- withCallingHandlers(
    force(expr),
    sitemix_warning = function(condition) {
      captured[[length(captured) + 1L]] <<- condition
      invokeRestart("muffleWarning")
    }
  )
  stopifnot(
    length(captured) <= 1L,
    all(vapply(captured, inherits, logical(1), expected_class))
  )
  observations <- getOption("sitemix.vignette.warning_observations", list())
  observations[[length(observations) + 1L]] <- list(
    expected_class = expected_class,
    count = length(captured)
  )
  options(sitemix.vignette.warning_observations = observations)
  value
}
library(sitemix)
```

# Overview

This article is written for **methodologists** evaluating D1 cross-marginal
sensitivity who need the exact pairwise intervals `sitemix` reports and *why*
their PSD projections are labeled stress scenarios rather than bounds. We
cover, in order:

1. what an analyst sees and what `sitemix` encodes;
2. the local notation this derivation adds;
3. the Fréchet–Hoeffding bounds;
4. the formal-D1a vs heuristic-D1b distinction;
5. why PSD projection creates stress scenarios, not bounds;
6. the implementation invariants.

**Established vs. novel.** *Established:* the Fréchet–Hoeffding bounds
[@frechet_1951; @hoeffding_1940] and the copula framing
[@nelsen2006introduction]. *This package:* the formal-D1a vs heuristic-D1b
distinction, and the PSD projection that yields labeled stress scenarios
(explicitly NOT lower/upper bounds).

| Result | Attribution |
|:-------|:------------|
| Fréchet–Hoeffding bounds | Fréchet (1951), Hoeffding (1940) |
| copula framing | Nelsen (2006) |
| D1a/D1b distinction + projected stress scenarios | this package (sitemix) |

> **About the example data.** All results here are computed live from
> `alprek_subset`, an anonymized, illustrative 50-site sample of Alabama
> First Class Pre-K records shipped in the package (see `?alprek_subset`).
> It is not a real accountability dataset and must not be cited as empirical
> Pre-K results. Every code block runs offline with a fixed random seed.

# 1. What an analyst sees, what `sitemix` encodes

In Scenario D1, the publisher provides marginal proportions
$\pi_{jt,k}$ for $K$ indicators but **not** the joint
proportions $\pi_{jt,kk'}$. The cross-marginal covariance is
unidentified from the marginals alone (M5 §4). The Fréchet-
Hoeffding bounds are the tightest interval the marginals can
support; any joint distribution consistent with the marginals
must lie in this interval.

# 2. Notation map

| Symbol | Meaning | Code | Range |
|:-------|:--------|:-----|:------|
| $\pi_x, \pi_y$ | Marginal proportions for two indicators | `theta_raw` | $[0, 1]$ |
| $\pi_{\mathrm{lower}}$ | Fréchet lower bound on joint | (internal) | $\max(0, \pi_x + \pi_y - 1)$ |
| $\pi_{\mathrm{upper}}$ | Fréchet upper bound on joint | (internal) | $\min(\pi_x, \pi_y)$ |
| $\sigma_{kk'}^{D1a}$ | Formal raw pairwise covariance interval | `raw_pairwise_intervals` | pairwise Fréchet endpoints |
| $\Sigma^-_{\mathrm{stress}}, \Sigma^+_{\mathrm{stress}}$ | PSD-projected corner stress matrices | `projected_negative_dependence_stress` / `projected_positive_dependence_stress` | not ordered bounds |

# 3. The Fréchet-Hoeffding bounds

**Proposition 1 (Fréchet-Hoeffding bounds).** *For binary
marginals $\pi_x, \pi_y \in [0, 1]$, every joint proportion
$\pi_{xy}$ consistent with the marginals satisfies*

$$
\boxed{
\max(0, \pi_x + \pi_y - 1) \;\le\; \pi_{xy} \;\le\; \min(\pi_x, \pi_y).
}
\tag{M7.1}
$$

**Attribution.** Classical Fréchet-Hoeffding bounds
[@frechet_1951; @hoeffding_1940]; see also
@nelsen2006introduction §2.5.

The probability bounds imply pairwise correlation bounds. For formal D1a,
the raw covariance endpoint is
$\sigma_{kk'} = (\pi_{xy} - \pi_x \pi_y) / n_{jt}$ (M3.1).
`raw_pairwise_intervals` stores the two marginals, common denominator, joint
probability endpoints, correlation endpoints, and these raw covariance
endpoints directly. Formal D1a therefore requires explicit `same_units`
provenance, one finite common denominator per site-year, the IID plug-in
variance rule, and no FPC or `binomial_bc` correction.

For heuristic D1b, there is no justified common $n_{jt}$. The object records
`n_common = NA` and constructs a heuristic covariance range as

$$
V_{kk'}^{\mathrm{corner}} = \rho_{kk'}^{\mathrm{corner}} s_k s_{k'},
$$

where $s_k$ and $s_{k'}$ are raw-scale row standard errors. The
`covariance_construction` and `interval_scope` columns keep this heuristic
range separate from the formal `/n` result.

# 4. D1a vs D1b — formal vs heuristic

**D1a (formal).** When the publisher marginals refer to the same sampled
units and satisfy the compatible denominator/design contract, each raw
pairwise interval is *formally valid*: any feasible pairwise joint
distribution must lie in that interval. Equal denominators alone do not
establish same-unit provenance.

**D1b (heuristic).** When the marginals are *subgroup-
conditional* (e.g., FRPM rate among ELL students vs FRPM rate
among non-ELL students), the marginals refer to different
populations and the Fréchet interval is *not* a formal joint constraint.
The returned ranges and projected matrices are then heuristic stress
quantities under an explicit counterfactual dependence construction.

`sm_frechet_envelope()` requires explicit acknowledgement of
this distinction via the `population_regime` argument
(`"d1a"` or `"d1b"`).

# 5. PSD projection creates scenarios, not bounds

The matrices assembled from all negative or all positive pairwise endpoints
are not guaranteed to be PSD. The function
applies one of two PSD repair methods:

-   **`psd_method = "higham"`** (default) — Higham's
    nearest-PSD algorithm via `Matrix::nearPD()`. Iterates
    using relative eigenvalue and convergence tolerances. The final
    PSD check requires the smallest eigenvalue to be no smaller than
    minus `psd_tol` times the matrix eigenvalue scale, plus a
    machine-range absolute floor. A `nearPD()` result whose `converged`
    flag is false is rejected. Supported `...` controls are sanitized and
    stored; package-owned diagonal, tolerance, iteration, symmetry, and output
    controls cannot be overridden [@higham_2002].

The public relative tolerance is restricted to
`0 < psd_tol <= sqrt(.Machine$double.eps)`. This keeps it a floating-point
allowance rather than a policy knob that could relabel a materially indefinite
corner as already PSD.
-   **`psd_method = "shrink"`** — line search for the largest
    shrinkage weight $\alpha \in (0, 1]$ such that
    $\alpha \cdot \Sigma + (1 - \alpha) \cdot \mathrm{diag}(\Sigma)$
    is PSD. Preserves the diagonal exactly. Automatic search errors when its
    alpha interval has not converged by `psd_max_iter`. A fixed alpha is the
    requested raw-corner retention and is applied exactly for `K > 2`, even
    when the source corner is already PSD; an infeasible value errors.

For `K <= 2`, both projected fields are exact identities of the source corners;
no Higham or shrink projection is attempted.

For `K > 2`, projection is a global matrix operation. An off-diagonal can
change sign, the projected negative-corner value can exceed the projected
positive-corner value, and either projected value can leave its raw pairwise
interval. Consequently the projected matrices are named stress scenarios.
`projection_diagnostics` is canonical long data with one row per site-year and
stress scenario. In addition to sign changes, projected-order reversals, and
raw-interval violations, it records scale labels, method/status,
attempted/converged state, iterations, relative and realized absolute
tolerances, eigenvalue scales and minima, requested/applied shrinkage,
projection distance, and diagonal/symmetry/PSD invariants. Deprecated
`psd_diagnostics` remains a wide compatibility view.

# 6. Implementation invariants

| ID  | Layer            | Claim                                                                  |
|:----|:-----------------|:------------------------------------------------------------------------|
| FR1 | frechet | D1a `raw_pairwise_intervals` reconstructs every covariance endpoint as $(q-p_kp_l)/n$. |
| FR2 | frechet | Each projected corner stress matrix satisfies the scale-aware relative `psd_tol` criterion and preserves the raw plug-in diagonal. |
| FR3 | frechet | For `K=2`, no projection is needed, so both stress matrices equal their source corners and pairwise ordering is preserved. |
| FR4 | frechet | For `K>2`, sign/order/range changes are diagnostics, not evidence of multivariate bounds. |
| FR5 | frechet | Every `K>2` projected stress matrix is deterministically replayable from its source corner and stored `projection_config`; nonconvergence is a hard error. |

Verify on a synthetic D1 slice:

```{r invariants}
counts <- readRDS(system.file("extdata", "alprek_subset_counts.rds",
                              package = "sitemix"))
d1 <- counts[counts$year == 2024, ]
d1_long <- rbind(
  data.frame(site_id = d1$site_id, year = d1$year,
             indicator = "frpm", c_jt = d1$c_jt_frpm, n_jt = d1$n_jt),
  data.frame(site_id = d1$site_id, year = d1$year,
             indicator = "snap", c_jt = d1$c_jt_snap, n_jt = d1$n_jt)
)
est_d1 <- capture_expected_sitemix_warning(
  sm_estimate_from_aggregates(
    d1_long, family = "multivariate",
    indicator_col = "indicator",
    sampling_relation = "same_units",
    vjt = TRUE
  ),
  "sitemix_warning_working_independence_default"
)
env <- sm_frechet_envelope(est_d1, population_regime = "d1a")

V_indep <- as.matrix(env$V_independence[[1L]])
raw_intervals <- env$raw_pairwise_intervals
V_negative <- as.matrix(env$projected_negative_dependence_stress[[1L]])
V_positive <- as.matrix(env$projected_positive_dependence_stress[[1L]])

# FR1: direct formal /n oracle
row1 <- raw_intervals[raw_intervals$site_key == env$site_keys$site_key[[1L]], ][1L, ]
stopifnot(all.equal(
  row1$pairwise_covariance_lower,
  (row1$joint_probability_lower - row1$p_1 * row1$p_2) / row1$n_common
))
stopifnot(all.equal(
  row1$pairwise_covariance_upper,
  (row1$joint_probability_upper - row1$p_1 * row1$p_2) / row1$n_common
))

# FR2: PSD after repair (relative tolerance + machine-range floor)
eig_negative <- eigen(V_negative, symmetric = TRUE, only.values = TRUE)$values
eig_positive <- eigen(V_positive, symmetric = TRUE, only.values = TRUE)$values
tol_negative <- env$psd_tol * max(abs(eig_negative)) +
  nrow(V_negative) * .Machine$double.xmin
tol_positive <- env$psd_tol * max(abs(eig_positive)) +
  nrow(V_positive) * .Machine$double.xmin
stopifnot(min(eig_negative) >= -tol_negative)
stopifnot(min(eig_positive) >= -tol_positive)

# FR3: K=2 shortcuts preserve their source corners
stopifnot(nrow(V_negative) == 2L)
stopifnot(all.equal(
  V_negative,
  as.matrix(env$unprojected_negative_dependence_corner[[1L]])
))
stopifnot(all.equal(
  V_positive,
  as.matrix(env$unprojected_positive_dependence_corner[[1L]])
))
stopifnot(all.equal(diag(V_negative), diag(V_indep), tolerance = 1e-8))
stopifnot(all.equal(diag(V_positive), diag(V_indep), tolerance = 1e-8))
```

The first three invariants hold for the two-indicator bundled example
(relative iterative-solver tolerance `1e-8`). For `K > 2`, pairwise
Fréchet feasibility does not by itself guarantee global joint
compatibility; the PSD step produces diagnostic scenarios and cannot promote
pairwise intervals into multivariate bounds.

# 7. Where to go next

-   [M5 · Aggregate engines D0 / D1](m5-aggregate-engines.html)
    for the working-independence baseline and D1 provenance contract.
-   [A7 · Variance smoothing and Fréchet](a7-variance-smoothing-and-frechet.html)
    for the applied diagnostic-envelope workflow.
-   [M8 · Output contract](m8-output-contract.html)
    for the package-neutral scalar and covariance contract, including why
    these projections stay separate from ordinary covariance.

# References
